Distinct transcriptomic profile of satellite cells contributes to preservation of neuromuscular junctions in extraocular muscles of ALS mice

Version 1

Abstract

Abstract

Amyotrophic lateral sclerosis (ALS) is a fatal neuromuscular disorder characterized by progressive weakness of almost all skeletal muscles, whereas extraocular muscles (EOMs) are comparatively spared. While hindlimb and diaphragm muscles of end-stage SOD1G93A (G93A) mice (a familial ALS mouse model) exhibit severe denervation and depletion of Pax7+ satellite cells (SCs), we found that the pool of SCs and the integrity of neuromuscular junctions (NMJs) are maintained in EOMs. In cell sorting profiles, SCs derived from hindlimb and diaphragm muscles of G93A mice exhibit denervation-related activation, whereas SCs from EOMs of G93A mice display spontaneous (non-denervation-related) activation, similar to SCs from wild-type mice. Specifically, cultured EOM SCs contain more abundant transcripts of axon guidance molecules, including Cxcl12, along with more sustainable renewability than the diaphragm and hindlimb counterparts under differentiation pressure. In neuromuscular co-culture assays, AAV-delivery of Cxcl12 to G93A-hindlimb SC-derived myotubes enhances motor neuron axon extension and innervation, recapitulating the innervation capacity of EOM SC-derived myotubes. G93A mice fed with sodium butyrate (NaBu) supplementation exhibited less NMJ loss in hindlimb and diaphragm muscles. Additionally, SCs derived from G93A hindlimb and diaphragm muscles displayed elevated expression of Cxcl12 and improved renewability following NaBu treatment in vitro. Thus, the NaBu-induced transcriptomic changes resembling the patterns of EOM SCs may underlie the beneficial effects observed in G93A mice. More broadly, the distinct transcriptomic profile of EOM SCs may offer novel therapeutic targets to slow progressive neuromuscular functional decay in ALS and provide possible “response biomarkers” in pre-clinical and clinical studies.

Introduction

Amyotrophic lateral sclerosis (ALS) is a fatal disorder characterized by progressive motor neuron (MN) loss and skeletal muscle wasting, resulting in respiratory failure and death 3–5 years following diagnosis [1,2]. While not entirely spared, the extraocular muscles (EOMs) of ALS patients and mutant mouse models exhibit comparatively preserved structure and function and retained neuromuscular junctions (NMJ) [3–6]. The reason for this differential involvement remains unclear, especially considering that eye movements are preferentially involved in other neuromuscular disorders like myasthenia gravis, some muscular dystrophies, and mitochondrial myopathies [7,8].

While ALS is classically considered predominantly a “dying-forward” process of motor neurons (MN), accumulated evidence [9–17] support that early muscle dysfunction is an important contributor to ALS pathophysiology. The degree to which “dying-forward” from the central nerve system or “dying-back” process [18–20] starting distally at NMJs contributing to ALS progression remains unsettled and may vary in different patient subsets. Regardless of the direction of communication and relative contribution of different cell types to ALS progression (e.g. glia, neurons, myofibers) [20], it is widely accepted that NMJ disassembly and the resulting skeletal muscle denervation is a critical pathogenic event in both patients and animal models [21–24].

EOMs differ from limb and trunk muscles in many respects [25]. In contrast to other skeletal muscles, the individual motor neurons (MNs) controlling EOMs innervate fewer EOM myofibers, and individual myofibers often have multiple NMJs [26]. The multiple innervation phenotype refers to the co-existence of innervated “en plaque” and “en grappe” NMJs in a single EOM myofiber. The “en plaque” type of NMJs are located in the middle portion of myofibers and resemble NMJs seen in other types of muscles. Each EOM myofiber has one “un plaque” NMJ. “en grappe” NMJs are a series of smaller, grape-shaped NMJs that are located in the distal portion of some EOM myofibers [3,26]. MNs innervating EOMs express higher level of the Ca2+ binding proteins calbindin-D28k and parvalbumin, thus may be more resistant to the Ca2+-induced excitotoxicity [27–29]. The expression profile of EOM myofibers is also different from that of limb muscle in over 300 genes [30]. Previous studies have investigated the muscle-derived neuroprotective factors, including BDNF, GDNF, NT3 and NT4 and found that none of these classical neurotrophic factors was expressed (RNA or proteins) significantly higher in EOM myofibers compared to hindlimb myofibers in both mutant G93A and wild-type mice [31,32]. Thus, it is important to look for clues in other cell types that influence the microenvironment around NMJs, such as satellite cells (SCs).

An association between SC function, NMJ integrity, and resultant denervation has been demonstrated in animal models of aging and muscular dystrophy [33,34]. Using transgenic mice expressing GFP in the progeny of activated SCs, Liu and colleagues demonstrated that inducible depletion of adult skeletal muscle SCs impaired the regeneration of NMJs [33]. Furthermore, they found that the loss of SCs exacerbated age-related NMJ degeneration. Preservation of SCs attenuated aging-related NMJ loss and improved muscle performance in mice [34]. SCs from EOMs exhibit superior expansion and renewability (the capability to maintain dividing and undifferentiated stem cells after rounds of regeneration) than their limb counterparts [35], inspiring us to compare the phenotypes of SCs derived from EOMs and limb muscles and to conduct transcriptomic profiling of SCs to look for candidate genes that could preserve NMJ integrity in ALS.

In the current study, we found that the integrity of NMJs and the amount of Pax7+ SCs are maintained in EOMs in end-stage G93A mice, in contrast to the severe denervation and depletion of Pax7+ SCs in hindlimb and diaphragm muscles. Transcriptomic profiling revealed that the EOM SCs exhibit abundant transcripts of axon guidance molecules, such as Cxcl12, along with sustainable renewability compared with SCs from diaphragm and hindlimb muscles. Furthermore, EOM SC-derived myotubes enhance axon extension and innervation in co-culture experiments with spinal motor neurons compared to hindlimb SC-derived myotubes. Overexpressing Cxcl12 using adeno-associated viral vector (AAV8) in G93A hindlimb SC-derived myotubes recapitulated features of the EOM SC-derived myotubes.

Butyrate, a short-chain fatty acid produced by bacterial fermentation of dietary fibers in the colon, has been shown to extend the survival of ALS G93A mice [36]. We found that G93A mice fed with 2% sodium butyrate (NaBu) in drinking water decreased NMJ denervation and SC depletion in hindlimb and diaphragm muscles. Adding NaBu to the culture medium improved the renewability of SCs derived from the G93A hindlimb and diaphragm muscles, as well as increasing the expression of Cxcl12, resembling EOM SC transcriptomic pattern. Since muscle tissue can be biopsied during life, characterizing the distinct EOM SC transcriptomic pattern could provide potential predictive/cohort selective, and response/pharmacodynamic biomarkers in therapeutic trials in both ALS patients and animal models, in addition to identifying therapeutic targets.

Results

NMJ integrity and peri-NMJ SC abundance are preserved in EOMs, and NaBu-supplemented feeding reduces NMJ denervation and SC depletion in hindlimb and diaphragm muscles in G93A mice

We investigated the NMJ integrity and peri-NMJ SC abundance by whole-mount imaging of extensor digitorum longus (EDL), soleus, diaphragm, and EOMs derived from WT littermates and G93A mice with or without NaBu-supplemented feeding. For EOMs, only the “en plaque” type of NMJs were included in this comparison, because they resemble NMJs in other types of muscles in morphology and spatial distribution. The “en grappe” type of NMJs were not included. Z-stack scans of glycerol-cleared whole muscles were obtained using a high working distance lens in a confocal microscope. The z-stacks were compacted into 2D images by maximal intensity projection. Acetylcholine receptors (AChRs) at NMJs were labeled with Alexa Fluor conjugated α-Bungarotoxin (BTX). Axons and axon terminals of motor neurons were labeled with antibodies to neurofilament (NF) and synaptophysin (SYP) respectively (Figure 1A and Figure 1-figure supplement 1A). In order to quantify the extent of denervation in a categorical manner, NMJs were arbitrarily defined as “well innervated” if SYP staining was present in >60% of the BTX positive area, “partially innervated” if between 60% and 30%, and “poorly innervated” if SYP staining corresponded to less than 30% of the BTX positive area. Poorly innervated NMJs also generally lacked nearby NF+ axons (Figure 1-figure supplement 1A).

Characterizing NMJ integrity in muscles dissected from wild-type controls and end-stage G93A mice with or without NaBu feeding.

(A) Representative compacted z-stack scan images of whole-mount EDL, soleus diaphragm extraocular muscles stained with antibodies against neurofilament medium chain (NF, labeling axons), synaptophysin (SYP, labeling axon terminals) and Alexa Fluor 488 conjugated α-Bungarotoxin (BTX, labeling AChRs on muscle membrane). scale bars, 50 μm. (B) Mean ratios of well innervated NMJs (SYP signals are present in >60% of BTX positive area), partially innervated NMJs (SYP signals are present in 30-60% of BTX positive area) and poorly innervated NMJs (SYP signals are present in <30% of BTX positive area) and in different types of muscles dissected from WT controls and end-stage G93A mice with or without NaBu feeding (tens to hundreds of NMJs were measured per muscle type per mice, also see Figure 1-figure supplement 1 and Figure 1-table supplement 1). * P < 0.05; ** P < 0.01; **** P < 0.0001; ns, not significant.

At least 343 (average 758 ± 345 per group) NMJs from muscles (EDL, soleus, diaphragm and EOMs, respectively) dissected from WT or G93A mice with or without NaBu-feeding were examined (Figure 1-table supplement 1) in order to calculate the relative proportions of well-, partially-and poorly- innervated NMJs (Figure 1B). EDL, soleus and diaphragm muscles, but not EOMs, from end stage G93A mice (4-month-old) exhibited significant decrease in well-innervated NMJs, accompanied by an increase in poorly-innervate NMJs. The increase of “poorly-innervated NMJ” is slightly less dramatic in diaphragm compared to the hindlimb muscles, meanwhile the increase of “partially-innervated NMJ” ratio is the most significant in the diaphragm (Figure 1B), indicating that denervation process is most severe in hindlimb muscles, slightly milder in diaphragm and barely detectable in EOMs. Consistent with previous observation by our lab and others that NaBu treatment slows ALS progression [36], the hindlimb and diaphragm muscles from NaBu-fed mice (1 month, 2% in water, starting feeding at the age of 3-month-old) all exhibited significantly reduced denervation compared to littermates who were not fed NaBu (Figure 1B, n = 10 per group, gender matched).

qRT-PCR for Scn5a, the gene encoding sodium channel Nav1.5, was performed with RNA extracted from whole muscles and used as another marker of denervation [37–39] (Figure 1-figure supplement 1B). The original values of Scn5a expression levels are provided in Figure 1-table supplement 2. Consistent with the immunostaining quantification in Figure 1A, the elevation of Scn5a expression is most notable when comparing G93A hindlimb muscles to wildtype counterparts (>250 folds for EDL, >70 folds for soleus), milder for diaphragm (> 15 folds) and not significant for EOMs. Additionally, NaBu feeding significantly reduced the Scn5a transcripts in hindlimb muscles (Figure 1-figure supplement 1B).

Whole mount imaging of muscles stained with the satellite cell (SC) markers Pax7, BTX, SYP, and DAPI, was used to quantify SCs in the vicinity of NMJs. The Pax7 signal in each z-stack image was filtered using DAPI as a mask to identify Pax7+ cell nuclei. Afterwards the z-stacks were compacted into 2D images by maximal intensity projection (Figure 2A). The number of nuclear Pax7+ cells were counted within the circles of 75 µm diameter drawn around the NMJs (Figure 2B, see also Figure 2-table supplement 1). Similar to the NMJ denervation status, depletion of peri-NMJ SCs was most significant in hindlimb muscles of end-stage G93A mice and slightly milder in diaphragm, while no significant depletion was observed in EOMs (Figure 2C). G93A mice fed with NaBu had less depletion of SCs in hindlimb and diaphragm muscles.

Quantification of peri-NMJ SC abundance in muscles dissected from wild-type controls and end-stage G93A mice with or without NaBu feeding.

(A) Representative compacted z-stack scan images of whole-mount EDL, soleus diaphragm extraocular muscles stained with antibodies against Pax7 (labeling SCs), SYP and BTX-Alexa Fluor 488. scale bars, 50 μm. (B) To measure the abundance of SCs around NMJs, circles of 75 μm diameter were drawn around the NMJs and the number of nuclear Pax7+ cells (co-localized with DAPI) were counted. Scale bars, 10 μm. (C) Mean number of SCs around NMJs in different types of muscles dissected from WT controls and end-stage G93A mice with or without NaBu feeding (tens to hundreds of NMJs were measured per muscle type per mice, see Figure 2-table supplement 1). * P < 0.05; ** P < 0.01; **** P < 0.0001; ns, not significant.

The proliferation and differentiation properties are conserved in SCs derived from G93A EOMs

We conducted FACS based isolation of SCs from hindlimb (combining tibialis anterior, EDL, quadriceps and gastrocnemius), diaphragm, and EOMs using a modified version of a published protocol [40]. Soleus was excluded because it has higher content of slow-twitch myofibers (37-55%), which is more similar to diaphragm than the other fast-twitch fiber-dominant hindlimb muscles [41–44]. Vcam1 was used as a marker for quiescent SC. CD31 (endothelial cell marker), CD45 (hematopoietic cell marker) and Sca-1 (mesenchymal stem cell marker) positive cells were discarded. Thus, in the FACS profile the target SC population (Vcam1+CD31-CD45-Sca-1-) was in gate P5-4 (Figure 3A). Single antibody staining controls are shown in Figure 3-figure supplement 1. We also applied methyl-green (20 µg/ml) staining [45] to evaluate the viability of cells (Figure 3-figure supplement 2), only about 1% of Vcam1 positive population were methyl green stained (Figure 3-figure supplement 2D, Events P5-2 / (P5-2 + P5-4)*100%). Thus, most cells were alive when loaded into the flow cytometer.

Isolation of SCs from wildtype and G93A muscles using fluorescence activated cell sorting.

(A) Representative FACS profiles during the isolation of Vcam1+CD31-CD45-Sca1- cells (gate P5-4). The corresponding population hierarchies are shown right to the plots. For single antibody staining controls, unstained controls and viability test, see Figure 3-figure supplement 1 and Figure 3-figure supplement 2. (B) Comparing the percentage of events in the P5-4 gate between cells isolated from WT and G93A mice. * P < 0.05; *** P < 0.001; ns: not significant.

SCs isolated from WT hindlimb and diaphragm muscles exhibited relatively homogenous levels of Vcam1, which facilitated separation from other cells and debris (Figure 3A, gate P5-4), and also indicates they are mostly quiescent under resting conditions. In contrast, SCs derived from the G93A hindlimb and diaphragm displayed heterogeneous levels of Vcam1, potentially due to ongoing activation/differentiation triggered by disease progression (Figure 3A). Importantly, EOM SCs from both WT and G93A mice were heterogeneous in Vcam1 levels, implying they are spontaneously activated even in the absence of pathological stimulation (Figure 3A). While significant decreases of P5-4 population were observed in the G93A hindlimb and diaphragm derived SCs compared to WT, no significant difference was detected between G93A and WT EOM derived SCs (Figure 3B).

FACS-isolated SCs were next cultured in growth medium for 4 days, fixed, and stained for MyoD to confirm their myogenic lineage. Close to 100% of the cells were MyoD positive (Figure 4A). We examined the ratio of actively proliferating cells (Ki67+) among the SCs (labeled with Pax7) [46]. SCs isolated from both WT and G93A EOMs exhibited the highest Pax/Ki67 double positive ratio, and the lowest Pax7/Ki67 double negative ratio (Figure 4A). These observations are consistent with previous reports of superior proliferative potential of EOM SCs compared to limb and diaphragm muscles [35]. The direct comparison between WT and G93A muscle SCs is presented in Figure 4-figure supplement 1.

Proliferation and differentiation properties of FACS-isolated SCs in culture.

(A) Representative images of FACS-isolated SCs cultured for 4 days in growth medium stained with antibodies against Pax7, Ki67 (proliferating cell marker), MyoD (myogenic lineage marker). Measurement results for the ratios of Pax7+Ki67+, Pax7+Ki67- and Pax7-Ki67- cells are shown in the right two panels. Scale bars, 100 μm. * P < 0.05; ** P < 0.01; **** P < 0.0001; ns, not significant. (B) Representative images of FACS-isolated SCs cultured for 4 days in growth medium and 2 days in differentiation medium stained with antibodies against Pax7, Ki67, myosin heavy chain (MHC, differentiated myotube marker). Measurement results for the ratios of Pax7+Ki67+ and Pax7+Ki67- cells, as well as the fusion indices are shown in the right two panels.

We also compared myotube formation and cell proliferation of cultured SCs under differentiation conditions. After 4 days in growth medium, we switched to differentiation medium and cultured the cells for 2 more days. The cells were then stained for sarcomere myosin heavy chain (MHC), Pax7 and Ki67. The EOM-derived SCs exhibited a higher number of Pax7 positive cells (Pax7+/Ki67+ and Pax7+/Ki67-) compared to SCs derived from hindlimb and diaphragm muscles – from both WT and G93A mice (Figure 4B). In contrast, the fusion indices, which calculate the ratio of nucleus within the MHC positive myotubes, were the lowest for EOM SCs (Figure 4B). This does not necessarily mean the EOM SCs have deficiencies in differentiation after they are committed, but more likely a reflection of their superior preservation of actively proliferating population [47]. In support of this view, previous TA engraftment experiments showed that more myofibers were generated by EOM SCs than their limb counterparts [35].

Butyrate induces EOM-like preservation of renewability in FACS-isolated SCs under differentiation conditions in vitro

Since NaBu-supplemented feeding reduced SC depletion in end stage G93A mice (Figure 2) in vivo, we tested NaBu treatment on cultured SCs. G93A hindlimb SCs were cultured with 0, 0.1, 0.5, 1.0, 2.5 and 5 mM of NaBu added to the growth medium for 1 day. RNAs were collected for qRT-PCR to evaluate the short-term impact of NaBu on cell growth. The relative expression of the proliferation markers (HMGA2 and CCND1) peaked at 0.5 mM. In contrast, the relative expression of proliferation inhibitor P21 bottomed at 0.5 mM (Figure 5A), indicating 0.5 mM as the optimal concentration of NaBu application in vitro.

Proliferation and differentiation properties of FACS-isolated SCs with different NaBu treatment conditions.

(A) qRT-PCR based relative quantification of Hmga2, Ccnd1 and P21 expression in FACS-isolated G93A HL SCs cultured with different doses of NaBu in growth medium for 1 day (n = 6). *** P < 0.001; **** P < 0.0001. (B) Left two panels show the representative images of FACS-isolated G93A HL and diaphragm SCs cultured in growth medium for 4 days and differentiation medium for 2 days and experienced different NaBu treatment conditions and stained with Pax7, Ki67 and MHC antibodies. Cells in top row were not treated with NaBu. Cells in the middle row were treated with NaBu for 1 day before the induction of differentiation (Day 4 of culture). Cells in the bottom row were treated with NaBu continuously for 3 days from the day before the induction of differentiation (Day 4-6 of culture). Scale bars, 100 μm. Measurement results for the ratios of Pax7+Ki67+ and Pax7+Ki67- cells, as well as the fusion indices are shown in the right two panels. * P < 0.05; ns, not significant.

We designed two NaBu treatment plans for G93A diaphragm and hindlimb SCs as follows: one with 0.5 mM NaBu applied in the growth medium for 1 day before switching to differentiation medium (1 day in growth medium), the other with NaBu continuously applied for 3 days (1 day in growth medium + 2 days in differentiation medium). As shown in Figure 5B, the G93A hindlimb and diaphragm SCs in culture showed significant increase in the ratio of Pax7 positive cells under the 1-day NaBu treatment, associated with a decrease in fusion index. This trend was more pronounced under the 3-day NaBu treatment. These data indicate that NaBu treatment results in EOM-like preservation of renewability of the G93A SCs (Figure 4B). Aside from the drop of fusion indices after 3-day treatment with NaBu, there was notable decrease of MHC levels, indicating long-term exposure to NaBu did inhibit the differentiation process (Figure 5B), which is in line with the published result observed in C2C12 cells [48].

EOM-derived SCs show distinct transcriptome profile supporting self-renewal and axonal growth when compared to the SCs derived from other muscle origins

We extracted RNAs from SCs of hindlimb, diaphragm and EOMs under the following two culture conditions: first, in growth medium for 4 days (with suffix “G”); second, in growth medium for 4 days and then in differentiation medium for 2 days (with suffix “D”). The RNAs were send for bulk RNA-Seq. Additionally, RNA-Seq analyses were conducted for SCs of G93A hindlimb and diaphragm muscles cultured with NaBu for 3 days (with suffix “NaBu3”) (Figure 6).

Comparing the transcriptomic profile and homeostasis preferences of SCs from different muscle origins cultured in growth and differentiation medium.

(A) PCA analysis and hierarchical clustering results of different RNA-Seq samples. Samples with suffix “G” were SCs cultured in growth medium for 4 days. Suffix “D” denote SCs cultured in growth medium for 4 days and differentiation medium for 2 days. Suffix “NaBu3” denote SCs treated with 0.5 mM NaBu continuously for 3 days (Day 4-6 of culture). Dashed-line circles in PCA plot highlight samples close to each other in distance. (B) We identify genes differentially expressed in EOM SCs compared to their diaphragm and HL counterparts cultured in growth medium (leftmost panel) and differentiation medium (middle panel) by screening for genes shared by the four DEG lists (gold, magenta, pink and green in color) shown in the Venn diagrams. All DEG lists were generated based on the standard of |log2FC|>= 0.4 and FDR<= 0.001. Yellow box highlights the shared genes. To identify NaBu treatment signature genes, we screened for genes shared by the two DEG lists (green and gold in color) shown in the rightmost Venn diagram. (C) Dot plots in the upper row show the Log2FC of activation signature genes (magenta) and differentiation signature genes (cyan) identified in the 1119, 3318 and 3018 genes highlighted in yellow boxes in (B). Dot plots in the lower row show the Log2 FC of quiescence signature genes identified in the yellow-box highlighted genes in (B). Additionally, genes belonging to the following subgroups of quiescent signature genes: most commonly used SC markers, Notch signaling components and pluripotency signaling components are colored in dark blue, red and gold, respectively.

Similarities of the transcriptomic profiles of RNA-Seq samples were examined using the Principle Component Analysis (PCA) and hierarchical clustering (Figure 6A). We found that both WT and G93A EOM SCs cultured in differentiation medium were more similar to hindlimb and diaphragm SCs in growth medium than their counterparts in differentiation medium, highlighting the superior renewability of EOM SCs in differentiation medium and is consistent with the immunostaining results showed in Figure 4B.

To understand RNA-Seq data in the perspective of SC homeostasis, which refers to the transition between quiescence (dormant, slow self-renewal), activation (rapid amplification with symmetric and asymmetric cell division) and differentiation states during tissue regeneration [47], we compiled the lists of quiescence, activation and differentiation signature genes respectively. For quiescence signature genes, a list of 507 genes highly upregulated in quiescent SCs compared to activated SCs have been generated previously by Fukada et al [49]. We adopted this list together with other 18 genes denoted by other investigators, including Notch1, Notch2, Hes1, Hey1, Rbpj, Dtx4, Jag1, Sox8, Sdc3, Itga7, Itgb1, Met, Vcam1, Ncam1, Cdh2, Cdh15, Cxcr4, Cav1 [47,50–53]. The activation signature gene list was generated from our RNA-seq data using those expressed significantly higher in SCs (both from WT and G93A) in growth medium (1187 genes in total, |log2FC| ≥ 0.4 and FDR ≤ 0.001). Accordingly, the differentiation signature gene list was generated from those expressed significantly higher in SCs cultured in differentiation medium (1349 genes in total).

To verify that our signature gene lists recapitulate the SC properties at activation and differentiation state reported in previous studies, we did functional annotation analysis (KEGG pathway, Gene ontology_biological process, Gene ontology_cellular component). As shown in Figure 6-figure supplement 1, the top functional annotations enriched in our activation signature genes are all related to cell cycle and DNA replication, while the top functional annotations for our differentiation signature genes are associated with muscle contraction, muscle-specific subcellular structures and metabolic changes.

To explore whether the homeostatic states of EOM SCs are different from that of SCs from other muscle origins, we looked for differentially expressed genes (DEGs) shared-in-common among the following four groups: WT EOM SC-vs-WT diaphragm SC; WT EOM SC-vs-WT HL SC; G93A EOM SC-vs-G93A diaphragm SC; G93A EOM SC-vs-G93A HL SC (Figure 6B, left two panels). Within the 1119 shared DEGs identified in SCs cultured in growth medium, slightly more activation signature genes and quiescence signature genes were expressed higher in EOM SCs compared to diaphragm and HL derived SCs. Meanwhile more differentiation signature genes were expressed lower in EOM SCs (Figure 6C, leftmost panel and Figure 6-table supplement 1). These trends became much more prominent in the 3318 shared DEGs identified in SCs cultured in differentiation medium, implying that EOM SCs are notably superior at preserving renewability in a differentiation environment than their diaphragm and hindlimb counterparts (Figure 6C, middle panel). Consistently, subgroups of quiescence signature genes belonging to commonly used stem cell markers (Pax7, Cd34, Cxcr4, Vcam1, Sdc3, Sdc4, Cdh15, Met, Cav1), Notch signaling pathway (Notch1, Notch2, Notch3, Jag1, Heyl, Dtx4, Msc) and pluripotency signaling pathway (Klf4, Igf1, Bmp2, Bmp4, Fzd4, Fgfr4, Pik3r1, Lifr) were all expressed higher in EOM SCs in differentiation medium than SCs derived from diaphragm and hindlimb muscles (Figure 6C, middle panel and Figure 6-table supplement 2).

In parallel, we examined whether the 3-day NaBu treatment could alter the homeostasis of G93A diaphragm and hindlimb SCs (Figure 6B, rightmost panel). There were 3018 DEGs shared between the following two comparing groups: G93A diaphragm SC_D_NaBu3-vs-G93A diaphragm SC_D and G93A HL SC_D_NaBu3-vs-G93A HL SC_D. Similar to the trends observed in EOM SCs in differentiation medium, the majority of activation and quiescence signature genes found in these DEGs were upregulated in the NaBu treated group, while the differentiation signature genes were downregulated (Figure 6C, rightmost panel). The elevated quiescence signature genes of the NaBu treated group include commonly used stem cell markers (Pax7, Cd34, Vcam1, Sdc4, Cdh2, Met, Cav1, Itga7), Notch signaling pathway members (Notch3, Rbpj) and pluripotency signaling pathway components (Fzd4, Bmp4). However, other three pluripotency related genes (Klf4, Wnt4 and Bmp2) were found downregulated (Figure 6-table supplement 3), indicating that NaBu treatment also has a mixed impact on the stemness of the SCs.

To compile EOM SCs signature gene list, we identified 621 genes expressed higher in EOM SCs than diaphragm and hindlimb SCs cultured in growth medium from the 1119 DEGs (Figure 6B, leftmost panel). Next, we pulled out 2424 genes that are expressed higher in EOM SCs than diaphragm and hindlimb SCs cultured in differentiation medium from the 3318 DEGs (Figure 6B, middle panel). These two lists have 478 genes shared-in-common. Particularly, many of these genes are related to axon guidance, cell migration and neuron projection (Figure 7A). Figure 7A-figure supplement 1, Figure 7-table supplement 1 and 2 further demonstrate this phenotype when the functional annotation analysis was done separately for the growth or differentiation culture conditions. The functional annotation analysis for the 3018 genes altered by the 3-day NaBu treatment (Figure 6B, rightmost panel) also spotted axon guidance, cell migration and neuron projection on the top of the annotation lists (Figure 7B and Figure7-table supplement 3). Importantly, the chemokine Cxcl12 shows up in all three lists. Cxcl12 is known to promote axonal extension and NMJ regeneration [54]. These data suggest that enhanced attraction of motor neuron axons by peri-NMJ SCs might be a shared mechanism underlying the NMJ preservation in EOMs and NaBu treatment observed in G93A mice.

Functional annotation analysis of EOM SC signature genes and NaBu treatment signature genes and qRT-PCR results.

(A) Top KEGG pathway, GO_P and GO_C annotations of the 478 EOM signature genes (expressed higher in EOM SCs compared to diaphragm and HL counterparts cultured in both growth medium and differentiation medium). Arrows highlight annotations related to axon guidance and cell migration. (B) Top KEGG pathway, GO_P and GO_C annotations of the 3018 NaBu treatment signature genes shown in Figure 6B, rightmost panel. (C) qRT-PCR based relative quantification of Hmga2 (activation signature gene), Actn3 (differentiation signature gene), Notch3 (quiescence signature gene) and Cxcl12 (axon guidance molecule) expression in HL, diaphragm and EOM SCs cultured in growth and differentiation medium, respectively. * P < 0.05; ** P < 0.01; *** P < 0.001; **** P < 0.0001; ns, not significant. RNA samples were collected from 3-6 rounds of sorting and sorted cells were split into 3 replicates. (D) qRT-PCR based relative quantification of Hmga2, Actn3, Notch3 and Cxcl12 expression in G93A HL and diaphragm SCs with or without the 3-day NaBu treatment. RNA samples were collected from 6 rounds of sorting and sorted cells were split into 3 replicates.

We selected one representative gene from each of the following categories: activation (Hmga2), differentiation (Actn3) and quiescence (Notch3) signature genes, as well as an axon guidance gene (Cxcl12) for qRT-PCR. For each muscle type and culture condition, we collected samples from 3-6 rounds of sorting (the individual results are listed in Figure 7-table supplement 4-6). We observed significantly higher expression of Hmga2 in EOM SCs compared to diaphragm and hindlimb SCs cultured in either growth or differentiation medium (Figure 7C, leftmost panel), confirming the superior renewability of EOM SCs. The 3-day NaBu treatment of diaphragm and hindlimb SCs derived from G93A mice also upregulated Hmga2 (Figure 7D, leftmost panel). Actn3 was expressed lower in EOM SCs than diaphragm SCs cultured in either growth or differentiation medium, but not necessarily lower than hindlimb SCs (Figure 7C, 2nd panel from the left). Meanwhile NaBu treatment significantly decreased Actn3 expression (Figure 7D, 2nd panel from the left). Consistent with the RNA-Seq data, the quiescence marker Notch3 was expressed significantly higher in EOM SCs than diaphragm and hindlimb SCs cultured in differentiation medium but not in growth medium (Figure 7C, 2nd panel from the right). Similarly, NaBu treatment also elevated Notch3 expression (Figure 7D, 2nd panel from the right) in G93A diaphragm and hindlimb SCs cultured in differential medium. Importantly, the axon guidance molecule Cxcl12 was expressed significantly higher in EOM SCs than diaphragm and hindlimb SCs cultured in both growth and differentiation medium (Figure 7C, rightmost panel). NaBu treatment elevated Cxcl12 expression as well (Figure 7D, rightmost panel), confirming enhanced axon attraction as a common feature shared by EOM SCs and NaBu treatment.

AAV mediated overexpression of Cxcl12 in myotubes promotes motor neuron axon extension

Since recombinant Cxcl12 is known to promote axon extension [54], we tested whether overexpressing Cxcl12 in myotubes via adeno-associated virus (AAV) transduction could achieve the same effect. For these experiments, we co-cultured motor neurons and SC-derived myotubes on compartmentalized microfluidic chambers (XonaChip). Because AAV vectors do not transduce satellite cells efficiently [55,56], we first seeded the satellite cells into one compartment of the XonaChips, then induced them to differentiate for two days before AAV transduction. Four types of AAV-transduced myotubes were used for the coculture assays: a) WT hindlimb SC-derived myotubes transduced with AAV-CMV-eGFP; b) G93A hindlimb SC-derived myotubes transduced with AAV-CMV-eGFP; c) G93A hindlimb SC-derived myotubes transduced with AAV-CMV-Cxcl12-IRES-eGFP; and d) G93A EOM SC-derived myotubes transduced with AAV-CMV-eGFP. One day after AAV transduction, rat spinal motor neurons (RSMN) were seeded into the other compartment of the XonaChips and the culture medium switched to Neurobasal Plus Medium (NBPM) (Figure 8A). The cocultures were maintained for 9 more days. Immunostaining confirmed increased presence of Cxcl12 protein in the form of cytosolic vesicles in myotubes transduced with AAV8-CMV-Cxcl12-IRES-eGFP but not those transduced with AAV8-CMV-eGFP (Figure 8-figure supplement 1). These are likely vesicles for secretion [57].

Coculture of AAV transduced SC derived myotubes with rat spinal motor neurons in compartmentalized microfluidic chambers. (A)

Schematic representation of XonaChip spatial configuration and coculture experiment timeline. NBPM: Neurobasal Plus Medium; RSMNs: rat spinal motor neurons. (B) Representative composite images of RSMNs cocultured with: 1. WT HL SC-derived myotubes transduced with AAV-CMV-eGFP; 2. G93A HL SC-derived myotubes transduced with AAV-CMV-eGFP; 3. G93A HL SC-derived myotubes transduced with AAV-CMV-Cxcl12-IRES-eGFP; 4. G93A EOM SC-derived myotubes transduced with AAV-CMV-eGFP. Arrows highlight RSMN neurites crossed the microgrooves but stayed close to the edges and grew backward. Also see Figure 8-figure supplement 2A. Boxed regions are enlarged below. Scale bars: 50 μm. (C) Zoom-in views of RSMN neurites innervated or missed the AChR clusters (BTX positive patches). Boxed regions are further enlarged below. Scale bars: 50 μm, 10 μm. (D) Measured lengths of the longest neurites derived from RSMN cells (presumably axons) crossing the 450 μm microgrooves. n = 24, 31, 34,31 cells for the 4 groups, respectively (from 3 rounds of coculture experiments).

By measuring the lengths of the longest RSMN neurites (presumably axons) crossing the 450 μm microgrooves, we noticed that hindlimb-SC derived myotubes from G93A mice, but not from WT littermates, appeared repulsive to distal axons. The axons stayed close to the edge of the microgrooves and even grew backward towards the neuronal compartment, which we termed “edge effect” (Figure 8B and Figure 8-figure supplement 2A, arrows). This was rarely seen in the WT hindlimb SC-derived myotubes. In contrast, RSMN axon extension was dramatically enhanced in the co-culture system either with G93A hindlimb SC-derived myotubes overexpressing Cxcl12, or G93A EOM SC-derived myotubes overexpressing GFP (as control) (Figure 8B, C). Those neurites located close to the microgrooves were often seen to closely grow along nearby myotubes (Figure 8-figure supplement 2B). G93A hindlimb SC-derived myotubes overexpressing Cxcl12 occasionally were even seen to overcome the “edge effect” and redirect the axons towards their proximities (Figure 8-figure supplement 2B, arrow) - a persuasive indication of chemoattractive effects of those myotubes. Meanwhile, RSMN innervation of AChR clusters (BTX positive patches) were not seen in co-cultures with G93A hindlimb SC-derived myotubes overexpressing GFP, while they were consistently observed in the other three types of AAV-transduced myotubes (Figure 8C).

The overall innervation incidence was too low for statistical analysis due to the sparsity of axons in the myotube compartments, thus we conducted co-culture experiments where RSMNs were added to SC-derived myotubes in a uni-compartmental co-culture setting. As shown in Figure 9A-C and Figure 9-figure supplement 1A-C, when co-cultured in a single compartment, Cxcl12 overexpressing G93A hindlimb SC-derived myotubes exhibited increased innervation compared to GFP overexpressing controls. Furthermore, individual Cxcl12 overexpressing EOM SC derived myotubes were frequently innervated by multiple (3 or more) times (Figure 9A-C and Figure 9-figure supplement 1D), recapitulating the “multi-innervation” property of EOMs in vivo [26]. The extra “anchor points” seemed to encourage better spatial alignment of the RSMN axon network to the myotubes globally in the co-cultures. This spatial alignment of RSMN axons and myotubes was not observed in co-cultures of RSMNs and hindlimb SC-derived myotubes (without Cxcl12 overexpression), which also did not exhibit multi-innervation (Figure 9A-C and Figure 9-figure supplement 1A-C). Under higher magnification, we often observed that axons sending branches to adjacent BTX positive patches on EOM SC-derived myotubes (Figure 9B, arrows). Of note, there was higher expression of several Netrin and Semaphorin family members reported to promote axon-branching in EOM SCs compared to other muscle-derived SCs cultured in differentiation medium (Figure 7-table supplement 2) [58–63]. This milieu, comprised of abundant axon guidance molecules, in combination with the robust self-renewability of satellite cells, may contribute to the relative EOM NMJ resistance in ALS.

Coculture of AAV transduced SC derived myotubes with rat spinal motor neurons within the same compartment. (A)

Representative composite images of RSMNs cocultured with: 1. WT HL SC-derived myotubes transduced with AAV-CMV-eGFP; 2. G93A HL SC-derived myotubes transduced with AAV-CMV-eGFP; 3. G93A HL SC-derived myotubes transduced with AAV-CMV-Cxcl12-IRES-eGFP; 4. G93A EOM SC-derived myotubes transduced with AAV-CMV-eGFP. The timeline was the same as that in Figure 8 but RSMNs were seeded on top of myotubes. Boxed regions are enlarged in (B). Scale bars: 50 μm. (B) Zoom-in views of innervated and non-innervated AChR clusters on myotubes having contact with RSMN neurites. Arrows highlight branching phenomena observed in RSMN neurites innervating AChR clusters. Scale bars: 10 μm. (C) Quantification of innervation ratios of AChR clusters (area > 10 μm2) in myotubes in contact with RSMN neurites. n = 37, 48, 40, 52 z-stack images for the 4 groups, respectively (from 3 rounds of coculture experiments). Also see methods for detailed description. *** P < 0.001; **** P < 0.0001; ns, not significant. (D) Graphic summary of the potential mechanisms contributing to EOM NMJ integrity maintenance during ALS progression mimicked by NaBu treatment.

Discussion

We have demonstrated an association between the degeneration of NMJs and depletion of peri-NMJ SCs in hindlimb and diaphragm muscles in vivo in end-stage G93A mice. In contrast, EOM NMJs exhibited no signs of denervation or SCs depletion. The FACS profile of SCs isolated from EOM suggests spontaneous activation even without pathological stimulation, which is supported by previous in vivo BrdU labeling and staining results [64]. Furthermore, immunostaining, qRT-PCR and RNA-Seq analyses all imply that EOM SCs in vitro possess superior renewability better than hindlimb and diaphragm SCs, under differentiation-inducing culture conditions. Comparative functional annotation analyses of the transcriptomes highlight higher expression of axon guidance molecules in EOM SCs. Neuromuscular coculture experiments demonstrate that EOM SC-derived myotubes are more attractive to potential axons derived from rat spinal motor neurons than their hindlimb counterparts – which in fact appeared to repel the branching axons. Thus, the unique homeostasis regulation and axon attraction properties of EOM SCs may contribute to the resistance to denervation of EOM under ALS (Figure 9D).

Earlier report of SC quantification from muscle biopsy samples using electron microscopy detected no significant difference between ALS patients and non-ALS controls [65]. Other studies using primary cultured SCs from muscle biopsy samples implied compromised differentiation program in ALS SCs compared to controls [66,67]. As to proliferation, one study reported enhanced proliferation in ALS SCs [67], while the other cued for deficits in proliferation in ALS SCs [66]. The various findings may reflect variations in the disease stage of patients during sampling. For animal studies, Manzano et al measured the amount of Pax7+ SCs in dissociated EDL and soleus myofibers from G93A mice and WT controls and observed large variations across muscle-type and disease stage [68]. In another study carried out by the same group, primarily cultured EDL and soleus SCs from G93A mice generally showed lower proliferation potential than the WT counterparts [69].

In our whole-mount muscle imaging data and FACS profiles of isolated SCs, SC depletion, at least for the most quiescent population, is observed in hindlimb and diaphragm muscles from end-stage G93A mice. Additionally, the heterogeneity of Vcam1 levels observed in hindlimb and diaphragm SCs derived from end stage G93A mice compared to those from WT controls indicates occurrence of denervation-related pathological activation. Although this activation may transiently elevate the number of SCs, the continuous dominance of asymmetric over symmetric cell division, as well as the differentiation pressure to regenerate myofibers could gradually exhaust the quiescent stem cell pool, leading to the depletion phenotype seen in the end stage G93A mice. In contrast, the superior preservation of renewability of EOM SCs seen in differentiation medium is an effective measure against this catastrophe. However, the spontaneous activation and superior renewability of EOM SCs may not always be an advantage. As mentioned in the introduction section, EOMs are preferentially involved in some other neuromuscular disorders. In the case of mitochondrial myopathies, mitochondrial DNA defects accumulate over time, resulting in compromised oxidative phosphorylation represented by cytochrome c oxidase (COX) deficient myofibers in patients [7,8]. The proportion of COX-deficient myofibers is higher in EOMs than other muscles, which may be related to accelerated spreading of mitochondrial DNA defects in myofibers by the distinct properties of EOM SCs.

In this study, overexpressing Cxcl12 in G93A hindlimb SC-derived myotubes promotes motor neuron axon extension under the co-cultured condition. Cxcl12 has been known as a critical axon guidance cue for motor neurons during development [70,71]. In adult animals, peri-synaptic Schwann cells have recently been reported to transiently produce Cxcl12 after acute motor axon terminal degeneration [54]. Intraperitoneal injection of antibodies neutralizing secreted Cxcl12 protein slows down the regeneration process. In contrast, the exposure of primary motor neurons to recombinant Cxcl12 protein stimulates neurite outgrowth and directional axonal extension in compartmentalized culture setup, which could be reversed by the treatment of the Cxcr4 specific inhibitor AMD3100 [54]. Since myoblasts also express Cxcr4 and will migrate along the Cxcl12 protein gradient during muscle regeneration [72,73], the abundant peri-NMJ SCs in EOM expressing Cxcl12 may achieve synergistic effect with peri-synaptic Schwann cells, improving the long-term integrity of NMJs.

Epigenetic regulation of SC transcriptome by histone deacetylase (HDAC) inhibitors, such as butyrate [74], could modulate the myogenesis process both in vitro and in vivo [48,75–77]. Butyrate is a gut-bacteria fermentation product that also acts as an energy source to improve mitochondrial respiration [74,78]. Previous studies by our lab and others showed NaBu-supplemented feeding prolongs the life span of G93A mice and reverses CypD-related mitoflash phenotypes in myofibers derived from G93A mice, which implies a potential decrease of oxidative stress. Additionally, NaBu treatment also improved mitochondrial respiratory function of NSC34 motor neuron cell line overexpressing hSOD1G93A [36,79,80]. Yet it was not known whether the beneficiary effects of NaBu are linked to any phenotypic changes of SCs, and whether NaBu treatment induced epigenetic modulations similar to that of EOM SCs. The study here revealed its beneficiary role on the preservation of NMJ integrity and peri-NMJ SCs affected in hindlimb and diaphragm muscles by ALS progression. NaBu treatment of G93A hindlimb and diaphragm SCs in culture induced transcriptomic changes mimicking some features seen in EOM SCs, including the upregulation of activation and quiescence signature genes and the downregulation of differentiation signature genes, indicating NaBu improved the renewability of G93A hindlimb and diaphragm SCs. Additionally, some axon guidance molecules, particularly Cxcl12, were also induced by NaBu application (Figure 9D). Thus, NaBu-induced EOM SC-like transcriptomic pattern could be a crucial contributor to its therapeutic effect against ALS progression.

Overall, our studies have defined several mechanisms which (1) may explain why EOMs are relatively spared in ALS, and (2) provides a rationale for targeting these differences as a therapeutic strategy. Furthermore, analysis of these signatures in patient muscle biopsies could be used to select subsets of ALS patients who may be most likely to benefit from therapies targeting this signaling network, and to provide possible “response biomarkers” in pre-clinical and clinical studies. As a cocktail of sodium phenylbutyrate (a derivative of NaBu) and Taurursodiol (an antioxidant) under the Commercial name of Relyvrio has been recently approved in US and Canada as a new ALS therapy [81], our study could help gain more insights into the molecular mechanisms related to this type of ALS therapy.

Materials and methods

Animals

All animal experiments were carried out in accordance with the recommendations in the Guide for the Care and Use of Laboratory Animals of the National Institutes of Health. The protocol on the usage of mice was approved by the Institutional Animal Care and Use Committee of the University of Texas at Arlington (A19.001, approval date: 09/20/2018). WT mice used in this study were of B6SJL background. The ALS transgenic mouse model (hSOD1G93A) with genetic background of B6SJL was originally generated by Drs. Deng and Siddique’s group at Northwestern University and deposited to the Jackson Lab as B6SJL-Tg (SOD1*G93A) [82]. For the butyrate feeding experiment, the animals were fed with 2% NaBu from the age of 3 months (at the ALS disease onset) and lasting for 4 weeks.

Primary culture of satellite cells isolated by fluorescence-activated cell sorting

The hindlimb muscles including tibialis anterior (TA), extensor digitorum longus (EDL), gastrocnemius, and quadriceps were dissected from 1-2 mice. For diaphragm, both the crural and costal domains were dissected from 1-2 mice. For EOMs, the four rectus muscles in the eye sockets were dissected from 5-6 mice [83]. The collected muscles were placed in 0 mM Ca2+ Ringer’s solution. Excessive connective tissue, tendons, fat, blood clots and vessels were removed as much as possible. After cleanup, the muscles were transferred into the digestion media (DMEM containing 5% FBS, 1% Pen/Strep and 10 mM HEPES) and minced into small blocks using scissors. Collagenase II (Worthington LS004176), dispase II (Sigma D46931G), hyaluronidase (Worthington LS002592) and Dnase I (Worthington LS002139) were added at the final concentration of 0.26%, 0.24%, 0.16% and 0.04%, respectively. The digestion system was placed in an orbital shaker running at 70 rpm at 37 °C for 45 min. The digested muscles were triturated 10 times with a 20-gauge needle attached to a 5 ml syringe. Afterwards the triturated mixture was pipetted onto a pre-wetted 100 μm strainer and centrifuge at 100 g for 2.5 min to precipitate the bulky myofiber debris. The supernatant was transferred onto a pre-wetted 40 um strainer and cells were collected by centrifuged at 1200 g for 6.5 min. The cells were resuspended in sorting media (PBS containing 0.3% BSA, 10 mM HEPES and 1 mM EGTA) and stained with antibodies or dyes at 4 °C for 45 min with shaking. The antibodies used include PE anti-mVcam1 (Biolegend 50604858, 600 ng/106 cells), APC anti-mCD31 (Biolegend 50162682, 400 ng/106 cells), APC anti-mCD45 (Biolegend 50162782, 400 ng/106 cells) APC anti-mSca-1 (Biolegend 50163216, 200 ng/106 cells). This antibody combination followed a previous published protocol [40]. Vcam1 is a SC marker, CD31 is a endothelial cell marker, CD45 is a hematopoietic cell marker and Sca-1 is a mesenchymal stem cell marker [40]. For sorting, our target population is VCAM+CD31-CD45-Sca-1-. When needed, the cells were stained with methyl green (2 µg/ml) to check mortality rate. As a positive control, half of the cells were transferred to a new Eppendorf tube and methanol (precooled at −20°C) was added dropwise with gentle mixing and left on ice for 5 min, the dead cells were wash twice before being mixed with the rest half of cells for methyl green staining. The sorting process was conducted using BD FACS Melody Cell Sorter using a 100 µm nozzle. The excitation laser and emission filter setup for methyl green is the same as APC. The flow rate was tuned between 1-7 to adjust the event rate to 2000-4000/sec. Gating conditions of FSC-A/SSC-A, SSC-H/SSC-W, FSC-H/FSC-W, Vcam1/CD31_CD45_Sca-1 and Vcam1/Methyl green plots were shown in Figure 3-figure supplement 1 and Figure 3-figure supplement 2. The sorted cells were collected in growth media (DMEM containing 20%FBS, 1% Pen/Strep, 1% chick embryonic extract and 10 mM HEPES). Afterwards the cells were seeded into laminin (Corning 354232) -coated plates or glass-bottom dishes containing growth media (without HEPES). Culture media was changed on the third day and the fourth day. To induce differentiation, cells were washed three times with PBS and cultured in differentiation media (DMEM containing 2% horse serum) for 2 more days.

Neuro-muscular coculture assay

For neuromuscular coculture in compartmentalized chambers (XonaCHIPS XC450), XoanCHIPS were coated with poly-L-ornithine (0.5 mg/ml) for 1 hr at 37 °C followed by laminin (0.1 mg/ml) for 1 hr at 37 °C. Primarily cultured satellite cells were seeded at 1×10^5 into the myotube compartment in growth medium for attachment overnight and then switched to differentiation medium and cultured for 2 more days. Afterwards AAV8-CMV-eGFP (Vector Biolabs, 1×10^11 GC/ml) or AAV8-CMV-Cxcl12-IRES-eGFP (Vector Biolabs, 1.5×10^11 GC/ml, the Cxcl12 gene is of mouse origin) transduction was conducted for one day. The usage of higher titer for the second virus is for matching GFP signal intensity, as IRES-dependent expression is significantly lower than those directly driven by CMV [84]. The chambers were washed with Neurobasal Plus Medium twice before seeding rat spinal motor neurons into the neuronal compartment (1×10^5). The coculture was kept in Neurobasal Plus Medium for 9 more days and half medium change was conducted every 2 days.

For neuromuscular coculture in the same glass-bottom chamber, 8-well chambered cover glasses (Cellvis C8-1.5H-N) were coated with poly-L-ornithine and laminin as described above. Primarily cultured satellite cells were seeded at 1×10^5/chamber in growth medium for attachment overnight and then switched to differentiation medium and cultured for 2 more days. AAV transduction was conducted for one day as described above. The infected myotubes were washed with Neurobasal Plus Medium (Thermo Fisher A3582901, with B-27 and GlutaMax supplement) before the addition of rat spinal motor neurons (4×10^3/chamber, iXCells 10RA-033) to the same chamber. The coculture was kept in Neurobasal Plus Medium for 9 more days and half medium change was conducted every 2 days.

Immunofluorescence (IF) and imaging of whole mount muscle samples

EDL, soleus and EOM (the four rectus muscles) were dissected out and fixed in 4% paraformaldehyde (PFA) for 1 hr. Afterwards the samples were washed with PBS containing 1% glycine once and two more times with PBS. Further fixation and permeabilization was done with methanol at −20°C for 10 min. For diaphragm, the costal muscles and their surrounding rib cages were dissected and immersed in methanol directly for fixation (PFA fixation was skipped because it made the diaphragm harder to flatten and introduced more non-specific noises in staining). The samples were rehydrated and washed with PBS. The epimysium and tendons were removed as much as possible. EDL and soleus were further split into smaller bundles along the longitudinal axis. Then the samples were immersed in blocking buffer (PBS containing 2% BSA, 2% horse serum, 0.1% Tween-20, 0.1% Triton X-100 and 0.05% sodium azide) overnight at 4 °C. Next day the samples were incubated with primary antibodies for 1-2 days. After washing with PBS for three times, the samples were incubated with Alexa Fluor labelled secondary antibodies (Thermo Fisher 1:1000) and Alexa Fluor 488 conjugated α- Bungarotoxin (Thermo Fisher B13422, 1:1000) for 1-2 days. The samples were then washed with PBS, counterstained with DAPI and mounted in antifade mounting media (20 mM Tris, 0.5% N-propyl gallate, 80% glycerol) for imaging. Primary antibodies used: neurofilament (DSHB 2H3 concentrate, 1:250), synaptophysin (Fisher PA11043, 1:400), Pax7 (SCBT sc-81648, 1:100), Pax7 (Thermo Fisher PA1-117, 1:200), CXCL12 (R&D MAB350, 1:200).

To quantify SC number surrounding NMJs, z-stack images were compacted using maximum projection and circles (75 µm in diameter) were drawn around each NMJ in Image J. The DAPI channel was used as a mask to filter out nonspecific staining of Pax7 in the extracellular matrix.

Immunofluorescence and imaging of cultured myoblasts/myotubes and motor neurons

The cells were fixed in 4% PFA for 10 min and then washed with 1% glycine and PBS. Further fixation and permeabilization was done either with methanol at −20°C for 10 min or 0.25% TritonX-100 for 15 min. The samples were then washed with PBS and immersed in blocking buffer for 40 min. Primary antibody incubation was done at 4 °C overnight. Next day the cells were washed with PBS and incubated with Alexa Fluor labelled secondary antibodies (1:1000) for 2 hours at room temperature, counterstained with DAPI and mounted in antifade mounting media for imaging. Primary antibodies used: Alexa Fluor647 conjugated α-Bungarotoxin (Thermo Fisher B13422, 1:1000), Pax7 (SCBT sc-81648, 1:100), MyoD (SCBT sc-377460, 1:100), Ki-67 (CST 12202S, 1:300), MYH (DSHB MF20 concentrate, 1: 200), neurofilament (DSHB 2H3 concentrate, 1:250), eGFP (Novus NBP-22111AF488, 1:250).

For quantifying innervation ratio, AChR clusters (BTX patches) larger than 10 μm^2 were counted on myotubes that have contact with RSMN neurites. Those myotubes having 0 pixel overlap with RSMN neurites were not included in measurement. The ratio of AChR clusters having overlaps with the RSMN neurites after visual inspection of the whole z-stacks of each view were calculated. The innervation ratio of each view is one dot in the box-and-dot plot in Figure 9C. The measurements were done for 3 rounds of coculture assays.

RNA extraction and qRT-PCR

For cultured SCs, cell culture medium was removed before RNA extraction and Trizol reagent was added to the well before scratching cells off using pipette tips. The homogenate was transferred into tubes containing 5 PRIME Phase Lock Gel (Quantabio 2302830), and 1/10 volume of 1-bromo-3-chloroporpane (Alfa Aesar A11395) was added. The samples were shaken vigorously for 12 s and centrifuged for 15 min at 16,000× g. Afterwards, the upper phase was transferred to another Eppendorf tube and isopropanol (1/2 volume of the initial homogenate) was added for precipitation overnight at −20 °C. Next day, after 15 min centrifugation at 16000 g the precipitate was washed briefly with 75% ethanol. After another 5 min centrifugation, the ethanol was removed. The precipitate was briefly dried and resuspended in RNase free water. For whole muscle samples, the freshly dissected muscles were snap-frozen in liquid nitrogen, kept at −80 °C freezer overnight and then soaked in RNALater ICE (Thermo Fisher AM7030) for 2-3 days at −20 °C. Before RNA extraction, leftover tendons and neurites were removed and muscles were transferred to 1 ml Trizol. Homogenization was conducted in FastPrep-24 Classic bead beating grinder (MP Biomedicals 116004500). The tissue homogenate was transfer to centrifuge tubes containing phase lock gel (QuantaBio 2302830) and 1/10 volume of 1-bromo-3-chloropropane was added. The tubes were hand shaken for 12 seconds, left at bench top for 2 minutes and then centrifuged at 16000 g at 4 °C for 15 minutes. The upper phase was transferred to a new centrifuge tube and mixed with equal volume of ethanol. The following steps of RNA purification was conducted with Direct-zol RNA Miniprep Plus kit (Zymo Research R2070). RNA concentration was measured with Quantus Fluorometer (Promega E6150). RNA quality was assessed by RNA ScreenTape analysis at Genomics Core Facility at the University of North Texas Health Science Center. Only RNAs with RIN ≥ 9 and 28s/18s ≥ 1 were used for qRT-PCR or RNA-Seq. Reverse transcription was done using Promega GoScript Reverse Transcription system. The qPCR reactions used Maxima SYBR Green/ROX qPCR Master Mix (Thermo Fisher) and were carried out by the StepOnePlus Real-Time PCR system (Thermo Fisher). Relative quantification (RQ) of gene expression was generated by ΔΔCt method. The sequences of primers used for qPCR were listed in Table 1.

Bulk RNA-Seq and analysis

Non-stranded mRNA sequencing was carried out by BGI Americas Corporation. The average number of clean reads per sample was 45.16 million (ranging from 44.11 to 48.51 million). The average percentage of clean reads was 93.32% (ranging from 92.42% to 93.97%). The average percentage of aligned reads (to Genome Reference Consortium Mouse Build 38 patch release 6) was 96.01% (ranging from 95.40% to 96.67%) and the average percentage of uniquely aligned reads was 86.46% (ranging from 85.12% to 87.63%). To identify differentially expressed genes (DEGs) between different samples (using Poisson distribution method to compare uniquely mapped reads of the gene), only genes with |log2FC| ≥ 0.4 and FDR ≤ 0.001 were considered. To identify DEGs between different groups (containing more than one sample, using DESeq2 method), only genes with |log2FC| ≥ 0.4 and Qvalue ≤ 0.05 were considered. PCA analysis, hierarchical clustering, functional annotation analysis (KEGG pathway, GO_C, GO_P) were performed in Dr. TOM platform of BGI. Dot plots showing the distribution of activation, differentiation and quiescence signature genes were generated in ggplot2 in R. The BAM files, TPM values will be uploaded to GEO database after acceptance of the manuscript.

Abbreviations

acetylcholine receptor

amyotrophic lateral sclerosis

extensor digitorum longus

extraocular muscle

flexor digitorum brevis

hindlimb

motor neuron

neuromuscular junction

tibialis anterior

sodium butyrate

Supporting information

eLife Assessment

This manuscript describes valuable information on how the extraocular muscles (EOM) are preserved in a mouse model of familial Amyotrophic lateral sclerosis (ALS) that carries a G93A mutation in the Sod1 gene. The authors provide convincing evidence of how the integrity of neuromuscular junction is preserved in EOM but not in limb and diaphragm muscles of G93A mice. Overall, this interesting work provides new evidence regarding the etiopathogenesis of ALS and insights for the development of therapeutic targets to slow the loss of neuromuscular function in ALS.

Reviewer #1 (Public Review):

Summary:

The study explores the mechanisms that preserve satellite cell function in extraocular muscles (EOMs) in a mouse model of familial Amyotrophic lateral sclerosis (ALS) that carries the G93A mutation in the Sod1 gene. ALS is a fatal neuromuscular disorder driven by motor neuron degeneration, leading to progressive wasting of most skeletal muscles but not EOM. The study first established that integrity of neuromuscular junction (NMJ) is preserved in EOM but not in limb and diaphragm muscles of G93A mice, and sodium butyrate (NaBu) treatment partially improves NMJ integrity in limb and diaphragm muscles of G93A mice. They also found a loss of synaptic satellite cells and renewability of cultured myoblasts in hindlimb and diaphragm muscles of G93A mice, but not in EOM, and NaBu treatment restores myoblast renewability. Using RNA-seq analysis, they identify that exon guidance molecules, particularly Cxcl12, are highly expressed in EOM myoblasts, along with more sustainable renewability. Using a neuromuscular co-culture model, they convincingly show that AAV-mediated Cxcl12 expression in G93A myotubes enhances motor axon extension and innervation. Strikingly, NaBu-mediated preservation of NMJ in limb muscles of G93A mice is associated with elevated expression of Cxcl12 in satellite cells and improved renewability of myoblasts. These results together offer molecular insights into genes critical for maintaining satellite cell function and revealing a mechanism through which NaBu ameliorates ALS.

Strengths:

Combination of in vivo and cell culture models.

Nice imaging of NMJ and associated satellite cells.

Using motoneuron-myotube coculture to establish the mechanism.

Tested and illustrated a mechanism through which a clinically used drug ameliorates ALS.

Weaknesses:

Data presentation could be improved (see details in the Recommendation for Authors).

It would have been nice to have included G93A motoneurons in the coculture study.

Reviewer #2 (Public Review):

Summary:

The work is potentially interesting as it outlines the role of satellite cells in supporting the functional decline of skeletal muscle due to the denervation process. In this context the authors analyze the functional and molecular characteristics of satellite cells in different muscle types differently affected by the degenerative process in the ALS model.

Strengths:

The work illustrates a relevant aspect of the differences in stem cell potential in different skeletal muscles in a mouse model of the disease through a considerable amount of data and experimental models.

Weaknesses:

However, there are some criticisms of the structuring of the results:

It is not clear how many animals were used in each experimental group (Figs 1 and 2, Fig. 2-9). In particular, it is unclear whether the dots in the histograms represent biological or technical replicates. Furthermore, the gender used in experimental groups is never specified. This last point appears to be important considering the gender differences observed in the SOD1G93A mouse model.

The first paragraph of the results lacks a functional analysis of the motor decline of the animals after the administration of sodium butyrate. The authors, in fact, administered NaBu around 90 days of age while in previous work the drug had been administered at a pre-symptomatic age. It would therefore be useful, to make the message more effective, to characterize the locomotor functions of the treated animals in parallel with the histological evidence of the integrity of the NMJ.

Figure 5 should be completed with the administration of NaBu also to the satellite cells isolated from the WT mouse, the same for figure 9 where AAV-CMV-Cxcl12 transduction of WT myotubes is missing.

In the experiment illustrated in Figure 8, treatment of cell cultures with NaBu would improve the outcome as well as the interference of Cxcl12 expression in myotubes derived from G93A EOM SC (Fig.9) would strengthen the specificity of this protein in axon guidance in this NMJ typical of a spared muscle in ALS.

In the "materials and methods" section the paragraph relating to the methods used for statistical analysis is missing.

Reviewer #3 (Public Review):

Summary:

In their paper, Li et al. investigate the transcriptome of satellite cells obtained from different muscle types including hindlimb, diaphragm, and extraocular muscles (EOM) from wild-type and G93A transgenic mice (end-stage ALS) in order to identify potential factors involved in the maintenance of the neuromuscular junction. The underlying hypothesis is that since EOMs are largely spared from this debilitating disease, they may secrete NMJ-protective factors. The results of their transcriptome analysis identified several axon guidance molecules including the chemokine Cxcl12, which are particularly enriched in EOM-derived satellite cells. Transduction of hindlimb-derived satellite cells with AAV encoding Cxcl12 reverted hindlimb-derived myotubes from the G93A mice into myotubes sharing phenotypic characteristics similar to those of EOM-derived satellite cells. Additionally, the authors were able to demonstrate that EOM-derived satellite cell myotube cultures are capable of enhancing axon extensions and innervation in co-culture experiments.

Strengths:

The strength of the paper is that the authors successfully isolated and purified different populations of satellite cells, compared their transcriptomes, identified specific factors released by EOM-derived satellite cells, overexpressed one of these factors (the chemokine Cxcl12) by AAV-mediated transduction of hindlimb-derived satellite cells. The transduced cells were then able to support axon guidance and NMJ integrity. They also show that administration of Na butyrate to mice decreased NMJ denervation and satellite cell depletion of hind limbs. Furthermore, the addition of Na Butyrate to hindlimb-derived satellite cell myotube cultures increased Cxcl12 expression. These are impressive results providing important insights for the development of therapeutic targets to slow the loss of neuromuscular function characterizing ALS.

Weaknesses:

Several important aspects have not been addressed by the authors, these include the following points which weaken the conclusions and interpretation of the results.

a) Na Butyrate was shown to extend the survival of G93A mice by Zhang et al. Na butyrate has a variety of biological effects, for example, anti-inflammatory effects inhibit mitochondrial oxidative stress, positively influence mitochondrial function, is a class I / II HDAC inhibitor, etc. What is the mechanism underlying its beneficial effects both in the context of mouse muscle function in the ALS G93A mice and in the in vitro myotube assay? Cytokine quantification as well as histone acetylation/methylation can be assessed experimentally and this is an important point that has not been appropriately investigated.

b) In the context of satellite cell characterization, on lines 151-152 the authors state that soleus muscles were excluded from further studies since they have a higher content of slow twitch fibers and are more similar to the diaphragm. This justification is not valid in the context of ALS as well as many other muscle disorders. Indeed, soleus and diaphragm muscles contain a high proportion of slow twitch fibers (up to 80% and 50% respectively) but soleus muscles are more spared than diaphragm muscles. What makes soleus muscles (and EOMs) more resistant to ALS NMJ injury? Satellite cells from soleus muscles need to be characterized in detail as well.

Furthermore, EOMs are complex muscles, containing many types of fibers and expressing different myosin heavy chain isoforms and muscle proteins. The fact that in mice both the globular layer and orbital layers of EOMs express slow myosin heavy chain isoform as well as myosin heavy chain 2X, 2A, and 2B (Zhou et al., 2010 IOVIS 51:6355-6363) also indicates that the sparing is not directly linked to the fast or slow twitch nature of the muscle fiber. This needs to be considered.

c) In the context of myotube formation from cultured satellite cells on lines 178-179 the authors stained the myotubes for myosin heavy chain. Because of the diversity of myosin heavy chain isoforms and different muscle origins of the satellite cells investigated, the isoform of myosin heavy chain expressed by the myotubes needs to be tested and described. It is not sufficient to state anti-MYH.

d) The original RNAseq results have not been deposited and while it is true that the authors have analyzed the results and described them in Figures 6 and 7 and relative supplements, the original data needs to be shown both as an xls list as a Volcano plots (q value versus log2 fold change). This will facilitate the independent interpretation of the results by the readers as some transcripts may not be listed. As presented it is rather difficult to identify which transcripts aside from Cxcl12 are commonly upregulated. Can the data be presented in a more visual way?

e) There is no section describing the statistical analysis methods used. In many figures, more than 2 groups are compared so the authors need to use an ANOVA followed by a post hoc test.

The authors have achieved their aim in showing that satellite cells derived from EOMs have a distinct transcriptome and that this may be the basis of their sparing in ALS. Furthermore, this work may help develop future therapeutic interventions for patients with ALS.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment

This manuscript describes valuable information on how the extraocular muscles (EOM) are preserved in a mouse model of familial Amyotrophic lateral sclerosis (ALS) that carries a G93A mutation in the Sod1 gene. The authors provide convincing evidence of how the integrity of neuromuscular junction is preserved in EOM but not in limb and diaphragm muscles of G93A mice. Overall, this interesting work provides new evidence regarding the etiopathogenesis of ALS and insights for the development of therapeutic targets to slow the loss of neuromuscular function in ALS.

Public Reviews:

Reviewer#1 (Public Review):

Summary:

The study explores the mechanisms that preserve satellite cell function in extraocular muscles (EOMs) in a mouse model of familial Amyotrophic lateral sclerosis (ALS) that carries the G93A mutation in the Sod1 gene. ALS is a fatal neuromuscular disorder driven by motor neuron degeneration, leading to progressive wasting of most skeletal muscles but not EOM. The study first established that integrity of neuromuscular junction (NMJ) is preserved in EOM but not in limb and diaphragm muscles of G93A mice, and sodium butyrate (NaBu) treatment partially improves NMJ integrity in limb and diaphragm muscles of G93A mice. They also found a loss of synaptic satellite cells and renewability of cultured myoblasts in hindlimb and diaphragm muscles of G93A mice, but not in EOM, and NaBu treatment restores myoblast renewability. Using RNA-seq analysis, they identify that exon guidance molecules, particularly Cxcl12, are highly expressed in EOM myoblasts, along with more sustainable renewability. Using a neuromuscular co-culture model, they convincingly show that AAV-mediated Cxcl12 expression in G93A myotubes enhances motor axon extension and innervation. Strikingly, NaBu-mediated preservation of NMJ in limb muscles of G93A mice is associated with elevated expression of Cxcl12 in satellite cells and improved renewability of myoblasts. These results together offer molecular insights into genes critical for maintaining satellite cell function and revealing a mechanism through which NaBu ameliorates ALS.

Strengths:

Combination of in vivo and cell culture models. Nice imaging of NMJ and associated satellite cells. Using motoneuron-myotube coculture to establish the mechanism. Tested and illustrated a mechanism through which a clinically used drug ameliorates ALS.

Weaknesses:

Data presentation could be improved (see details in the Recommendation for Authors).

It would have been nice to have included G93A motoneurons in the coculture study.

This is indeed a plan of our future study. In the revised version, we discussed the limitation of not including G93A motor neurons in the coculture assay. (Page 11, Line 445-448)

“However, it is possible that motor neurons carrying ALS mutations will respond differently to Cxcl12 mediated axon guidance than WT motor neurons. This is a limitation of the current study which will be investigated in future co-culture studies.”

Reviewer #2 (Public Review):

Summary:

The work is potentially interesting as it outlines the role of satellite cells in supporting the functional decline of skeletal muscle due to the denervation process. In this context the authors analyze the functional and molecular characteristics of satellite cells in different muscle types differently affected by the degenerative process in the ALS model.

Strengths:

The work illustrates a relevant aspect of the differences in stem cell potential in different skeletal muscles in a mouse model of the disease through a considerable amount of data and experimental models.

Weaknesses:

However, there are some criticisms of the structuring of the results:

It is not clear how many animals were used in each experimental group (Figs 1 and 2, Fig. 2-9). In particular, it is unclear whether the dots in the histograms represent biological or technical replicates. Furthermore, the gender used in experimental groups is never specified. This last point appears to be important considering the gender differences observed in the SOD1G93A mouse model.

The original quantification data and mouse gender specification were actually listed in the corresponding supplementary tables. We now added the gender specification and number of the mice used in all corresponding figure legends. The number of mice used for sorting SCs from different muscles were also specified in the Methods section in the revised manuscript. (Page 12, Line 489-493).

We also added one more supplementary figure (Figure 1-figure supplement 2) to compare the innervation status between male and female mice. The following description has been added in the updated manuscript (Page 3-4; Line 125-130):

“The data shown in Figure 1B has also been replotted to compare the innervation status between male and female mice (Figure 1- figure supplement 2). In terms of well- or partially- innervated ratios, there are no significant gender difference observed in our experimental condition, in which the muscle samples were collected at the end stage of the disease, although there is marginally lower “poorly innervated ratio” in the EDL muscle of G93A female mice compared to G93A male mice.”

However, we acknowledge that the current study has limitations to fully detect cross-gender differences in our experiments due to low “n” numbers per gender. We hope this is understandable as we have to split limited resource of ALS G93A mice between different kinds of experiments, including NMJ integrity assessment, peri-nuclear SC abundance assessment, whole muscle-qPCR, cell sorting for imaging, cell sorting for RNA-Seq, cell-sorting for qPCR, cell-sorting for neuromuscular co-culture, etc., in this pioneer study. However, we do intend to gradually build up “n” numbers for characterization of cross-gender difference in our ongoing studies.

As to what the dots in each plot represent, we have inserted the description in each relevant figure legend as detailed below:

For Fig 1, each dot represents quantification result from a single mouse. Please see Figure 1-figure supplement 1, Figure 1-figure supplement 2 and Figure 1-table supplement 1 for NMJs measured per muscle type per gender. Briefly, EDL, soleus and diaphragm muscles were from 4 male and 6 female mice per group; WT EOM group was from 4 male and 4 female mice; G93A EOM group was from 3 male and 4 female mice; G93A EOM with NaBu feeding group was from 6 female mice.

For Fig 2, each dot represents quantification result from a single mouse. Please see Figure 2-table supplement 1 for NMJs measured per muscle type per gender. Briefly, WT EDL group was from 2 male and 2 female mice; G93A EDL group was from 3 male and 3 female mice; G93A EDL with NaBu feeding group was from 2 male and 4 female mice; WT soleus group was from 2 male and 3 female mice; G93A soleus group was from 3 male and 2 female mice; G93A soleus with NaBu feeding group was from 1 male and 4 female mice; WT diaphragm group was from 1 male and 4 female mice; G93A diaphragm group was from 1 male and 4 female mice; G93A diaphragm with NaBu feeding group was from 4 female mice; WT EOM group was from 1 male and 3 female mice; G93A EOM group was from 5 female mice; G93A EOM with NaBu feeding group was from 1 male and 3 female mice.

For Fig 3, each dot in the box-and-dot plots represents result from one round of sorting. WT HL SCs were from 8 male and 6 female mice; G93A HL SCs were from 9 male and 5 female mice; WT diaphragm SCs were from 6 male and 3 female mice; G93A diaphragm SCs were from 12 male and 5 female mice. WT EOM SCs were from 6 batches of male and 1 batch of female mice (each batch contains 5-6 mice of the same gender). G93A EOM SCs were from 5 batches of male and 2 batches of female mice.

*Please note these results were from sorting in which the FACS profiles were recorded. Not all rounds of sorting were with FACS profile recorded.

For Fig 4A, each dot in the box-and-dot plots represents one image analyzed. For WT HL SCs, 94 images from 3 rounds of sorting; For WT Dia SCs, 107 images from 3 rounds of sorting; For WT EOM SCs, 75 images from 3 rounds of sorting; For G93A HL SCs, 96 images from 3 rounds of sorting; For G93A Dia SCs, 62 images from 3 rounds of sorting; For G93A EOM SCs, 79 images from 3 rounds of sorting. For the 3 rounds of sorting, 1 was from male and 2 were from female mice.

*Please note that the number of mice used for sorting SCs in different muscles were specified in the Method Section in the revised manuscript. (Page 12, Line 489-493)

For Fig 4B, each dot in the box-and-dot plots represents one image analyzed. For WT HL SCs, 52 images from 3 rounds of sorting; For WT Dia SCs, 51 images from 3 rounds of sorting; For WT EOM SCs, 51 images from 3 rounds of sorting; For G93A HL SCs, 52 images from 3 rounds of sorting; For G93A Dia SCs, 47 images from 3 rounds of sorting; For G93A EOM SCs, 56 images from 3 rounds of sorting. For the 3 rounds of sorting, 1 was from male and 2 were from female mice.

For Fig 5A, each dot in the box-and-dot plots represents one replicate of culture. HL SCs were from male mice.

For Fig 5B, each dot in the box-and-dot plots represents one image analyzed. For G93A HL SCs, 52 images from 3 rounds of sorting; 1-day NaBu treatment, 45 images from 3 rounds of sorting; 3-day NaBu treatment, 51 images from 3 rounds of sorting; For G93A Dia SCs, 47 images from 3 rounds of sorting; 1-day NaBu treatment, 60 images from 3 rounds of sorting; 3-day NaBu treatment, 57 images from 3 rounds of sorting. For the 3 rounds of sorting, 2 were from male and 1 was from female mice.

For Fig 6, all samples used for bulk RNA-Seq were from female mice.

For Fig 7C, each dot in the box-and-dot plots represents one replicate of culture. RNA samples were collected from 3-6 rounds of sorting and sorted cells were seeded into 3 dishes as replicates. WT HL SCs were from 3 male and 1 female mice. WT diaphragm SCs were from 2 male and 2 female mice; WT EOM SCs were from 3 male mice; G93A HL SCs were from 4 male and 2 female mice. G93A diaphragm SCs were from 1 male and 3 female mice; G93A EOM SCs were from 3 male mice.

For Fig 7D, each dot in the box-and-dot plots represents one replicate of culture. RNA samples were collected from 6 rounds of sorting and sorted cells were seeded into 3 dishes as replicates. G93A HL SCs were from 4 male and 2 female mice; G93A diaphragm SCs were from 2 male and 4 female mice.

For Fig 8D, each dot in the box-and-dot plot represents one neurite measured. HL and EOM SCs used for co-culture experiments were all from male mice.

For Fig 9D, each dot in the box-and-dot plot represents one image analyzed. HL and EOM SCs used for co-culture experiments were all from male mice.

For Figure 1-figure supplement 1, each dot in the box-and-dot plots represents quantification result from one mouse. Please also see Figure 1-table supplement 2. Briefly, muscles in WT and G93A groups were from 3 male and 3 female mice per group; G93A EDL with NaBu feeding group was from 3 male and 3 female mice. G93A soleus with NaBu feeding group was from 2 male and 3 female mice; G93A diaphragm with NaBu feeding group was from 2 male and 4 female mice; G93A EOM with NaBu feeding group was from 4 male and 2 female mice.

The first paragraph of the results lacks a functional analysis of the motor decline of the animals after the administration of sodium butyrate. The authors, in fact, administered NaBu around 90 days of age while in previous work the drug had been administered at a pre-symptomatic age. It would therefore be useful, to make the message more effective, to characterize the locomotor functions of the treated animals in parallel with the histological evidence of the integrity of the NMJ.

We are still in the process of collecting locomotor function data for G93A mice with and without NaBu treatment. We plan to report them in a future manuscript while this manuscript focuses on the molecular and histological aspect. Additionally, in the revised manuscript, we revised the rationale of the NaBu treatment starting after the disease onset. (Page 4, Line 131-134)

“In the previous study, NaBu treatment initiated at a pre-symptomatic age delayed disease progression in G93A mice. As treatment of ALS patients is initiated after symptoms appear, we further tested whether NaBu treatment started after disease onset (at the age of 3 months, 2% NaBu in water for 1 month) was effective in preserving NMJ integrity.”

Figure 5 should be completed with the administration of NaBu also to the satellite cells isolated from the WT mouse, the same for figure 9 where AAV-CMV-Cxcl12 transduction of WT myotubes is missing. We appreciate the reviewer’s suggestion of conducting the additional experiment with AAV-delivery of CXCL12 into the myotubes derived from the WT mice. Extensive studies by other investigators have been performed with butyrate on satellite cells derived from WT mice. To name a few here: Fiszman et al., 1980 (DOI: 10.1016/0014-4827(80)90467-X); Johnston et al., 1992 (DOI: 10.1128/mcb.12.11.5123-5130.1992); Lezzi et al., 2002 (DOI: 10.1073/pnas.112218599). To avoid performing redundant experiments, we focus on the effect of butyrate on the proliferation and differentiation of SCs derived from G93A mice. Thanks to the reviewer’s comment, we added additional discussion in the Results section (Page 6, line 216-217). Regarding the effect of Cxcl12, published studies have demonstrated its role in promoting axon growth. To name a few here: Negro et al., 2017 (DOI: 10.15252/emmm.201607257); Lieberam et al., 2005 (DOI: 10.1016/j.neuron.2005.08.011); Whitman et al., 2018 (DOI: 10.1167/iovs.18-25190). (Page 10, line 434, 440-442).

In the experiment illustrated in Figure 8, treatment of cell cultures with NaBu would improve the outcome as well as the interference of Cxcl12 expression in myotubes derived from G93A EOM SC (Fig.9) would strengthen the specificity of this protein in axon guidance in this NMJ typical of a spared muscle in ALS.

This is a great suggestion. Our study demonstrated the overexpression of CXCL12 in G93A myotube can enhance the axonal guidance and innervation of the co-cultured myotube/moto-neurons. We have also demonstrated the NaBu treatment can enhance the expression of CXCL12 and slow ALS progression. Combining NaBu treatment with CXCL12 overexpression may indeed have additive therapeutic benefits to slow ALS progression. We have added this statement in the revised Discussion. (Page 11, Line 466-468)

In the "materials and methods" section the paragraph relating to the methods used for statistical analysis is missing.

We have added it accordingly. (Page 15, Line 631-636)

Reviewer #3 (Public Review):

Summary:

In their paper, Li et al. investigate the transcriptome of satellite cells obtained from different muscle types including hindlimb, diaphragm, and extraocular muscles (EOM) from wild-type and G93A transgenic mice (end-stage ALS) in order to identify potential factors involved in the maintenance of the neuromuscular junction. The underlying hypothesis is that since EOMs are largely spared from this debilitating disease, they may secrete NMJ-protective factors. The results of their transcriptome analysis identified several axon guidance molecules including the chemokine Cxcl12, which are particularly enriched in EOM-derived satellite cells. Transduction of hindlimb-derived satellite cells with AAV encoding Cxcl12 reverted hindlimb-derived myotubes from the G93A mice into myotubes sharing phenotypic characteristics similar to those of EOM-derived satellite cells. Additionally, the authors were able to demonstrate that EOM-derived satellite cell myotube cultures are capable of enhancing axon extensions and innervation in co-culture experiments.

Strengths:

The strength of the paper is that the authors successfully isolated and purified different populations of satellite cells, compared their transcriptomes, identified specific factors released by EOM-derived satellite cells, overexpressed one of these factors (the chemokine Cxcl12) by AAV-mediated transduction of hindlimb-derived satellite cells. The transduced cells were then able to support axon guidance and NMJ integrity. They also show that administration of Na butyrate to mice decreased NMJ denervation and satellite cell depletion of hind limbs. Furthermore, the addition of Na Butyrate to hindlimb-derived satellite cell myotube cultures increased Cxcl12 expression. These are impressive results providing important insights for the development of therapeutic targets to slow the loss of neuromuscular function characterizing ALS.

Weaknesses:

Several important aspects have not been addressed by the authors, these include the following points which weaken the conclusions and interpretation of the results.

(a) Na Butyrate was shown to extend the survival of G93A mice by Zhang et al. Na butyrate has a variety of biological effects, for example, anti-inflammatory effects inhibit mitochondrial oxidative stress, positively influence mitochondrial function, is a class I / II HDAC inhibitor, etc. What is the mechanism underlying its beneficial effects both in the context of mouse muscle function in the ALS G93A mice and in the in vitro myotube assay? Cytokine quantification as well as histone acetylation/methylation can be assessed experimentally and this is an important point that has not been appropriately investigated.

Great suggestion by the reviewer.

Our previous publications (DOI: 10.3390/biom12020333; DOI: 10.3390/ijms22147412) have shown the beneficial roles of NaBu in ameliorating mitochondrial function in both motor neuron-like cells and adult muscle fibers. A focus of the current study is to test whether NaBu treatment also affect the SCs by regulating their gene transcription. Regarding the potential on HDAC/acetylation modification, there are previous studies by other investigators. We have added these references in the Discussion (Page 11, line 466-468).

(b) In the context of satellite cell characterization, on lines 151-152 the authors state that soleus muscles were excluded from further studies since they have a higher content of slow twitch fibers and are more similar to the diaphragm. This justification is not valid in the context of ALS as well as many other muscle disorders. Indeed, soleus and diaphragm muscles contain a high proportion of slow twitch fibers (up to 80% and 50% respectively) but soleus muscles are more spared than diaphragm muscles. What makes soleus muscles (and EOMs) more resistant to ALS NMJ injury? Satellite cells from soleus muscles need to be characterized in detail as well.

We agree with the reviewer’s comment that our original statement is misleading regarding the difference between soleus and diaphragm muscles in terms of the content of slow twitch fibers. Our histological studies revealed similar defects in denervation of diaphragm and soleus muscles derived from the G93A mice. Most importantly, the degree of NMJ degeneration and atrophy is less severe in soleus compared to other hindlimb muscles, such as EDL, during ALS progression. We have cited related studies such as Valdez et al., 2012 (DOI: 10.1371/journal.pone.0034640), Atkin et al., 2005 (DOI: 10.1016/j.nmd.2005.02.005). To avoid any confusion, we have removed the original statement and revised the paragraph (Page 4, line 159-162).

“The three groups were determined because they represent the most severely affected, moderately affected and least affected muscles by ALS progression, respectively. Soleus was not included in the hindlimb SCs pool because its less affected than other hindlimb muscles based on our study and others [6,42].”

Furthermore, EOMs are complex muscles, containing many types of fibers and expressing different myosin heavy chain isoforms and muscle proteins. The fact that in mice both the globular layer and orbital layers of EOMs express slow myosin heavy chain isoform as well as myosin heavy chain 2X, 2A, and 2B (Zhou et al., 2010 IOVIS 51:6355-6363) also indicates that the sparing is not directly linked to the fast or slow twitch nature of the muscle fiber. This needs to be considered.

We greatly appreciate your suggestions and have included these points in the revised Discussion. “It is known that EOMs are complex muscles. Besides the developmental myosin isoforms, EOMs also express both adult fast and slow myosin contractile elements (Zhou et al., 2010 IOVIS 51:6355-6363), suggesting that the sparing may not be solely linked to the fast or slow twitch nature of the muscle fiber, rather the changes in SCs may play a pivotal role in preserving the EOM function during the progression of ALS. ” (Page 9, line 389-392)

(c) In the context of myotube formation from cultured satellite cells on lines 178-179 the authors stained the myotubes for myosin heavy chain. Because of the diversity of myosin heavy chain isoforms and different muscle origins of the satellite cells investigated, the isoform of myosin heavy chain expressed by the myotubes needs to be tested and described. It is not sufficient to state anti-MYH.

We used the pan-anti-MYH antibody (MF20 from DSHB) for the immunostaining of myosin heavy chain for identification of the differentiated myotubes. As described in the commercial website: https://dshb.biology.uiowa.edu/MF-20), FM20 recognizes all myosin heavy chain isoforms. We are happy to examine whether specific myosin heavy chain isoforms may contribute to the differences observed in future studies.

(d) The original RNAseq results have not been deposited and while it is true that the authors have analyzed the results and described them in Figures 6 and 7 and relative supplements, the original data needs to be shown both as an xls list as a Volcano plots (q value versus log2 fold change). This will facilitate the independent interpretation of the results by the readers as some transcripts may not be listed. As presented it is rather difficult to identify which transcripts aside from Cxcl12 are commonly upregulated. Can the data be presented in a more visual way?

We have uploaded the Fastq files and the text files containing TPM values to the Gene Expression Omnibus (GEO) database and included the GEO access number GSE249484 in the revised text. Per recommendation of the reviewer, we have added supplementary tables for Figure 6, to list the top 20 differentially expressed genes (ranked by Log2FC, both the upregulated and downregulated) comparing 1) EOM SCs to their hindlimb and diaphragm counterparts (Figure 6-table supplement 1); 2) G93A SCs to WT SCs of the same muscle origin (Figure 6-table supplement 2); 3) G93A hindlimb and diaphragm SCs with 3 day-NaBu treatment to those without (Figure 6-table supplement 3). (Page 6, Line 237-257)

(e) There is no section describing the statistical analysis methods used. In many figures, more than 2 groups are compared so the authors need to use an ANOVA followed by a post hoc test.

Thank for the comments. We have added it accordingly. (Page 15, Line 631-636)

The authors have achieved their aim in showing that satellite cells derived from EOMs have a distinct transcriptome and that this may be the basis of their sparing in ALS. Furthermore, this work may help develop future therapeutic interventions for patients with ALS.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

The prevailing hypothesis of ALS is that motoneuron degeneration subsequently induces muscle atrophy and wasting. However, evidence also suggests that ALS is a muscle disease independent of motoneuron degeneration. The results from the current study support the latter. The RNA-seq data from cultured myoblasts (without innervation) suggest cell cell-autonomous effect of G93A on muscle cells. While the current analyses in this study identify axon guidance pathways in EOM satellite cells that may underlie their unique gene program that enhances motoneuron function, the powerfulness of the RNA-seq data is underutilized. I suggest that the authors explore the RNA-seq further by comparing genes and pathways altered by G93A in various muscles to better pinpoint how G93A influences satellite cell function.

Thanks for the comments and advice. Further analysis of the RNA-seq data is planned. As our original sequencing provider became unavailable to us since last year, we are currently negotiating with other sequencing providers. We have deposited the raw data files into the GEO database (GSE249484) to foster further analyses by other researcher teams.

To address the reviewer’s concern, we have added three more supplementary tables for Figure 6, which list the top 20 differentially expressed genes (DEG) (ranked by Log2FC, both the upregulated and downregulated) comparing 1) EOM SCs to their hindlimb and diaphragm counterparts (new Figure 6-table supplement 1); 2) G93A SCs to WT SCs of the same muscle origin (new Figure 6-table supplement 2); 3) G93A hindlimb and diaphragm SCs with and without 3 day-NaBu treatment (new Figure 6-table supplement 3). These three DEG lists are discussed in the results section of the revised manuscript as following (Page 6, Line 237-257).

Figure 4 presentation could be improved by adopting a similar comparison (WT vs G93A) as used in Figure 1-3. The current comparison is not straightforward. In addition, a magnified image of panel A would demonstrate the loss of myoblast homeostasis more clearly. (AKA Figure 2B)

The WT vs G93A comparison was presented in the supplementary figure of Figure 4 (Figure 4-figure supplement 1 in the previous version, and now in Figure 4-figure supplement 2 in the revised version).

As requested, we have added magnified single channel representative images of cultured SCs in the new Figure 4-figure supplement 1 in the revised manuscript.

Co-culture results in Figure 8 are very impressive. It would be nice if the data were quantified. The figure legend states that panel D is the quantification, but I don't see panel D. As the study used rat motoneurons (presumably SOD1 wildtype), it is unknown if G93A motoneurons would respond to muscle-derived CXCL12 similarly to the wildtype motoneurons. This information is crucial for understanding whether the SOD1 mutant ALS1 is a motoneuron disease or muscle disease or both. Some discussion should be provided to reflect the limitation (of not including G93A motoneurons in the coculture).

Panel D (the quantification data) was presented in the original figure setting (but may not be obvious). We have now revised Figure 8 to enlarge panel D to clearly present the quantification data.

We acknowledge the limitation of not including mutant G93A motor neurons in the coculture assay, and have added this important point (and our future plans to do so) in the discussion section of the revised manuscript: (Page 11, Line 445-448)

“However, it is possible that motor neurons carrying ALS mutations may respond differently to Cxcl12 mediated axon guidance than WT motor neurons. This is a limitation of the current study, which will be investigated in future co-culture studies.”

Reviewer #2 (Recommendations For The Authors):

Line 108. The sentence: "Z-stack scans of glycerol-cleared 109 whole muscles were obtained using a high working distance lens in a confocal microscope. The z-stacks were compacted into 2D images by maximal intensity projection" and should be moved to the material and methods section.

Removed from the Result section and added to the Method section as recommended (Page 13, Line 564-568).

Linea 113. The sentence: " In order to quantify the extent of denervation in a categorical manner, NMJs were arbitrarily defined as "well innervated" if SYP staining was present in >60% of the BTX positive area, "partially innervated" if between 60% and 30%, and "poorly innervated" if SYP staining corresponded to less than 30% of the BTX positive area" has already been written in the figure legend.

Thanks for the advice. We have rephrased the sentence to remove the redundant part.

In lines 445-7, it would be better to indicate the enzymatic units instead of the concentrations.

We included enzymatic units for the four enzymes in the Methods Section of revised manuscript (Page 12, Line 497-499).

Reviewer #3 (Recommendations for The Authors):

There are several points that need to be addressed by the authors including:

(a) The authors need to provide experimental evidence as to the mode of action of Na Butyrate and more specifically whether its beneficial effect is mediated by its anti-inflammatory action, inhibition of HDACs, or the combination of several mechanisms. Additionally, it should be clearer why Na Butyrate was administered. The sentence referring to reference 36 is not sufficient and some mechanistic insight needs to be provided in the results section.

Thanks for the great suggestion. We have revised the Results section accordingly to clarify the rationale for NaBu usage (please also see our detailed response to your suggestion above). (Page 4, line 131-134)

(b) Their reason for excluding soleus-derived-satellite cells from the analysis is not valid. Soleus muscles are "more" speared than diaphragm muscles and analysis may help shed light on this observation.

Please see our response to your question (b) in the above public review section.

(c) DATA AVAILABILITY: The RNAseq raw untransformed data has not been provided and Volcano plots are also not shown. I find it quite difficult to follow the results of the RNAseq experiments and this is central to the interpretation of the paper's results. Ideally, one should be able to look at the data and draw his/her own conclusions but as it stands this is difficult to do.

We have uploaded the raw FastQ files and the excel files containing TPM values to the GEO database with the access number GSE249484.

(d) A detailed description of all statistical tests that were used needs to be provided.

Yes, this has been added to the revised manuscript.

(e) Many figure legends are incomplete and some panels are not described appropriately, indicating that the authors need to thoroughly revise all aspects of the manuscript.

We have extensively edited the figure legends to address the issues raised by reviewers.

(f) Line 96-98: it is unlikely that muscles from ALS patients will be biopsied frequently. Furthermore, what biomarkers exactly could be followed in patients in response to therapy? This is unclear.

While it is true that it is not generally part of the diagnostic workup for ALS, muscle biopsy is increasingly being used pre- and post-treatment in ALS clinical trials to examine responses to potential new therapies. Muscle biopsy is also being explored in several ongoing studies as a potential ALS-relevant peripheral tissue amenable to biopsy (as opposed to brain or spinal cord) for predictive, pharmacodynamic, and prognostic biomarkers. This includes studies attempting to recapitulate pathophysiological patient clusters observed in CNS autopsy tissues and studies to detect aberrant TDP-43 aggregates in intramuscular nerve twigs, among others. Indeed, Dr. Ostrow’s clinical duties include performing muscle biopsies and interpreting muscle pathology, and he is involved in several ongoing studies attempting to correlate postmortem CNS and muscle analyses for these purposes.

To avoid potential controversy on the feasibility of multiple biopsies, we rephrased the sentence as follows (Page 3, Line 96-98)

“Characterizing the distinct EOM SC transcriptomic pattern could provide clues for identifying potential biomarkers in therapeutic trials in both ALS patients and animal models, in addition to identifying therapeutic targets.”

(g) Line 388-389. What do the authors mean by this sentence? It is not clear.

Thanks for the comment, we have expended the discussion to make it clearer in the revision. (Page 10, Line 428-431)

“It is possible that the more frequent self-renewal and spontaneous activation of EOM SCs contribute to higher rate of mitochondrial DNA replication, leading to accelerated spreading of mitochondrial DNA defects, resulting in higher proportion of COX-deficient myofibers than other muscles”.

(h) Were the experimenters blinded as to the results shown in Figures 2, 7, 8, and 9?

We endeavored to blind experiments whenever possible. Not all experiments were blinded due to logistic complexity and the clear difference in microscopic and gross appearances of wild-type and mutant muscle. The differences observed in Figures 2, 7, 8, 9 are qualitative (ie more than just quantitative), which should minimize the impact of possible human bias. Additionally, we employed multiple different experimental approaches to assess our hypotheses.

For Fig 2, the physical appearance is notably different between G93A and WT muscles. The different innervation status (Fig 2A) is also not amenable to blinding.

For Fig 7, the expression level of Hmga2, Notch3 and Cxcl12 detected by the qPCR assay are substantially greater in EOM derived SCs than counterparts from other muscles, and these results are also consistent with RNA-Seq, immunofluorescence assays. For Fig 8, the overexpression of Cxcl12 and the coculture with EOM SC derived myotubes not only increased the length of the longest neurites but also promoted axon branching, which can be easily observed.

For Fig 9, only the EOM SC derived myotubes were capable of aligning the neurites along with them on a global scale. This qualitative difference is easy to appreciate, even under low magnification.

(i) Line 64 -65 The authors refer to a very old paper by Fischer et al in 2002 for the expression profile of EOMs. There are more recent papers including that of Eckhardt et al. (eLife 2023, 12:e83618) showing the differences in proteome between EOMs and soleus and EOMs and EDL muscles. There are more than 2000 (and not 300!!) differentially expressed proteins.

Thank you for the newly published reference. We have revised the Introduction section to include this new proteomic study. (Page 2, Line 64-69)

(j) Figure 7 C. The Y axis is mislabeled as they should be log2 fold change and not the growth conditions.

Thank you for catching this. We have fixed it.

(k) In all figures, if each symbol represents the results obtained on 1 mouse, this needs to be clearly stated. What do the panels on the right of Figures 4 and 5B show?

Thanks for the comments. For Figure 1B and 2C, as well as Figure 1-figure supplement 1B, one dot in the box-and-dot plots represents result obtained from 1 mouse. For Figure 3B, one dot represents one round of sorting. Generally, one mouse was euthanized for each round of sorting for HL and diaphragm SCs. But the sorting of EOM SCs could take up to 6 mice (as the EOMs are much smaller). For Figure 4 and 5B, each dot represents one image analyzed. All images were collected from three rounds of sorting. For Figure 5A, each dot represents one replicate of culture. For Figure 5B, each dot represents one image analyzed. All images were collected from three rounds of sorting. We have indicated those details in the revision.

Please also see our response to the 1st question of Reviewer 2 in the public review section.

(l) Figure 6 Table supplement 3 does NOT show the FDR but only the log2 fold change. Please amend.

We have amended the supplementary table accordingly.

Version 2

Abstract

Abstract

Amyotrophic lateral sclerosis (ALS) is a fatal neuromuscular disorder characterized by progressive weakness of almost all skeletal muscles, whereas extraocular muscles (EOMs) are comparatively spared. While hindlimb and diaphragm muscles of end-stage SOD1G93A (G93A) mice (a familial ALS mouse model) exhibit severe denervation and depletion of Pax7+satellite cells (SCs), we found that the pool of SCs and the integrity of neuromuscular junctions (NMJs) are maintained in EOMs. In cell sorting profiles, SCs derived from hindlimb and diaphragm muscles of G93A mice exhibit denervation-related activation, whereas SCs from EOMs of G93A mice display spontaneous (non-denervation-related) activation, similar to SCs from wild-type mice. Specifically, cultured EOM SCs contain more abundant transcripts of axon guidance molecules, including Cxcl12, along with more sustainable renewability than the diaphragm and hindlimb counterparts under differentiation pressure. In neuromuscular co-culture assays, AAV-delivery of Cxcl12 to G93A-hindlimb SC-derived myotubes enhances motor neuron axon extension and innervation, recapitulating the innervation capacity of EOM SC-derived myotubes. G93A mice fed with sodium butyrate (NaBu) supplementation exhibited less NMJ loss in hindlimb and diaphragm muscles. Additionally, SCs derived from G93A hindlimb and diaphragm muscles displayed elevated expression of Cxcl12 and improved renewability following NaBu treatment in vitro. Thus, the NaBu-induced transcriptomic changes resembling the patterns of EOM SCs may underlie the beneficial effects observed in G93A mice. More broadly, the distinct transcriptomic profile of EOM SCs may offer novel therapeutic targets to slow progressive neuromuscular functional decay in ALS and provide possible "response biomarkers" in pre-clinical and clinical studies.

Introduction

Amyotrophic lateral sclerosis (ALS) is a fatal disorder characterized by progressive motor neuron (MN) loss and skeletal muscle wasting, resulting in respiratory failure and death 3–5 years following diagnosis [1,2]. While not entirely spared, the extraocular muscles (EOMs) of ALS patients and mutant mouse models exhibit comparatively preserved structure and function and retained neuromuscular junctions (NMJ) [3-6]. The reason for this differential involvement remains unclear, especially considering that eye movements are preferentially involved in other neuromuscular disorders like myasthenia gravis, some muscular dystrophies, and mitochondrial myopathies [7,8].

While ALS is classically considered predominantly a “dying-forward” process of motor neurons (MN), accumulated evidence [9-17] support that early muscle dysfunction is an important contributor to ALS pathophysiology. The degree to which “dying-forward” from the central nerve system or “dying-back” process [18-20] starting distally at NMJs contributing to ALS progression remains unsettled and may vary in different patient subsets. Regardless of the direction of communication and relative contribution of different cell types to ALS progression (e.g. glia, neurons, myofibers) [20], it is widely accepted that NMJ disassembly and the resulting skeletal muscle denervation is a critical pathogenic event in both patients and animal models [21-24].

EOMs differ from limb and trunk muscles in many respects [25]. In contrast to other skeletal muscles, the individual motor neurons (MNs) controlling EOMs innervate fewer EOM myofibers, and individual myofibers often have multiple NMJs [26]. The multiple innervation phenotype refers to the co-existence of innervated "en plaque" and "en grappe" NMJs in a single EOM myofiber. The "en plaque" type of NMJs are located in the middle portion of myofibers and resemble NMJs seen in other types of muscles. Each EOM myofiber has one "un plaque" NMJ. "en grappe" NMJs are a series of smaller, grape-shaped NMJs that are located in the distal portion of some EOM myofibers [3,26]. MNs innervating EOMs express higher level of the Ca2+ binding proteins calbindin-D28k and parvalbumin, thus may be more resistant to the Ca2+-induced excitotoxicity [27-29]. Pronounced differences exist between EOM and limb muscles in their transcriptomic profiles [30]. In addition, a recent proteomic study revealed more than 2500 proteins differentially expressed between EOM and EDL muscles and more than 2000 between EOM and soleus muscles [31]. However, none of muscle-derived neuroprotective factors, including BDNF, GDNF, NT3 and NT4 were expressed (RNA or proteins) significantly higher in EOM myofibers compared to hindlimb myofibers in both mutant G93A and wild-type mice [31-33]. Thus, it is important to look for clues in other cell types that influence the microenvironment around NMJs, such as satellite cells (SCs).

An association between SC function, NMJ integrity, and resultant denervation has been demonstrated in animal models of aging and muscular dystrophy [34,35]. Using transgenic mice expressing GFP in the progeny of activated SCs, Liu and colleagues demonstrated that inducible depletion of adult skeletal muscle SCs impaired the regeneration of NMJs [34]. Furthermore, they found that the loss of SCs exacerbated age-related NMJ degeneration. Preservation of SCs attenuated aging-related NMJ loss and improved muscle performance in mice [35]. SCs from EOMs exhibit superior expansion and renewability (the capability to maintain dividing and undifferentiated stem cells after rounds of regeneration) than their limb counterparts [36], inspiring us to compare the phenotypes of SCs derived from EOMs and limb muscles and to conduct transcriptomic profiling of SCs to look for candidate genes that could preserve NMJ integrity in ALS.

In the current study, we found that the integrity of NMJs and the amount of Pax7+ SCs are maintained in EOMs in end-stage G93A mice, in contrast to the severe denervation and depletion of Pax7+ SCs in hindlimb and diaphragm muscles. Transcriptomic profiling revealed that the EOM SCs exhibit abundant transcripts of axon guidance molecules, such as Cxcl12, along with sustainable renewability compared with SCs from diaphragm and hindlimb muscles. Furthermore, EOM SC-derived myotubes enhance axon extension and innervation in co-culture experiments with spinal motor neurons compared to hindlimb SC-derived myotubes. Overexpressing Cxcl12 using adeno-associated viral vector (AAV8) in G93A hindlimb SC-derived myotubes recapitulated features of the EOM SC-derived myotubes.

Butyrate, a short-chain fatty acid produced by bacterial fermentation of dietary fibers in the colon, has been shown to extend the survival of ALS G93A mice [37]. We found that G93A mice fed with 2% sodium butyrate (NaBu) in drinking water decreased NMJ denervation and SC depletion in hindlimb and diaphragm muscles. Adding NaBu to the culture medium improved the renewability of SCs derived from the G93A hindlimb and diaphragm muscles, as well as increasing the expression of Cxcl12, resembling EOM SC transcriptomic pattern. Characterizing the distinct EOM SC transcriptomic pattern could provide clues for identifying potential biomarkers in therapeutic trials in both ALS patients and animal models, in addition to identifying therapeutic targets.

Results

NMJ integrity and peri-NMJ SC abundance are preserved in EOMs, and NaBu-supplemented feeding reduces NMJ denervation and SC depletion in hindlimb and diaphragm muscles in G93A mice

We investigated the NMJ integrity and peri-NMJ SC abundance by whole-mount imaging of extensor digitorum longus (EDL), soleus, diaphragm, and EOMs derived from WT littermates and G93A mice with or without NaBu-supplemented feeding. Acetylcholine receptors (AChRs) at NMJs were labeled with Alexa Fluor conjugated α-Bungarotoxin (BTX). Axons and axon terminals of motor neurons were labeled with antibodies to neurofilament (NF) and synaptophysin (SYP) respectively (Figure 1A and Figure 1-figure supplement 1A). In order to quantify the extent of denervation in a categorical manner, NMJs were arbitrarily defined as “well innervated”, “partially innervated”, and “poorly innervated” (Figure 1-figure supplement 1A). Poorly innervated NMJs also generally lacked nearby NF+ axons. For EOMs, only the “en plaque” type of NMJs were included in this comparison, because they resemble NMJs in other types of muscles in morphology and spatial distribution. The “en grappe” type of NMJs were not included.

Characterizing NMJ integrity in muscles dissected from wild-type controls and end-stage G93A mice with or without NaBu feeding.

(A) Representative compacted z-stack scan images of whole-mount EDL, soleus diaphragm extraocular muscles stained with antibodies against neurofilament medium chain (NF, labeling axons), synaptophysin (SYP, labeling axon terminals) and Alexa Fluor 488 conjugated α-Bungarotoxin (BTX, labeling AChRs on muscle membrane). scale bars, 50 μm. (B) Mean ratios of well innervated NMJs (SYP signals are present in >60% of BTX positive area), partially innervated NMJs (SYP signals are present in 30-60% of BTX positive area) and poorly innervated NMJs (SYP signals are present in <30% of BTX positive area) and in different types of muscles dissected from WT controls and end-stage G93A mice with or without NaBu feeding (see Figure 1-figure supplement 1, Figure 1-figure supplement 2 and Figure 1-table supplement 1 for NMJs measured per muscle type per gender. Briefly, EDL, soleus and diaphragm muscles were from 4 male and 6 female mice per group; WT EOM group was from 4 male and 4 female mice; G93A EOM group was from 3 male and 4 female mice; G93A EOM with NaBu feeding group was from 6 female mice). Each dot in the box-and-dot plots represents quantification result from a single mouse. * P < 0.05; ** P < 0.01; **** P < 0.0001; ns, not significant (t-test). ANOVA P values are also shown.

The relative proportions of well-, partially- and poorly-innervated NMJs in different muscle origins of WT and G93A mice are shown in Figure 1B (including both genders). At least 343 (average 758 ± 345 per group) NMJs from muscles (EDL, soleus, diaphragm and EOMs, respectively) dissected from WT or G93A mice with or without NaBu-feeding were examined. The original mouse information (including gender specification) and quantification data are listed in Figure 1-table supplement 1. EDL, soleus and diaphragm muscles, but not EOMs, from end stage G93A mice (4-month-old) exhibited significant decrease in well-innervated NMJs, accompanied by an increase in poorly-innervate NMJs. The increase of “poorly-innervated NMJ” is slightly less dramatic in diaphragm compared to the hindlimb muscles, meanwhile the increase of “partially-innervated NMJ” ratio is the most significant in the diaphragm (Figure 1B), indicating that denervation process is most severe in hindlimb muscles, slightly milder in diaphragm and barely detectable in EOMs. The data showed in Figure 1B have also been replotted to compare the innervation status between male and female mice (Figure 1-figure supplement 2). In the term of well- or partially-innervated ratios, there are no significant cross-gender difference observed in our experimental condition, in which the muscle samples were collected at the end stage of the disease, although there is marginally lower “poorly innervated ratio” in the EDL muscle of G93A female mice compared to G93A male mice.

In the previous study, the NaBu treatment of G93A mice started at the pre-symptomatic age (2∼2.5-month-old) significantly delayed ALS progression in G93A mice [37]. As the treatment for ALS patients is initiated after symptoms appear, we further tested whether NaBu treatment started after the disease onset (at the age of 3 months, 2% NaBu in water for 1 month) was effective in preserving NMJ integrity. Consistent with previous observation, the hindlimb and diaphragm muscles from NaBu-fed mice (1 month, 2% in water, starting feeding at the age of 3-month-old) all exhibited significantly reduced denervation compared to littermates who were not fed NaBu (Figure 1B, n = 10 per group, gender matched).

qRT-PCR for Scn5a, the gene encoding sodium channel Nav1.5, was performed with RNA extracted from whole muscles and used as another marker of denervation [38-40] (Figure 1-figure supplement 1B). The original values of Scn5a expression levels are provided in Figure 1-table supplement 2. Consistent with the immunostaining quantification in Figure 1A, the elevation of Scn5a expression is most notable when comparing G93A hindlimb muscles to wildtype counterparts (>250 folds for EDL, >70 folds for soleus), milder for diaphragm (> 15 folds) and not significant for EOMs. Additionally, NaBu feeding significantly reduced the Scn5a transcripts in hindlimb muscles (Figure 1-figure supplement 1B).

Whole mount imaging of muscles stained with the satellite cell (SC) markers Pax7, BTX, SYP, and DAPI, was used to quantify SCs in the vicinity of NMJs. The Pax7 signal in each z-stack image was filtered using DAPI as a mask to identify Pax7+ cell nuclei. Afterwards the z-stacks were compacted into 2D images by maximal intensity projection (Figure 2A). The number of nuclear Pax7+ cells were counted within the circles of 75 µm diameter drawn around the NMJs (Figure 2B, see also Figure 2-table supplement 1). Similar to the NMJ denervation status, depletion of peri-NMJ SCs was most significant in hindlimb muscles of end-stage G93A mice and slightly milder in diaphragm, while no significant depletion was observed in EOMs (Figure 2C). G93A mice fed with NaBu had less depletion of SCs in hindlimb and diaphragm muscles.

Quantification of peri-NMJ SC abundance in muscles dissected from wild-type controls and end-stage G93A mice with or without NaBu feeding.

(A) Representative compacted z-stack scan images of whole-mount EDL, soleus diaphragm extraocular muscles stained with antibodies against Pax7 (labeling SCs), SYP and BTX-Alexa Fluor 488. Scale bars, 50 μm. (B) To measure the abundance of SCs around NMJs, circles of 75 μm diameter were drawn around the NMJs and the number of nuclear Pax7+ cells (co-localized with DAPI) were counted. Scale bars, 10 μm. (C) Mean number of SCs around NMJs in different types of muscles dissected from WT controls and end-stage G93A mice with or without NaBu feeding (see Figure 2-table supplement 1 for NMJs measured per muscle type per gender. Briefly, WT EDL group was from 2 male and 2 female mice; G93A EDL group was from 3 male and 3 female mice; G93A EDL with NaBu feeding group was from 2 male and 4 female mice; WT soleus group was from 2 male and 3 female mice; G93A soleus group was from 3 male and 2 female mice; G93A soleus with NaBu feeding group was from 1 male and 4 female mice; WT diaphragm group was from 1 male and 4 female mice; G93A diaphragm group was from 1 male and 4 female mice; G93A diaphragm with NaBu feeding group was from 4 female mice; WT EOM group was from 1 male and 3 female mice; G93A EOM group was from 5 female mice; G93A EOM with NaBu feeding group was from 1 male and 3 female mice). Each dot in the box-and-dot plots represents quantification result from a single mouse. * P < 0.05; ** P < 0.01; **** P < 0.0001; ns, not significant (t-test). ANOVA P values are also shown.

The proliferation and differentiation properties are conserved in SCs derived from G93A EOMs

We conducted FACS based isolation of SCs from hindlimb (combining tibialis anterior, EDL, quadriceps and gastrocnemius), diaphragm, and EOMs using a modified version of a published protocol [41]. The three groups were determined because they represent the most severely affected, moderately affected and least affected muscles by ALS progression, respectively. Soleus was not included in the hindlimb SCs pool because its less affected than other hindlimb muscles based on our study and others [6,42]. To separate different cell populations, Vcam1 was used as a marker for quiescent SCs, while CD31 (endothelial cell marker), CD45 (hematopoietic cell marker) and Sca-1 (mesenchymal stem cell marker) positive cells were discarded. Thus, in the FACS profile the target SC population (Vcam1+CD31-CD45-Sca-1-) was in gate P5-4 (Figure 3A). Single antibody staining controls are shown in Figure 3-figure supplement 1. We also applied methyl-green (20 µg/ml) staining [43] to evaluate the viability of cells (Figure 3-figure supplement 2), only about 1% of Vcam1 positive population were methyl green stained (Figure 3-figure supplement 2D, Events P5-2 / (P5-2 + P5-4)*100%). Thus, most cells were alive when loaded into the flow cytometer.

Isolation of SCs from wildtype and G93A muscles using fluorescence activated cell sorting.

(A) Representative FACS profiles during the isolation of Vcam1+CD31-CD45-Sca1- cells (gate P5-4). The corresponding population hierarchies are shown right to the plots. For single antibody staining controls, unstained controls and viability test, see Figure 3-figure supplement 1 and Figure 3-figure supplement 2. (B) Comparing the percentage of events in the P5-4 gate between cells isolated from WT and G93A mice. Each dot in the box-and-dot plots represents result from one round of sorting. WT HL SCs were from 8 male and 6 female mice; G93A HL SCs were from 9 male and 5 female mice; WT diaphragm SCs were from 6 male and 3 female mice; G93A diaphragm SCs were from 12 male and 5 female mice. WT EOM SCs were from 6 batches of male and 1 batch of female mice (each batch contains 5-6 mice of the same gender). G93A EOM SCs were from 5 batches of male and 2 batches of female mice. * P < 0.05; *** P < 0.001; ns: not significant (t-test).

SCs isolated from WT hindlimb and diaphragm muscles exhibited relatively homogenous levels of Vcam1, which facilitated separation from other cells and debris (Figure 3A, gate P5-4), and also indicates they are mostly quiescent under resting conditions. In contrast, SCs derived from the G93A hindlimb and diaphragm displayed heterogeneous levels of Vcam1, potentially due to ongoing activation/differentiation triggered by disease progression (Figure 3A). Importantly, EOM SCs from both WT and G93A mice were heterogeneous in Vcam1 levels, implying they are spontaneously activated even in the absence of pathological stimulation (Figure 3A). While significant decreases of P5-4 population were observed in the G93A hindlimb and diaphragm derived SCs compared to WT, no significant difference was detected between G93A and WT EOM derived SCs (Figure 3B).

FACS-isolated SCs were next cultured in growth medium for 4 days, fixed, and stained for MyoD to confirm their myogenic lineage. Close to 100% of the cells were MyoD positive (Figure 4-figure supplement 1A). We examined the ratio of actively proliferating cells (Ki67+) among the SCs (labeled with Pax7) [44]. SCs isolated from both WT and G93A EOMs exhibited the highest Pax/Ki67 double positive ratio, and the lowest Pax7/Ki67 double negative ratio (Figure 4A). These observations are consistent with previous reports of superior proliferative potential of EOM SCs compared to limb and diaphragm muscles [36]. The direct comparison between WT and G93A muscle SCs is presented in Figure 4-figure supplement 2.

Proliferation and differentiation properties of FACS-isolated SCs in culture.

(A) Representative images of FACS-isolated SCs cultured for 4 days in growth medium stained with antibodies against Pax7, Ki67 (proliferating cell marker), MyoD (myogenic lineage marker). Measurement results for the ratios of Pax7+Ki67+, Pax7+Ki67- and Pax7-Ki67- cells are shown in the right two panels. Scale bars, 100 μm. Each dot in the box-and-dot plots represents one image analyzed. For WT HL SCs, 94 images from 3 rounds of sorting; For WT Dia SCs, 107 images from 3 rounds of sorting; For WT EOM SCs, 75 images from 3 rounds of sorting; For G93A HL SCs, 96 images from 3 rounds of sorting; For G93A Dia SCs, 62 images from 3 rounds of sorting; For G93A EOM SCs, 79 images from 3 rounds of sorting. For the 3 rounds of sorting, 1 was from male and 2 were from female mice. * P < 0.05; ** P < 0.01; **** P < 0.0001; ns, not significant (t-test). ANOVA P values are also shown. (B) Representative images of FACS-isolated SCs cultured for 4 days in growth medium and 2 days in differentiation medium stained with antibodies against Pax7, Ki67, myosin heavy chain (MHC, differentiated myotube marker). Measurement results for the ratios of Pax7+Ki67+ and Pax7+Ki67- cells, as well as the fusion indices are shown in the right two panels. Each dot in the box-and-dot plots represents one image analyzed. For WT HL SCs, 52 images from 3 rounds of sorting; For WT Dia SCs, 51 images from 3 rounds of sorting; For WT EOM SCs, 51 images from 3 rounds of sorting; For G93A HL SCs, 52 images from 3 rounds of sorting; For G93A Dia SCs, 47 images from 3 rounds of sorting; For G93A EOM SCs, 56 images from 3 rounds of sorting. For the 3 rounds of sorting, 1 was from male and 2 were from female mice. ANOVA P values are also shown.

We also compared myotube formation and cell proliferation of cultured SCs under differentiation conditions. After 4 days in growth medium, we switched to differentiation medium and cultured the cells for 2 more days. The cells were then stained for sarcomere myosin heavy chain (MHC), Pax7 and Ki67. The EOM-derived SCs exhibited a higher number of Pax7 positive cells (Pax7+/Ki67+ and Pax7+/Ki67-) compared to SCs derived from hindlimb and diaphragm muscles – from both WT and G93A mice (Figure 4B). In contrast, the fusion indices, which calculate the ratio of nucleus within the MHC positive myotubes, were the lowest for EOM SCs (Figure 4B). This does not necessarily mean the EOM SCs have deficiencies in differentiation after they are committed, but more likely a reflection of their superior preservation of actively proliferating population [45]. In support of this view, previous TA engraftment experiments showed that more myofibers were generated by EOM SCs than their limb counterparts [36].

Butyrate induces EOM-like preservation of renewability in FACS-isolated SCs under differentiation conditions in vitro

Since NaBu-supplemented feeding reduced SC depletion in end stage G93A mice (Figure 2) in vivo, we tested NaBu treatment on cultured SCs. G93A hindlimb SCs were cultured with 0, 0.1, 0.5, 1.0, 2.5 and 5 mM of NaBu added to the growth medium for 1 day. RNAs were collected for qRT-PCR to evaluate the short-term impact of NaBu on cell growth. The relative expression of the proliferation markers (HMGA2 and CCND1) peaked at 0.5 mM. In contrast, the relative expression of proliferation inhibitor P21 bottomed at 0.5 mM (Figure 5A), indicating 0.5 mM as the optimal concentration of NaBu application in vitro.

Proliferation and differentiation properties of FACS-isolated SCs with different NaBu treatment conditions.

(A) qRT-PCR based relative quantification of Hmga2, Ccnd1 and P21 expression in FACS-isolated G93A HL SCs cultured with different doses of NaBu in growth medium for 1 day (n = 6, each dot in the box-and-dot plots represents one replicate of culture). HL SCs were from male mice. *** P < 0.001; **** P < 0.0001 (t-test). ANOVA P values are also shown. (B) Left two panels show the representative images of FACS-isolated G93A hindlimb and diaphragm SCs cultured in growth medium for 4 days and differentiation medium for 2 days and experienced different NaBu treatment conditions and stained with Pax7, Ki67 and MHC antibodies. Cells in top row were not treated with NaBu. Cells in the middle row were treated with NaBu for 1 day before the induction of differentiation (Day 4 of culture). Cells in the bottom row were treated with NaBu continuously for 3 days from the day before the induction of differentiation (Day 4-6 of culture). Scale bars, 100 μm. Measurement results for the ratios of Pax7+Ki67+ and Pax7+Ki67- cells, as well as the fusion indices are shown in the right two panels. Each dot in the box-and-dot plots represents one image analyzed. For G93A HL SCs, 52 images from 3 rounds of sorting; 1-day NaBu treatment, 45 images from 3 rounds of sorting; 3-day NaBu treatment, 51 images from 3 rounds of sorting; For G93A Dia SCs, 47 images from 3 rounds of sorting; 1-day NaBu treatment, 60 images from 3 rounds of sorting; 3-day NaBu treatment, 57 images from 3 rounds of sorting. For the 3 rounds of sorting, 2 were from male and 1 was from female mice. * P < 0.05; ns, not significant (t-test). ANOVA P values are also shown.

We designed two NaBu treatment plans for G93A diaphragm and hindlimb SCs as follows: one with 0.5 mM NaBu applied in the growth medium for 1 day before switching to differentiation medium (1 day in growth medium), the other with NaBu continuously applied for 3 days (1 day in growth medium + 2 days in differentiation medium). As shown in Figure 5B, the G93A hindlimb and diaphragm SCs in culture showed significant increase in the ratio of Pax7 positive cells under the 1-day NaBu treatment, associated with a decrease in fusion index. This trend was more pronounced under the 3-day NaBu treatment. These data indicate that NaBu treatment results in EOM-like preservation of renewability of the G93A SCs (Figure 4B). Aside from the drop of fusion indices after 3-day treatment with NaBu, there was notable decrease of MHC levels, indicating long-term exposure to NaBu did inhibit the differentiation process (Figure 5B), which is in line with the published result observed in myoblasts of different origins [46-48].

EOM-derived SCs show distinct transcriptome profile supporting self-renewal and axonal growth when compared to the SCs derived from other muscle origins

We extracted RNAs from SCs of hindlimb, diaphragm and EOMs under the following two culture conditions: first, in growth medium for 4 days (with suffix “G”); second, in growth medium for 4 days and then in differentiation medium for 2 days (with suffix “D”). The RNAs were send for bulk RNA-Seq. Additionally, RNA-Seq analyses were conducted for SCs of G93A hindlimb and diaphragm muscles cultured with NaBu for 3 days (with suffix “NaBu3”) (Figure 6).

Comparing the transcriptomic profile and homeostasis preferences of SCs from different muscle origins cultured in growth and differentiation medium.

(A) PCA analysis and hierarchical clustering results of different RNA-Seq samples. Samples with suffix “G” were SCs cultured in growth medium for 4 days. Suffix “D” denote SCs cultured in growth medium for 4 days and differentiation medium for 2 days. Suffix “NaBu3” denote SCs treated with 0.5 mM NaBu continuously for 3 days (Day 4-6 of culture). Dashed-line circles in PCA plot highlight samples close to each other in distance. All SC samples shown here were from female mice. (B) We identify genes differentially expressed in EOM SCs compared to their diaphragm and HL counterparts cultured in growth medium (leftmost panel) and differentiation medium (middle panel) by screening for genes shared by the four DEG lists (gold, magenta, pink and green in color) shown in the Venn diagrams. All DEG lists were generated based on the standard of |log2FC|>= 0.4 and FDR<= 0.001. Yellow box highlights the shared genes. To identify NaBu treatment signature genes, we screened for genes shared by the two DEG lists (green and gold in color) shown in the rightmost Venn diagram. (C) Dot plots in the upper row show the Log2FC of activation signature genes (magenta) and differentiation signature genes (cyan) identified in the 1119, 3318 and 3018 genes highlighted in yellow boxes in (B). Dot plots in the lower row show the Log2 FC of quiescence signature genes identified in the yellow-box highlighted genes in (B). Additionally, genes belonging to the following subgroups of quiescent signature genes: most commonly used SC markers, Notch signaling components and pluripotency signaling components are colored in dark blue, red and gold, respectively.

Similarities of the transcriptomic profiles of RNA-Seq samples were examined using the Principle Component Analysis (PCA) and hierarchical clustering (Figure 6A). We found that both WT and G93A EOM SCs cultured in differentiation medium were more similar to hindlimb and diaphragm SCs in growth medium than their counterparts in differentiation medium, highlighting the superior renewability of EOM SCs in differentiation medium and is consistent with the immunostaining results showed in Figure 4B.

Considering that cultured WT and G93A SCs from the same muscle origins exhibited relatively similar gross transcriptomic profiles in PCA and hierarchical clustering analysis, and that the proliferation and differentiation properties of cultured WT and G93A SCs from the same muscles were only marginally different in immunofluorescence assays (Figure 4-figure supplement 2), we speculate that the differential homeostasis seen in the FACS profiles of hindlimb and diaphragm SCs (Figure 3) may relate to the pathological milieu in vivo. When G93A SCs are isolated and cultured in vitro, their properties become similar to those of WT counterparts.

Additionally, we listed top 20 differentially expressed genes (DEGs) (ranked by Log2FC, both the upregulated and downregulated) by comparing: (1) EOM SCs to their hindlimb and diaphragm counterparts (Figure 6-table supplement 1); (2) G93A SCs to WT SCs of the same muscle origin (Figure 6-table supplement 2); (3) G93A hindlimb and diaphragm SCs with and without 3 day-NaBu treatment (Figure 6-table supplement 3).

The top 20 DEGs comparing EOM SCs to their hindlimb and diaphragm counterparts are mainly homeobox transcription factors (Figure 6-table supplement 1), reflecting different anatomic origin of these SCs [49]. For example, Alx4 and Pitx1 are expressed significantly higher in EOM SCs, while Lbx1, Tbx1, HoxA and HoxC genes are expressed significantly lower in EOM SCs than hindlimb and diaphragm counterparts acquired from either WT or G93A mice. These observations are consistent with previous reports [50,51]. Additionally, Pax3 is expressed significantly lower in EOM SCs than hindlimb and diaphragm counterparts cultured in growth medium, which reflects the unique Pax3-independent induction of myoD expression reported in EOM SCs [50]. It is also worth mentioning that BMP4 is expressed significantly higher in G93A EOM SCs compared to G93A hindlimb and diaphragm counterparts cultured in differentiation medium, potentially contributing to the continued expansion and slowed myogenic differentiation of EOM SCs [52]. This trend is also seen in SCs acquired from WT mice, but not pronounced enough to make it to the top 20 list. However, most of the top 20 differentially expressed genes between G93A and WT SCs (Figure 6-table supplement 2) are with poor functional annotation records and there exist great variations in the lists between muscles of different origins. We also examined the top 20 differentially expressed genes following 3 day-NaBu treatment in G93A hindlimb and diaphragm SCs. It did not turn out to be very informative (Figure 6-table supplement 3).

To better understand RNA-Seq data in the perspective of SC homeostasis, which refers to the transition between quiescence (dormant, slow self-renewal), activation (rapid amplification with symmetric and asymmetric cell division) and differentiation states during tissue regeneration [45], we compiled the lists of quiescence, activation and differentiation signature genes respectively. For quiescence signature genes, a list of 507 genes highly upregulated in quiescent SCs compared to activated SCs have been generated previously by Fukada et al [53]. We adopted this list together with other 18 genes denoted by other investigators, including Notch1, Notch2, Hes1, Hey1, Rbpj, Dtx4, Jag1, Sox8, Sdc3, Itga7, Itgb1, Met, Vcam1, Ncam1, Cdh2, Cdh15, Cxcr4, Cav1 [45,54-57]. The activation signature gene list was generated from our RNA-seq data using those expressed significantly higher in SCs (both from WT and G93A) in growth medium (1187 genes in total, |log2FC| ≥ 0.4 and FDR ≤ 0.001). Accordingly, the differentiation signature gene list was generated from those expressed significantly higher in SCs cultured in differentiation medium (1349 genes in total).

To verify that our signature gene lists recapitulate the SC properties at activation and differentiation state reported in previous studies, we did functional annotation analysis (KEGG pathway, Gene ontology_biological process, Gene ontology_cellular component). As shown in Figure 6-figure supplement 4, the top functional annotations enriched in our activation signature genes are all related to cell cycle and DNA replication, while the top functional annotations for our differentiation signature genes are associated with muscle contraction, muscle-specific subcellular structures and metabolic changes.

To explore whether the homeostatic states of EOM SCs are different from that of SCs from other muscle origins, we looked for differentially expressed genes (DEGs) shared-in-common among the following four groups: WT EOM SC-vs-WT diaphragm SC; WT EOM SC-vs-WT HL SC; G93A EOM SC-vs-G93A diaphragm SC; G93A EOM SC-vs-G93A HL SC (Figure 6B, left two panels). Within the 1119 shared DEGs identified in SCs cultured in growth medium, slightly more activation signature genes and quiescence signature genes were expressed higher in EOM SCs compared to diaphragm and hindlimb derived SCs. Meanwhile more differentiation signature genes were expressed lower in EOM SCs (Figure 6C, leftmost panel and Figure 6-table supplement 4). These trends became much more prominent in the 3318 shared DEGs identified in SCs cultured in differentiation medium, implying that EOM SCs are notably superior at preserving renewability in a differentiation environment than their diaphragm and hindlimb counterparts (Figure 6C, middle panel). Consistently, subgroups of quiescence signature genes belonging to commonly used stem cell markers (Pax7, Cd34, Cxcr4, Vcam1, Sdc3, Sdc4, Cdh15, Met, Cav1), Notch signaling pathway (Notch1, Notch2, Notch3, Jag1, Heyl, Dtx4, Msc) and pluripotency signaling pathway (Klf4, Igf1, Bmp2, Bmp4, Fzd4, Fgfr4, Pik3r1, Lifr) were all expressed higher in EOM SCs in differentiation medium than SCs derived from diaphragm and hindlimb muscles (Figure 6C, middle panel and Figure 6-table supplement 5).

In parallel, we examined whether the 3-day NaBu treatment could alter the homeostasis of G93A diaphragm and hindlimb SCs (Figure 6B, rightmost panel). There were 3018 DEGs shared between the following two comparing groups: G93A diaphragm SC_D_NaBu3-vs-G93A diaphragm SC_D and G93A HL SC_D_NaBu3-vs-G93A HL SC_D. Similar to the trends observed in EOM SCs in differentiation medium, the majority of activation and quiescence signature genes found in these DEGs were upregulated in the NaBu treated group, while the differentiation signature genes were downregulated (Figure 6C, rightmost panel). The elevated quiescence signature genes of the NaBu treated group include commonly used stem cell markers (Pax7, Cd34, Vcam1, Sdc4, Cdh2, Met, Cav1, Itga7), Notch signaling pathway members (Notch3, Rbpj) and pluripotency signaling pathway components (Fzd4, Bmp4). However, other three pluripotency related genes (Klf4, Wnt4 and Bmp2) were found downregulated (Figure 6-table supplement 6), indicating that NaBu treatment also has a mixed impact on the stemness of the SCs.

To compile EOM SCs signature gene list, we identified 621 genes expressed higher in EOM SCs than diaphragm and hindlimb SCs cultured in growth medium from the 1119 DEGs (Figure 6B, leftmost panel). Next, we pulled out 2424 genes that are expressed higher in EOM SCs than diaphragm and hindlimb SCs cultured in differentiation medium from the 3318 DEGs (Figure 6B, middle panel). These two lists have 478 genes shared-in-common. Particularly, many of these genes are related to axon guidance, cell migration and neuron projection (Figure 7A). Figure 7A-figure supplement 1, Figure 7-table supplement 1 and 2 further demonstrate this phenotype when the functional annotation analysis was done separately for the growth or differentiation culture conditions. The functional annotation analysis for the 3018 genes altered by the 3-day NaBu treatment (Figure 6B, rightmost panel) also spotted axon guidance, cell migration and neuron projection on the top of the annotation lists (Figure 7B and Figure7-table supplement 3). Importantly, the chemokine Cxcl12 shows up in all three lists. Cxcl12 is known to promote axonal extension and NMJ regeneration [58]. These data suggest that enhanced attraction of motor neuron axons by peri-NMJ SCs might be a shared mechanism underlying the NMJ preservation in EOMs and NaBu treatment observed in G93A mice.

Functional annotation analysis of EOM SC signature genes and NaBu treatment signature genes and qRT-PCR results.

(A) Top KEGG pathway, GO_P and GO_C annotations of the 478 EOM signature genes (expressed higher in EOM SCs compared to diaphragm and HL counterparts cultured in both growth medium and differentiation medium). Arrows highlight annotations related to axon guidance and cell migration. (B) Top KEGG pathway, GO_P and GO_C annotations of the 3018 NaBu treatment signature genes shown in Figure 6B, rightmost panel. (C) qRT-PCR based relative quantification (RQ, fold of change against WT HL SCs) of Hmga2 (activation signature gene), Actn3 (differentiation signature gene), Notch3 (quiescence signature gene) and Cxcl12 (axon guidance molecule) expression in HL, diaphragm and EOM SCs cultured in growth (upper row) and differentiation medium (lower row), respectively. RNA samples were collected from 3-6 rounds of sorting and sorted cells were seeded into 3 dishes as replicates. Each dot in the box-and-dot plots represents one replicate of culture. WT HL SCs were from 3 male and 1 female mice. WT diaphragm SCs were from 2 male and 2 female mice; WT EOM SCs were from 3 male mice; G93A HL SCs were from 4 male and 2 female mice. G93A diaphragm SCs were from 1 male and 3 female mice; G93A EOM SCs were from 3 male mice. * P < 0.05; ** P < 0.01; *** P < 0.001; **** P < 0.0001; ns, not significant (t-test). ANOVA P values are also shown. (D) qRT-PCR based relative quantification (fold of change against G93A HL SCs without NaBu treatment) of Hmga2, Actn3, Notch3 and Cxcl12 expression in G93A hindlimb and diaphragm SCs with or without the 3-day NaBu treatment. RNA samples were collected from 6 rounds of sorting and sorted cells were seeded into 3 dishes as replicates. G93A HL SCs were from 4 male and 2 female mice; G93A diaphragm SCs were from 2 male and 4 female mice. ANOVA P values are also shown.

We selected one representative gene from each of the following categories: activation (Hmga2), differentiation (Actn3) and quiescence (Notch3) signature genes, as well as an axon guidance gene (Cxcl12) for qRT-PCR. For each muscle type and culture condition, we collected samples from 3-6 rounds of sorting (the individual results are listed in Figure 7-table supplement 4-6). We observed significantly higher expression of Hmga2 in EOM SCs compared to diaphragm and hindlimb SCs cultured in either growth or differentiation medium (Figure 7C, leftmost panel), confirming the superior renewability of EOM SCs. The 3-day NaBu treatment of diaphragm and hindlimb SCs derived from G93A mice also upregulated Hmga2 (Figure 7D, leftmost panel). Actn3 was expressed lower in EOM SCs than diaphragm SCs cultured in either growth or differentiation medium, but not necessarily lower than hindlimb SCs (Figure 7C, 2nd panel from the left). Meanwhile NaBu treatment significantly decreased Actn3 expression (Figure 7D, 2nd panel from the left). Consistent with the RNA-Seq data, the quiescence marker Notch3 was expressed significantly higher in EOM SCs than diaphragm and hindlimb SCs cultured in differentiation medium but not in growth medium (Figure 7C, 2nd panel from the right). Similarly, NaBu treatment also elevated Notch3 expression (Figure 7D, 2nd panel from the right) in G93A diaphragm and hindlimb SCs cultured in differential medium. Importantly, the axon guidance molecule Cxcl12 was expressed significantly higher in EOM SCs than diaphragm and hindlimb SCs cultured in both growth and differentiation medium (Figure 7C, rightmost panel). NaBu treatment elevated Cxcl12 expression as well (Figure 7D, rightmost panel), confirming enhanced axon attraction as a common feature shared by EOM SCs and NaBu treatment.

AAV mediated overexpression of Cxcl12 in myotubes promotes motor neuron axon extension

Since recombinant Cxcl12 is known to promote axon extension [58], we tested whether overexpressing Cxcl12 in myotubes via adeno-associated virus (AAV) transduction could achieve the same effect. For these experiments, we co-cultured motor neurons and SC-derived myotubes on compartmentalized microfluidic chambers (XonaChip). Because AAV vectors do not transduce satellite cells efficiently [59,60], we first seeded the satellite cells into one compartment of the XonaChips, then induced them to differentiate for two days before AAV transduction. Four types of AAV-transduced myotubes were used for the coculture assays: a) WT hindlimb SC-derived myotubes transduced with AAV-CMV-eGFP; b) G93A hindlimb SC-derived myotubes transduced with AAV-CMV-eGFP; c) G93A hindlimb SC-derived myotubes transduced with AAV-CMV-Cxcl12-IRES-eGFP; and d) G93A EOM SC-derived myotubes transduced with AAV-CMV-eGFP. One day after AAV transduction, rat spinal motor neurons (RSMN) were seeded into the other compartment of the XonaChips and the culture medium switched to Neurobasal Plus Medium (NBPM) (Figure 8A). The cocultures were maintained for 9 more days. Immunostaining confirmed increased presence of Cxcl12 protein in the form of cytosolic vesicles in myotubes transduced with AAV8-CMV-Cxcl12-IRES-eGFP but not those transduced with AAV8-CMV-eGFP (Figure 8-figure supplement 1). These are likely vesicles for secretion [61].

Coculture of AAV transduced SC derived myotubes with rat spinal motor neurons in compartmentalized microfluidic chambers.

(A) Schematic representation of XonaChip spatial configuration and coculture experiment timeline. NBPM: Neurobasal Plus Medium; RSMNs: rat spinal motor neurons. (B) Representative composite images of RSMNs cocultured with: 1. WT HL SC-derived myotubes transduced with AAV-CMV-eGFP; 2. G93A HL SC-derived myotubes transduced with AAV-CMV-eGFP; 3. G93A HL SC-derived myotubes transduced with AAV-CMV-Cxcl12-IRES-eGFP; 4. G93A EOM SC-derived myotubes transduced with AAV-CMV-eGFP. Arrows highlight RSMN neurites crossed the microgrooves but stayed close to the edges and grew backward. Also see Figure 8-figure supplement 2A. Boxed regions are enlarged below. Scale bars: 50 μm. (C) Zoom-in views of RSMN neurites innervated or missed the AChR clusters (BTX positive patches). Boxed regions are further enlarged below. Scale bars: 50 μm, 10 μm. (D) Measured lengths of the longest neurites derived from RSMN cells (presumably axons) crossing the 450 μm microgrooves. n = 24, 31, 34,31 for the 4 groups, respectively (from 3 rounds of coculture experiments). Each dot in the box-and-dot plot represents one neurite measured. HL and EOM SCs used for co-culture experiments were all from male mice. ** P < 0.01; **** P < 0.0001 (t-test). ANOVA P value is also shown.

By measuring the lengths of the longest RSMN neurites (presumably axons) crossing the 450 μm microgrooves, we noticed that hindlimb-SC derived myotubes from G93A mice, but not from WT littermates, appeared repulsive to distal axons. The axons stayed close to the edge of the microgrooves and even grew backward towards the neuronal compartment, which we termed “edge effect” (Figure 8B and Figure 8-figure supplement 2A, arrows). This was rarely seen in the WT hindlimb SC-derived myotubes. In contrast, RSMN axon extension was dramatically enhanced in the co-culture system either with G93A hindlimb SC-derived myotubes overexpressing Cxcl12, or G93A EOM SC-derived myotubes overexpressing GFP (as control) (Figure 8B, C). Those neurites located close to the microgrooves were often seen to closely grow along nearby myotubes (Figure 8-figure supplement 2B). G93A hindlimb SC-derived myotubes overexpressing Cxcl12 occasionally were even seen to overcome the “edge effect” and redirect the axons towards their proximities (Figure 8-figure supplement 2B, arrow) - a persuasive indication of chemoattractive effects of those myotubes. Meanwhile, RSMN innervation of AChR clusters (BTX positive patches) were not seen in co-cultures with G93A hindlimb SC-derived myotubes overexpressing GFP, while they were consistently observed in the other three types of AAV-transduced myotubes (Figure 8C).

The overall innervation incidence was too low for statistical analysis due to the sparsity of axons in the myotube compartments, thus we conducted co-culture experiments where RSMNs were added to SC-derived myotubes in a uni-compartmental co-culture setting. As shown in Figure 9A-C and Figure 9-figure supplement 1A-C, when co-cultured in a single compartment, Cxcl12 overexpressing G93A hindlimb SC-derived myotubes exhibited increased innervation compared to GFP overexpressing controls. Furthermore, individual Cxcl12 overexpressing EOM SC derived myotubes were frequently innervated by multiple (3 or more) times (Figure 9A-C and Figure 9-figure supplement 1D), recapitulating the “multi-innervation” property of EOMs in vivo [26]. The extra “anchor points” seemed to encourage better spatial alignment of the RSMN axon network to the myotubes globally in the co-cultures. This spatial alignment of RSMN axons and myotubes was not observed in co-cultures of RSMNs and hindlimb SC-derived myotubes (without Cxcl12 overexpression), which also did not exhibit multi-innervation (Figure 9A-C and Figure 9-figure supplement 1A-C). Under higher magnification, we often observed that axons sending branches to adjacent BTX positive patches on EOM SC-derived myotubes (Figure 9B, arrows). Of note, there was higher expression of several Netrin and Semaphorin family members reported to promote axon-branching in EOM SCs compared to other muscle-derived SCs cultured in differentiation medium (Figure 7-table supplement 2) [62-67]. This milieu, comprised of abundant axon guidance molecules, in combination with the robust self-renewability of satellite cells, may contribute to the relative EOM NMJ resistance in ALS.

Coculture of AAV transduced SC derived myotubes with rat spinal motor neurons within the same compartment.

(A) Representative composite images of RSMNs cocultured with: 1. WT HL SC-derived myotubes transduced with AAV-CMV-eGFP; 2. G93A HL SC-derived myotubes transduced with AAV-CMV-eGFP; 3. G93A HL SC-derived myotubes transduced with AAV-CMV-Cxcl12-IRES-eGFP; 4. G93A EOM SC-derived myotubes transduced with AAV-CMV-eGFP. The timeline was the same as that in Figure 8 but RSMNs were seeded on top of myotubes. Boxed regions are enlarged in (B). Scale bars: 50 μm. (B) Zoom-in views of innervated and non-innervated AChR clusters on myotubes having contact with RSMN neurites. Arrows highlight branching phenomena observed in RSMN neurites innervating AChR clusters. Scale bars: 10 μm. (C) Quantification of innervation ratios of AChR clusters (area > 10 μm2) in myotubes in contact with RSMN neurites. n = 37, 48, 40, 52 z-stack images for the 4 groups, respectively (from 3 rounds of coculture experiments). Also see methods for detailed description. Each dot in the box-and-dot plot represents one image analyzed. HL and EOM SCs used for co-culture experiments were all from male mice. *** P < 0.001; **** P < 0.0001; ns, not significant (t-test). ANOVA P value is also shown. (D) Graphic summary of the potential mechanisms contributing to EOM NMJ integrity maintenance during ALS progression mimicked by NaBu treatment.

Discussion

It is known that the EOMs are complex muscles. Besides the developmental myosin isoforms, EOMs also express both adult fast and slow myosin contractile elements [68], suggesting that the sparing may not be solely linked to the fast or slow twitch nature of the muscle fibers, rather the changes in SCs may play a pivotal role in preserving the EOM function during the progression of ALS. We have demonstrated an association between the degeneration of NMJs and depletion of peri-NMJ SCs in hindlimb and diaphragm muscles in vivo in end-stage G93A mice. In contrast, EOM NMJs exhibited no signs of denervation or SCs depletion. The FACS profile of SCs isolated from EOM suggests spontaneous activation even without pathological stimulation, which is supported by previous in vivo BrdU labeling and staining results [69]. Furthermore, immunostaining, qRT-PCR and RNA-Seq analyses all imply that EOM SCs in vitro possess superior renewability better than hindlimb and diaphragm SCs, under differentiation-inducing culture conditions. Comparative functional annotation analyses of the transcriptomes highlight higher expression of axon guidance molecules in EOM SCs. Neuromuscular coculture experiments demonstrate that EOM SC-derived myotubes are more attractive to potential axons derived from rat spinal motor neurons than their hindlimb counterparts – which in fact appeared to repel the branching axons. Thus, the unique homeostasis regulation and axon attraction properties of EOM SCs may contribute to the resistance to denervation of EOM under ALS (Figure 9D).

Earlier report of SC quantification from muscle biopsy samples using electron microscopy detected no significant difference between ALS patients and non-ALS controls [70]. Other studies using primary cultured SCs from muscle biopsy samples implied compromised differentiation program in ALS SCs compared to controls [71,72]. As to proliferation, one study reported enhanced proliferation in ALS SCs [72], while the other cued for deficits in proliferation in ALS SCs [71]. The various findings may reflect variations in the disease stage of patients during sampling. For animal studies, Manzano et al measured the amount of Pax7+ SCs in dissociated EDL and soleus myofibers from G93A mice and WT controls and observed large variations across muscle-type and disease stage [73]. In another study carried out by the same group, primarily cultured EDL and soleus SCs from G93A mice generally showed lower proliferation potential than the WT counterparts [74].

In our whole-mount muscle imaging data and FACS profiles of isolated SCs, SC depletion, at least for the most quiescent population, is observed in hindlimb and diaphragm muscles from end-stage G93A mice. Additionally, the heterogeneity of Vcam1 levels observed in hindlimb and diaphragm SCs derived from end stage G93A mice compared to those from WT controls indicates occurrence of denervation-related pathological activation. Although this activation may transiently elevate the number of SCs, the continuous dominance of asymmetric over symmetric cell division, as well as the differentiation pressure to regenerate myofibers could gradually exhaust the quiescent stem cell pool, leading to the depletion phenotype seen in the end stage G93A mice. In contrast, the superior preservation of renewability of EOM SCs seen in differentiation medium is an effective measure against this catastrophe. However, the spontaneous activation and superior renewability of EOM SCs may not always be an advantage. As mentioned in the introduction section, EOMs are preferentially involved in some other neuromuscular disorders. In the case of mitochondrial myopathies, mitochondrial DNA defects accumulate over time, resulting in compromised oxidative phosphorylation represented by cytochrome c oxidase (COX) deficient myofibers in patients [7,8]. It is possible that the more frequent self-renewal and spontaneous activation of EOM SCs contribute to higher rate of mitochondrial DNA replication, leading to accelerated spreading of mitochondrial DNA defects, resulting in higher proportion of COX-deficient myofibers than other muscles.

In this study, overexpressing Cxcl12 in G93A hindlimb SC-derived myotubes promotes motor neuron axon extension under the co-cultured condition. Cxcl12 has been known as a critical axon guidance cue for motor neurons during development [75,76]. In adult animals, peri-synaptic Schwann cells have recently been reported to transiently produce Cxcl12 after acute motor axon terminal degeneration [58]. Intraperitoneal injection of antibodies neutralizing secreted Cxcl12 protein slows down the regeneration process. In contrast, the exposure of primary motor neurons to recombinant Cxcl12 protein stimulates neurite outgrowth and directional axonal extension in compartmentalized culture setup, which could be reversed by the treatment of the Cxcr4 specific inhibitor AMD3100. Published studies have demonstrated that local administration of Cxcl12 protein to neurotoxin injected hindlimb muscles of mice improved evoked junctional potentials of soleus [58]. Thus, the axon guidance role of Cxcl12 has been consolidated both in vitro and in vivo. Since myoblasts also express Cxcr4 and will migrate along the Cxcl12 protein gradient during muscle regeneration [77,78], the abundant peri-NMJ SCs in EOMs expressing Cxcl12 may achieve synergistic effects with peri-synaptic Schwann cells, improving the long-term integrity of NMJs. However, it is possible that motor neurons carrying ALS mutations may respond differently to Cxcl12 mediated axon guidance than WT motor neurons. This is a limitation of the current study, which will be investigated in future co-culture studies.

Epigenetic regulation of SC transcriptome by histone deacetylase (HDAC) inhibitors, such as butyrate [79], could modulate the myogenesis process both in vitro and in vivo [46,48,80,81]. Butyrate is a gut-bacteria fermentation product that also acts as an energy source to improve mitochondrial respiration [79,82]. Previous studies by our lab and others showed NaBu-supplemented feeding prolongs the life span of G93A mice and reverses CypD-related mitoflash phenotypes in myofibers derived from G93A mice, which implies a potential decrease of oxidative stress. Additionally, NaBu treatment also improved mitochondrial respiratory function of NSC34 motor neuron cell line overexpressing hSOD1G93A [37,83,84]. Yet it was not known whether the beneficiary effects of NaBu are linked to any phenotypic changes of SCs, and whether NaBu treatment induced epigenetic modulations similar to that of EOM SCs. The study here revealed its beneficiary role on the preservation of NMJ integrity and peri-NMJ SCs affected in hindlimb and diaphragm muscles by ALS progression. NaBu treatment of G93A hindlimb and diaphragm SCs in culture induced transcriptomic changes mimicking some features seen in EOM SCs, including the upregulation of activation and quiescence signature genes and the downregulation of differentiation signature genes, indicating NaBu improved the renewability of G93A hindlimb and diaphragm SCs. Additionally, some axon guidance molecules, particularly Cxcl12, were also induced by NaBu application (Figure 9D). Thus, NaBu-induced EOM SC-like transcriptomic pattern could be a crucial contributor to its therapeutic effect against ALS progression. NaBu treatment has been reported to directly promote neuronal survival and axonal transport by inhibiting HDAC6 [85-88]. Combining NaBu treatment with application on top of CXCL12 overexpression may indeed have additive therapeutic benefits to slow ALS progression.

Overall, our studies have defined several mechanisms which (1) may explain why EOMs are relatively spared in ALS, and (2) provides a rationale for targeting these differences as a therapeutic strategy. Furthermore, analysis of these signatures in patient muscle biopsies could be used to select subsets of ALS patients who may be most likely to benefit from therapies targeting this signaling network, and to provide possible "response biomarkers" in pre-clinical and clinical studies. As a cocktail of sodium phenylbutyrate (a derivative of NaBu) and Taurursodiol (an antioxidant) under the Commercial name of Relyvrio has been recently approved in US and Canada as a new ALS therapy [89], our study could help gain more insights into the molecular mechanisms related to this type of ALS therapy.

Materials and methods

Animals

All animal experiments were carried out in accordance with the recommendations in the Guide for the Care and Use of Laboratory Animals of the National Institutes of Health. The protocol on the usage of mice was approved by the Institutional Animal Care and Use Committee of the University of Texas at Arlington (A19.001, approval date: 09/20/2018). WT mice used in this study were of B6SJL background. The ALS transgenic mouse model (hSOD1G93A) with genetic background of B6SJL was originally generated by Drs. Deng and Siddique’s group at Northwestern University and deposited to the Jackson Lab as B6SJL-Tg (SOD1*G93A) [90]. For the butyrate feeding experiment, the animals were fed with 2% NaBu from the age of 3 months (at the ALS disease onset) and lasting for 4 weeks.

Primary culture of satellite cells isolated by fluorescence-activated cell sorting

The hindlimb muscles including tibialis anterior (TA), extensor digitorum longus (EDL), gastrocnemius, and quadriceps were dissected from one mouse for each round of sorting. For diaphragm, both the crural and costal domains were dissected from one mouse for each round of sorting. For EOMs, the four rectus muscles in the eye sockets were dissected from 5-6 mice of the same gender for each round of sorting [91]. The collected muscles were placed in 0 mM Ca2+ Ringer’s solution. Excessive connective tissue, tendons, fat, blood clots and vessels were removed as much as possible. After cleanup, the muscles were transferred into the digestion media (DMEM containing 5% FBS, 1% Pen/Strep and 10 mM HEPES) and minced into small blocks using scissors. Collagenase II (Worthington LS004176, ≥125 units/mg), dispase II (Sigma D46931G, ≥0.5 units/mg), hyaluronidase (Worthington LS002592, ≥300 USP/NF units/mg) and Dnase I (Worthington LS002139, ≥2,000 Kunitz units/mg) were added at the final concentration of 0.26%, 0.24%, 0.16% and 0.04% (weight/volume), respectively. The digestion system was placed in an orbital shaker running at 70 rpm at 37 °C for 45 min. The digested muscles were triturated 10 times with a 20-gauge needle attached to a 5 ml syringe. Afterwards the triturated mixture was pipetted onto a pre-wetted 100 μm strainer and centrifuge at 100 g for 2.5 min to precipitate the bulky myofiber debris. The supernatant was transferred onto a pre-wetted 40 um strainer and cells were collected by centrifuged at 1200 g for 6.5 min. The cells were resuspended in sorting media (PBS containing 0.3% BSA, 10 mM HEPES and 1 mM EGTA) and stained with antibodies or dyes at 4 °C for 45 min with shaking. The antibodies used include PE anti-mVcam1 (Biolegend 50604858, 600 ng/106 cells), APC anti-mCD31 (Biolegend 50162682, 400 ng/106 cells), APC anti-mCD45 (Biolegend 50162782, 400 ng/106 cells) APC anti-mSca-1 (Biolegend 50163216, 200 ng/106 cells). This antibody combination followed a previous published protocol [41]. Vcam1 is a SC marker, CD31 is a endothelial cell marker, CD45 is a hematopoietic cell marker and Sca-1 is a mesenchymal stem cell marker [41]. For sorting, our target population is VCAM+CD31- CD45-Sca-1-. When needed, the cells were stained with methyl green (2 µg/ml) to check mortality rate. As a positive control, half of the cells were transferred to a new Eppendorf tube and methanol (precooled at -20°C) was added dropwise with gentle mixing and left on ice for 5 min, the dead cells were wash twice before being mixed with the rest half of cells for methyl green staining. The sorting process was conducted using BD FACS Melody Cell Sorter using a 100 µm nozzle. The excitation laser and emission filter setup for methyl green is the same as APC. The flow rate was tuned between 1-7 to adjust the event rate to 2000-4000/sec. Gating conditions of FSC-A/SSC-A, SSC-H/SSC-W, FSC-H/FSC-W, Vcam1/CD31_CD45_Sca-1 and Vcam1/Methyl green plots were shown in Figure 3-figure supplement 1 and Figure 3-figure supplement 2. The sorted cells were collected in growth media (DMEM containing 20%FBS, 1% Pen/Strep, 1% chick embryonic extract and 10 mM HEPES). Afterwards the cells were seeded into laminin (Corning 354232) -coated plates or glass-bottom dishes containing growth media (without HEPES). Culture media was changed on the third day and the fourth day. To induce differentiation, cells were washed three times with PBS and cultured in differentiation media (DMEM containing 2% horse serum) for 2 more days.

Neuromuscular coculture assay

For neuromuscular coculture in compartmentalized chambers (XonaCHIPS XC450), XoanCHIPS were coated with poly-L-ornithine (0.5 mg/ml) for 1 hr at 37 °C followed by laminin (0.1 mg/ml) for 1 hr at 37 °C. Primary cultured satellite cells were seeded at 1×10^5 into the myotube compartment in growth medium for attachment overnight and then switched to differentiation medium and cultured for 2 more days. Afterwards AAV8-CMV-eGFP (Vector Biolabs, 1×10^11 GC/ml) or AAV8-CMV-Cxcl12-IRES-eGFP (Vector Biolabs, 1.5×10^11 GC/ml, the Cxcl12 gene is of mouse origin) transduction was conducted for one day. The usage of higher titer for the second virus is for matching GFP signal intensity, as IRES-dependent expression is significantly lower than those directly driven by CMV [92]. The chambers were washed with Neurobasal Plus Medium twice before seeding rat spinal motor neurons into the neuronal compartment (1×10^5). The coculture was kept in Neurobasal Plus Medium for 9 more days and half medium change was conducted every 2 days.

For neuromuscular coculture in the same glass-bottom chamber, 8-well chambered cover glasses (Cellvis C8-1.5H-N) were coated with poly-L-ornithine and laminin as described above. Primary cultured satellite cells were seeded at 1×10^5/chamber in growth medium for attachment overnight and then switched to differentiation medium and cultured for 2 more days. AAV transduction was conducted for one day as described above. The infected myotubes were washed with Neurobasal Plus Medium (Thermo Fisher A3582901, with B-27 and GlutaMax supplement) before the addition of rat spinal motor neurons (4×10^3/chamber, iXCells 10RA-033) to the same chamber. The coculture was kept in Neurobasal Plus Medium for 9 more days and half medium change was conducted every 2 days.

Immunofluorescence (IF) and imaging of whole mount muscle samples

EDL, soleus and EOM (the four rectus muscles) were dissected out and fixed in 4% paraformaldehyde (PFA) for 1 hr. Afterwards the samples were washed with PBS containing 1% glycine once and two more times with PBS. Further fixation and permeabilization was done with methanol at -20°C for 10 min. For diaphragm, the costal muscles and their surrounding rib cages were dissected and immersed in methanol directly for fixation (PFA fixation was skipped because it made the diaphragm harder to flatten and introduced more non-specific noises in staining). The samples were rehydrated and washed with PBS. The epimysium and tendons were removed as much as possible. EDL and soleus were further split into smaller bundles along the longitudinal axis. Then the samples were immersed in blocking buffer (PBS containing 2% BSA, 2% horse serum, 0.1% Tween-20, 0.1% Triton X-100 and 0.05% sodium azide) overnight at 4 °C. Next day the samples were incubated with primary antibodies for 1-2 days. After washing with PBS for three times, the samples were incubated with Alexa Fluor labelled secondary antibodies (Thermo Fisher 1:1000) and Alexa Fluor 488 conjugated α-Bungarotoxin (Thermo Fisher B13422, 1:1000) for 1-2 days. The samples were then washed with PBS, counterstained with DAPI and mounted in antifade mounting media (20 mM Tris, 0.5% N-propyl gallate, 80% glycerol) for imaging. Primary antibodies used: neurofilament (DSHB 2H3 concentrate, 1:250), synaptophysin (Fisher PA11043, 1:400), Pax7 (SCBT sc-81648, 1:100), Pax7 (Thermo Fisher PA1-117, 1:200), CXCL12 (R&D MAB350, 1:200).

Z-stack scans of glycerol-cleared whole muscles were obtained using 40x high working distance lens of Leica TCS SP8 confocal microscope. The z-stacks were compacted into 2D images by maximal intensity projection. To quantify SC number surrounding NMJs, circles (75 µm in diameter) were drawn around each NMJ in Image J. The DAPI channel was used as a mask to filter out nonspecific staining of Pax7 in the extracellular matrix.

Immunofluorescence and imaging of cultured myoblasts/myotubes and motor neurons

The cells were fixed in 4% PFA for 10 min and then washed with 1% glycine and PBS. Further fixation and permeabilization was done either with methanol at -20°C for 10 min or 0.25% TritonX-100 for 15 min. The samples were then washed with PBS and immersed in blocking buffer for 40 min. Primary antibody incubation was done at 4 °C overnight. Next day the cells were washed with PBS and incubated with Alexa Fluor labelled secondary antibodies (1:1000) for 2 hours at room temperature, counterstained with DAPI and mounted in antifade mounting media for imaging. Primary antibodies used: Alexa Fluor647 conjugated α-Bungarotoxin (Thermo Fisher B13422, 1:1000), Pax7 (SCBT sc-81648, 1:100), MyoD (SCBT sc-377460, 1:100), Ki-67 (CST 12202S, 1:300), MYH (DSHB MF20 concentrate, 1: 200), neurofilament (DSHB 2H3 concentrate, 1:250), eGFP (Novus NBP-22111AF488, 1:250).

For quantifying innervation ratio, AChR clusters (BTX patches) larger than 10 μm^2 were counted on myotubes that have contact with RSMN neurites. Those myotubes having 0 pixel overlap with RSMN neurites were not included in measurement. The ratio of AChR clusters having overlaps with the RSMN neurites after visual inspection of the whole z-stacks of each view were calculated. The innervation ratio of each view is one dot in the box-and-dot plot in Figure 9C. The measurements were done for 3 rounds of coculture assays.

RNA extraction and qRT-PCR

For cultured SCs, cell culture medium was removed before RNA extraction and Trizol reagent was added to the well before scratching cells off using pipette tips. The homogenate was transferred into tubes containing 5 PRIME Phase Lock Gel (Quantabio 2302830), and 1/10 volume of 1-bromo-3-chloroporpane (Alfa Aesar A11395) was added. The samples were shaken vigorously for 12 s and centrifuged for 15 min at 16,000× g. Afterwards, the upper phase was transferred to another Eppendorf tube and isopropanol (1/2 volume of the initial homogenate) was added for precipitation overnight at −20 °C. Next day, after 15 min centrifugation at 16000 g the precipitate was washed briefly with 75% ethanol. After another 5 min centrifugation, the ethanol was removed. The precipitate was briefly dried and resuspended in RNase free water. For whole muscle samples, the freshly dissected muscles were snap-frozen in liquid nitrogen, kept at -80 °C freezer overnight and then soaked in RNALater ICE (Thermo Fisher AM7030) for 2-3 days at -20 °C. Before RNA extraction, leftover tendons and neurites were removed and muscles were transferred to 1 ml Trizol. Homogenization was conducted in FastPrep-24 Classic bead beating grinder (MP Biomedicals 116004500). The tissue homogenate was transfer to centrifuge tubes containing phase lock gel (QuantaBio 2302830) and 1/10 volume of 1-bromo-3-chloropropane was added. The tubes were hand shaken for 12 seconds, left at bench top for 2 minutes and then centrifuged at 16000 g at 4 °C for 15 minutes. The upper phase was transferred to a new centrifuge tube and mixed with equal volume of ethanol. The following steps of RNA purification was conducted with Direct-zol RNA Miniprep Plus kit (Zymo Research R2070). RNA concentration was measured with Quantus Fluorometer (Promega E6150). RNA quality was assessed by RNA ScreenTape analysis at Genomics Core Facility at the University of North Texas Health Science Center. Only RNAs with RIN ≥ 9 and 28s/18s ≥ 1 were used for qRT-PCR or RNA-Seq. Reverse transcription was done using Promega GoScript Reverse Transcription system. The qPCR reactions used Maxima SYBR Green/ROX qPCR Master Mix (Thermo Fisher) and were carried out by the StepOnePlus Real-Time PCR system (Thermo Fisher). Relative quantification (RQ) of gene expression was generated by ΔΔCt method. The sequences of primers used for qPCR were listed in Table 1.

Bulk RNA-Seq and analysis

Non-stranded mRNA sequencing was carried out by BGI Americas Corporation. The average number of clean reads per sample was 45.16 million (ranging from 44.11 to 48.51 million). The average percentage of clean reads was 93.32% (ranging from 92.42% to 93.97%). The average percentage of aligned reads (to Genome Reference Consortium Mouse Build 38 patch release 6) was 96.01% (ranging from 95.40% to 96.67%) and the average percentage of uniquely aligned reads was 86.46% (ranging from 85.12% to 87.63%). To identify differentially expressed genes (DEGs) between different samples (using Poisson distribution method to compare uniquely mapped reads of the gene), only genes with |log2FC| ≥ 0.4 and FDR ≤ 0.001 were considered. To identify DEGs between different groups (containing more than one sample, using DESeq2 method), only genes with |log2FC| ≥ 0.4 and Qvalue ≤ 0.05 were considered. PCA analysis, hierarchical clustering, functional annotation analysis (KEGG pathway, GO_C, GO_P) were performed in Dr. TOM platform of BGI. Dot plots showing the distribution of activation, differentiation and quiescence signature genes were generated in ggplot2 in R. Fastq files and the excel file containing TPM values has been uploaded to GEO database with record number GSE249484.

Statistics

Box-and-dot plots were generated using the open-source ggplot2 data visualization package for the statistical programming language R. Two independent sample t-tests were carried out by adding the stat_compare_means function. For multi-group data, one-way ANOVA P-values were generated using the R function aov(). For each two-group comparison, we used two independent sample t-tests and used brackets to highlight the groups compared.

Supporting information

eLife Assessment

The manuscript by Jingsong Zhou and colleagues tries to uncover the reasons for the resistance of extraocular muscles (EOMs) to degenerative changes induced by amyotrophic lateral sclerosis (ALS). The findings of the study offer valuable information that EOMs are spared in ALS because they produce protective factors for the NMJ and, more specifically, factors secreted by EOM-derived satellite cells. While most of the experimental approaches are convincing, the use of sodium butyrate (NaBu) in this study needs further investigation, as NaBu might have a variety of biological effects. Overall, this work may help develop future therapeutic interventions for patients with ALS.

Reviewer #3 (Public Review):

Summary:

In their paper Li et al. investigate the transcriptome of satellite cells obtained from different muscle types including hindlimb, diaphragm and extraocular muscles (EOM) from wild type and G93A transgenic mice (end stage ALS) in order to identify potential factors involved in the maintenance of the neuromuscular junction. The underlying hypothesis being that since EOMs are largely spared from this debilitating disease, they may secrete NMJ-protective factors. The results of their transcriptome analysis identified several axon guidance molecules including the chemokine Cxcl12, which are particularly enriched in EOM-derived satellite cells. Transduction of hindlimb-derived satellite cells with AAV encoding Cxcl12 reverted hindlimb-derived myotubes from the G93A mice into myotubes sharing phenotypic characteristics similar to those of EOM-derived satellite cells. Additionally, the authors were able to demonstrate that EOM-derived satellite cell myotube cultures are capable of enhancing axon extensions and innervation in co-culture experiments.

Strengths:

The strength of the paper is that the authors successfully isolated and purified different populations of satellite cells, compared their transcriptomes, identified specific factors release by EOM-derived satellite cells, overexpressed one of these factors (the chemokine Cxcl12) by AAV-mediated transduction of hindlimb-derived satellite cells. The transduced cells were then able to support axon guidance and NMJ integrity. They also show that administration of Na butyrate to mice decreased NMJ denervation and satellite cell-depletion of hind limbs. Furthermore, addition of Na Butyrate to hindlimb derived satellite cell myotube cultures increased Cxcl12 expression. These are impressive results providing important insights for the development of therapeutic targets to slow the loss on neuromuscular function characterizing ALS.

Weaknesses:

Several important aspects have not been addressed by the authors, these include the following points which weaken the conclusions and interpretation of the results.

(a) Na Butyrate was shown to extend the survival of G93A mice by Zhang et al. Na butyrate has a variety of biological effects. For example, anti-inflammatory effects, inhibits mitochondrial oxidative stress, positively influences mitochondrial function, is a class I / II HDAC inhibitor etc. What is the mechanism underlying its beneficial effects both in the context of mouse muscle function in the ALS G93A mice and in the in vitro myotube assay? Cytokine quantification as well as histone acetylation/methylation can be assessed experimentally and this is an important point that has not been appropriately investigated.

(b) In the context of satellite cell characterization, on line 151-152 the authors state that soleus muscles were excluded from further studies since they have a higher content of slow twitch fibers and are more similar to diaphragm. This justification is not valid in the context of ALS as well as many other muscle disorders. Indeed, soleus and diaphragm muscles contain a high proportion of slow twitch fibers (up to 80% and 50% respectively) but soleus muscles are more spared than diaphragm muscles. What makes soleus muscles (and EOMs) more resistant to ALS NMJ injury? Satellite cells from soleus muscles need to be characterized in detail as well.

Furthermore, EOMs are complex muscles, containing many types of fibers and expressing different myosin heavy chain isoforms and muscle proteins. The fact that in mouse both the globular layer and orbital layers of EOMs express slow myosin heavy chain isoform as well as myosin heavy chain 2X, 2A and 2B (Zhou et al., 2010 IOVIS 51:6355-6363) also indicates that the sparing is not directly linked to the fast or slow twitch nature of the muscle fiber. This needs to be considered.

(c) In the context of myotube formation from cultured satellite cells on line 178-179 the authors stained the myotubes for myosin heavy chain. Because of the diversity of myosin heavy chain isoforms and different muscle origin of the satellite cells investigated, the isoform of myosin heavy chain expressed by the myotubes needs to be tested and described. It is not sufficient to state anti-MYH.

(d) The original RNAseq results have not been deposited and while it is true that the authors have analyzed the results and described them in Figures 6 and 7 and relative supplements, the original data needs to be shown both as an xls list as a Volcano plots (q value versus log2 fold change). This will facilitate the independent interpretation of the results by the readers as some transcripts may not be listed. As presented it is rather difficult to identify which transcripts aside Cxcl12 are commonly upregulated. Can the data be presented in a more visual way?

(e) There is no section describing the statistical analysis methods used. In many figures more than 2 groups are compared so the authors need to use an ANOVA followed by a post hoc test.

The authors have achieved their aim in showing that satellite cells derived from EOMs have a distinct transcriptome and that this may be the basis of their sparing in ALS. Furthermore, this work may help develop future therapeutic interventions for patients with ALS.

Reviewer #4 (Public Review):

Summary:

In this work, the authors have used a mouse model of familial Amyotrophic lateral sclerosis (ALS) that carries a G93A mutation in the Sod1 gen to understand how the extraocular muscles (EOM) are preserved in ALS while other muscles undergo degeneration. Interestingly, the authors demonstrate that the integrity of neuromuscular junctions (NMJ) is affected by ALS in the limb and diaphragm muscles of G93A mice, while EOM is mostly preserved. The authors also further demonstrate that NaBu treatment partially restores the integrity of NMJ in the limb and diaphragm muscles of G93A mice. The results also indicate that chemokine Cxcl12 is expressed at higher levels in EOM myoblasts, and transduction with AAV encoding Cxcl12 improved the phenotypic characteristics of hindlimb-derived satellite cells.

Strengths:

The authors have used both in vivo and cell culture models. The findings have a translational potential.

Weaknesses:

The use of NaBu could be an issue as it has multiple effects and targets in ALS.

The sample size of animal experiments still needs to be improved.

The molecular mechanism of how Cxcl12 improved the phenotypic characteristics of hindlimb-derived satellite cells is still being determined.

Author response:

The following is the authors’ response to the previous reviews.

eLife assessment

The manuscript by Jingsong Zhou and colleagues tries to uncover the reasons for the resistance of extraocular muscles (EOMs) to degenerative changes induced by amyotrophic lateral sclerosis (ALS). The findings of the study offer valuable information that EOMs are spared in ALS because they produce protective factors for the NMJ and, more specifically, factors secreted by EOM-derived satellite cells. While most of the experimental approaches are convincing, the use of sodium butyrate (NaBu) in this study needs further investigation, as NaBu might have a variety of biological effects. Overall, this work may help develop future therapeutic interventions for patients with ALS.

We agree with the editor that NaBu have a variety of biological effects that require further investigation. Our team previously have explored the effect of NaBu treatment on intestinal microbiota and intestinal epithelial permeability (DOI: 10.1016/j.clinthera.2016.12.014), on the mitochondrial respiratory function of NSC-34 motor neuron cell line overexpressing hSOD1G93A (DOI: 10.3390/biom12020333) and on the mitochondrial function of skeletal muscle myofibers of G93A mice (DOI: 10.3390/ijms22147412). Other research teams have also explored the role of NaBu (or HDAC inhibition) in neuronal survival and axonal transport (DOIs: 10.1073/pnas.0907935106; 10.1038/s41467-017-00911-y; 10.15252/embj.2020106177; 10.1093/hmg/ddt028).

Since the theme of this manuscript is the transcriptomic characteristics of EOM SCs, to include data of how NaBu affect cellular/molecular processes of other tissues will somewhat deviate from the theme. It would be more appropriate to develop a separate manuscript focusing on other tissues.

We appreciate the feedback from the Editors and reviewers. We realized that our previous description on butyrate’s beneficial role might be overstated in the Abstract Section. We have made two changes to avoid potential overstatement of our finding: (1) We modified the Abstract to state that “the NaBu-induced transcriptomic changes resembling the patterns of EOM SCs “may contribute to” (instead of “underlie”) the beneficial effects observed in G93A mice” (Page 1, Line 29); (2) We have edited the corresponding paragraph in the Discussion section to emphasize that the effect of NaBu treatment is multi-faceted (Page 11, Line 459-461).

Recommendations for the authors:

Reviewer #3 (Recommendations For The Authors):

line 388-389. The sentence has been corrected but is still not clear. What do the authors mean by ".....resulting in higher proportion of COX-deficient myofibers than other muscles». What other muscles do they refer to?

Other muscles refer to muscles whose stem cells remain dormant under physiological conditions (uninjured, innervated), such as EDL. We have edited the sentence accordingly. (Page 10, Line 431-432)

In reference to the results shown in Fig. 2, 7, 8 and 9. Since the experimenters were not blinded, this should be explicitly stated in the Methods section.

We have added the disclaimer in the current “Data analysis and statistics” section in Methods as follows: “The experimenters were not blinded to the samples in data collection and analysis.” (Page 15, Line 636)

Figure 7 C has been amended but now the inserted ANOVA values interfere with the correct visualization of Fig. 7D, can panels D be moved down so that they are better separated from panels in Fig. 7C

Thanks for the comment and we have edited Figure 7 accordingly.

Reviewer #4 (Recommendations For The Authors):

The authors have revised the manuscript per the reviewer's comments in this study. While most of the concerns were addressed, a few concerns remain.

The molecular basis of how AAV-mediated delivery of Cxcl12 improves the phenotype of satellite cells is still unclear.

Thanks for the comment. As one of the earliest discovered chemokines, the chemotactic role of Cxcl12-Cxcr4 axis on cells and cellular processes (such as axons) has been comprehensively investigated by different functional assays from overexpression to protein application to inhibitor application to knockdown by shRNAs in different types of tissues. To list a few examples, the establishment of the correct routing trajectories of mammalian motor axons and oculomotor axons during embryonic development (DOIs: 10.1016/j.neuron.2005.08.011; 10.1167/iovs.18-25190). The regeneration of injured motor axon terminals guided by terminal Schwann cells in adult mice (DOI: 10.15252/emmm.201607257). The migration of neural crest cells to sympathetic ganglia in the formation of sympathetic nerve system during embryogenesis (DOI: 10.1523/JNEUROSCI.0892-10.2010). The migration of myoblasts in the process of fusion into myotubes (DOIs: 10.1242/jcs.066241; 10.1111/boc.201200022; 10.1074/jbc.M706730200).

Because the existence of so many detailed mechanistic studies, our goal for this manuscript is not to identify a novel mechanism of how Cxcl12-mediated chemotaxis is achieved. Rather, we used it as one of the proof-of-concept mechanisms contributing to the resistance of EOMs against ALS and benefits of NaBu treatment. Certainly, it is not the sole mechanism.

To address the reviewer’s concern, we have expanded discussion about the previous studies regarding the chemotactic effect of Cxcl12 in the discussion section. (Page 10, Line 435-436, Page 11, Line 445-446)

The NaBu experiments may need additional support from other approaches. NaBu effects may not be directly related to satellite cells or muscle cells. Thus, the animal experiment results need to be carefully interpreted.

We agree that NaBu have a variety of biological effects that require further investigation. Our team previously have explored the effect of NaBu treatment on intestinal microbiota and intestinal epithelial permeability (DOI: 10.1016/j.clinthera.2016.12.014), on the mitochondrial respiratory function of NSC-34 motor neuron cell line overexpressing hSOD1G93A (DOI: 10.3390/biom12020333) and on the mitochondrial function of skeletal muscle myofibers of G93A mice (DOI: 10.3390/ijms22147412). Other research teams have also explored the role of NaBu (or HDAC inhibition) in neuronal survival and axonal transport (DOIs: 10.1073/pnas.0907935106; 10.1038/s41467-017-00911-y; 10.15252/embj.2020106177; 10.1093/hmg/ddt028).

Since the theme of this manuscript is the transcriptomic characteristics of EOM SCs, to include data of how NaBu affect cellular/molecular processes of other tissues will somewhat deviate from the theme. It would be more appropriate to develop a separate manuscript specifically addressing the impact of NaBu on other tissues.

We appreciate the feedback from the reviewers. We realized that our previous description on butyrate’s beneficial role might be overstated in the Abstract Section. In response, we have made two changes to avoid potential overstatement of our finding: (1) We modified the Abstract to state that “the NaBu-induced transcriptomic changes resembling the patterns of EOM SCs “may contribute to” (instead of “underlie”) the beneficial effects observed in G93A mice” (Page 1, Line 29); (2) We edited the corresponding paragraph in the Discussion section to emphasize that the effect of NaBu treatment is multi-faceted (Page 11, Line 459-461).

Version 3

Abstract

Abstract

Amyotrophic lateral sclerosis (ALS) is a fatal neuromuscular disorder characterized by progressive weakness of almost all skeletal muscles, whereas extraocular muscles (EOMs) are comparatively spared. While hindlimb and diaphragm muscles of end-stage SOD1G93A (G93A) mice (a familial ALS mouse model) exhibit severe denervation and depletion of Pax7+ satellite cells (SCs), we found that the pool of SCs and the integrity of neuromuscular junctions (NMJs) are maintained in EOMs. In cell sorting profiles, SCs derived from hindlimb and diaphragm muscles of G93A mice exhibit denervation-related activation, whereas SCs from EOMs of G93A mice display spontaneous (non-denervation-related) activation, similar to SCs from wild-type mice. Specifically, cultured EOM SCs contain more abundant transcripts of axon guidance molecules, including Cxcl12, along with more sustainable renewability than the diaphragm and hindlimb counterparts under differentiation pressure. In neuromuscular co-culture assays, AAV-delivery of Cxcl12 to G93A-hindlimb SC-derived myotubes enhances motor neuron axon extension and innervation, recapitulating the innervation capacity of EOM SC-derived myotubes. G93A mice fed with sodium butyrate (NaBu) supplementation exhibited less NMJ loss in hindlimb and diaphragm muscles. Additionally, SCs derived from G93A hindlimb and diaphragm muscles displayed elevated expression of Cxcl12 and improved renewability following NaBu treatment in vitro. Thus, the NaBu-induced transcriptomic changes resembling the patterns of EOM SCs may contribute to the beneficial effects observed in G93A mice. More broadly, the distinct transcriptomic profile of EOM SCs may offer novel therapeutic targets to slow progressive neuromuscular functional decay in ALS and provide possible "response biomarkers" in pre-clinical and clinical studies.

Introduction

Amyotrophic lateral sclerosis (ALS) is a fatal disorder characterized by progressive motor neuron (MN) loss and skeletal muscle wasting, resulting in respiratory failure and death 3–5 years following diagnosis [1,2]. While not entirely spared, the extraocular muscles (EOMs) of ALS patients and mutant mouse models exhibit comparatively preserved structure and function and retained neuromuscular junctions (NMJ) [3–6]. The reason for this differential involvement remains unclear, especially considering that eye movements are preferentially involved in other neuromuscular disorders like myasthenia gravis, some muscular dystrophies, and mitochondrial myopathies [7,8].

While ALS is classically considered predominantly a “dying-forward” process of motor neurons (MN), accumulated evidence [9–17] support that early muscle dysfunction is an important contributor to ALS pathophysiology. The degree to which “dying-forward” from the central nerve system or “dying-back” process [18–20] starting distally at NMJs contributing to ALS progression remains unsettled and may vary in different patient subsets. Regardless of the direction of communication and relative contribution of different cell types to ALS progression (e.g. glia, neurons, myofibers) [20], it is widely accepted that NMJ disassembly and the resulting skeletal muscle denervation is a critical pathogenic event in both patients and animal models [21–24].

EOMs differ from limb and trunk muscles in many respects [25]. In contrast to other skeletal muscles, the individual motor neurons (MNs) controlling EOMs innervate fewer EOM myofibers, and individual myofibers often have multiple NMJs [26]. The multiple innervation phenotype refers to the co-existence of innervated "en plaque" and "en grappe" NMJs in a single EOM myofiber. The "en plaque" type of NMJs are located in the middle portion of myofibers and resemble NMJs seen in other types of muscles. Each EOM myofiber has one "un plaque" NMJ. "en grappe" NMJs are a series of smaller, grape-shaped NMJs that are located in the distal portion of some EOM myofibers [3,26]. MNs innervating EOMs express higher level of the Ca2+ binding proteins calbindin-D28k and parvalbumin, thus may be more resistant to the Ca2+-induced excitotoxicity [27–29]. Pronounced differences exist between EOM and limb muscles in their transcriptomic profiles [30]. In addition, a recent proteomic study revealed more than 2500 proteins differentially expressed between EOM and EDL muscles and more than 2000 between EOM and soleus muscles [31]. However, none of muscle-derived neuroprotective factors, including BDNF, GDNF, NT3 and NT4 were expressed (RNA or proteins) significantly higher in EOM myofibers compared to hindlimb myofibers in both mutant G93A and wild-type mice [31–33]. Thus, it is important to look for clues in other cell types that influence the microenvironment around NMJs, such as satellite cells (SCs).

An association between SC function, NMJ integrity, and resultant denervation has been demonstrated in animal models of aging and muscular dystrophy [34,35]. Using transgenic mice expressing GFP in the progeny of activated SCs, Liu and colleagues demonstrated that inducible depletion of adult skeletal muscle SCs impaired the regeneration of NMJs [34]. Furthermore, they found that the loss of SCs exacerbated age-related NMJ degeneration. Preservation of SCs attenuated aging-related NMJ loss and improved muscle performance in mice [35]. SCs from EOMs exhibit superior expansion and renewability (the capability to maintain dividing and undifferentiated stem cells after rounds of regeneration) than their limb counterparts [36], inspiring us to compare the phenotypes of SCs derived from EOMs and limb muscles and to conduct transcriptomic profiling of SCs to look for candidate genes that could preserve NMJ integrity in ALS.

In the current study, we found that the integrity of NMJs and the amount of Pax7+ SCs are maintained in EOMs in end-stage G93A mice, in contrast to the severe denervation and depletion of Pax7+ SCs in hindlimb and diaphragm muscles. Transcriptomic profiling revealed that the EOM SCs exhibit abundant transcripts of axon guidance molecules, such as Cxcl12, along with sustainable renewability compared with SCs from diaphragm and hindlimb muscles. Furthermore, EOM SC-derived myotubes enhance axon extension and innervation in co-culture experiments with spinal motor neurons compared to hindlimb SC-derived myotubes. Overexpressing Cxcl12 using adeno-associated viral vector (AAV8) in G93A hindlimb SC-derived myotubes recapitulated features of the EOM SC-derived myotubes.

Butyrate, a short-chain fatty acid produced by bacterial fermentation of dietary fibers in the colon, has been shown to extend the survival of ALS G93A mice [37]. We found that G93A mice fed with 2% sodium butyrate (NaBu) in drinking water decreased NMJ denervation and SC depletion in hindlimb and diaphragm muscles. Adding NaBu to the culture medium improved the renewability of SCs derived from the G93A hindlimb and diaphragm muscles, as well as increasing the expression of Cxcl12, resembling EOM SC transcriptomic pattern. Characterizing the distinct EOM SC transcriptomic pattern could provide clues for identifying potential biomarkers in therapeutic trials in both ALS patients and animal models, in addition to identifying therapeutic targets.

Results

NMJ integrity and peri-NMJ SC abundance are preserved in EOMs, and NaBu-supplemented feeding reduces NMJ denervation and SC depletion in hindlimb and diaphragm muscles in G93A mice

We investigated the NMJ integrity and peri-NMJ SC abundance by whole-mount imaging of extensor digitorum longus (EDL), soleus, diaphragm, and EOMs derived from WT littermates and G93A mice with or without NaBu-supplemented feeding. Acetylcholine receptors (AChRs) at NMJs were labeled with Alexa Fluor conjugated α-Bungarotoxin (BTX). Axons and axon terminals of motor neurons were labeled with antibodies to neurofilament (NF) and synaptophysin (SYP) respectively (Figure 1A and Figure 1-figure supplement 1A). In order to quantify the extent of denervation in a categorical manner, NMJs were arbitrarily defined as “well innervated”, “partially innervated”, and “poorly innervated” (Figure 1-figure supplement 1A). Poorly innervated NMJs also generally lacked nearby NF+ axons. For EOMs, only the “en plaque” type of NMJs were included in this comparison, because they resemble NMJs in other types of muscles in morphology and spatial distribution. The “en grappe” type of NMJs were not included.

Characterizing NMJ integrity in muscles dissected from wild-type controls and end-stage G93A mice with or without NaBu feeding.

(A) Representative compacted z-stack scan images of whole-mount EDL, soleus diaphragm extraocular muscles stained with antibodies against neurofilament medium chain (NF, labeling axons), synaptophysin (SYP, labeling axon terminals) and Alexa Fluor 488 conjugated α-Bungarotoxin (BTX, labeling AChRs on muscle membrane). scale bars, 50 μm. (B) Mean ratios of well innervated NMJs (SYP signals are present in >60% of BTX positive area), partially innervated NMJs (SYP signals are present in 30-60% of BTX positive area) and poorly innervated NMJs (SYP signals are present in <30% of BTX positive area) and in different types of muscles dissected from WT controls and end-stage G93A mice with or without NaBu feeding (see Figure 1-figure supplement 1, Figure 1-figure supplement 2 and Figure 1-Source Data 1 for NMJs measured per muscle type per gender. Briefly, EDL, soleus and diaphragm muscles were from 4 male and 6 female mice per group; WT EOM group was from 4 male and 4 female mice; G93A EOM group was from 3 male and 4 female mice; G93A EOM with NaBu feeding group was from 6 female mice). Each dot in the box-and-dot plots represents quantification result from a single mouse. * P < 0.05; ** P < 0.01; **** P < 0.0001; ns, not significant (t-test). ANOVA P values are also shown.

The relative proportions of well-, partially- and poorly-innervated NMJs in different muscle origins of WT and G93A mice are shown in Figure 1B (including both genders). At least 343 (average 758 ± 345 per group) NMJs from muscles (EDL, soleus, diaphragm and EOMs, respectively) dissected from WT or G93A mice with or without NaBu-feeding were examined. The original mouse information (including gender specification) and quantification data are listed in Figure 1-Source Data 1. EDL, soleus and diaphragm muscles, but not EOMs, from end stage G93A mice (4-month-old) exhibited significant decrease in well-innervated NMJs, accompanied by an increase in poorly-innervate NMJs. The increase of “poorly-innervated NMJ” is slightly less dramatic in diaphragm compared to the hindlimb muscles, meanwhile the increase of “partially-innervated NMJ” ratio is the most significant in the diaphragm (Figure 1B), indicating that denervation process is most severe in hindlimb muscles, slightly milder in diaphragm and barely detectable in EOMs. The data showed in Figure 1B have also been replotted to compare the innervation status between male and female mice (Figure 1-figure supplement 2). In the term of well- or partially-innervated ratios, there are no significant cross-gender difference observed in our experimental condition, in which the muscle samples were collected at the end stage of the disease, although there is marginally lower “poorly innervated ratio” in the EDL muscle of G93A female mice compared to G93A male mice.

In the previous study, the NaBu treatment of G93A mice started at the pre-symptomatic age (2∼2.5-month-old) significantly delayed ALS progression in G93A mice [37]. As the treatment for ALS patients is initiated after symptoms appear, we further tested whether NaBu treatment started after the disease onset (at the age of 3 months, 2% NaBu in water for 1 month) was effective in preserving NMJ integrity. Consistent with previous observation, the hindlimb and diaphragm muscles from NaBu-fed mice (1 month, 2% in water, starting to feed at the age of 3-month-old) all exhibited significantly reduced denervation compared to littermates who were not fed NaBu (Figure 1B, n = 10 per group, gender matched).

qRT-PCR for Scn5a, the gene encoding sodium channel Nav1.5, was performed with RNA extracted from whole muscles and used as another marker of denervation [38–40] (Figure 1-figure supplement 1B). The original values of Scn5a expression levels are provided in Figure 1-Source Data 2. Consistent with the immunostaining quantification in Figure 1A, the elevation of Scn5a expression is most notable when comparing G93A hindlimb muscles to wildtype counterparts (>250 folds for EDL, >70 folds for soleus), milder for diaphragm (> 15 folds) and not significant for EOMs. Additionally, NaBu feeding significantly reduced the Scn5a transcripts in hindlimb muscles (Figure 1-figure supplement 1B).

Whole mount imaging of muscles stained with the satellite cell (SC) markers Pax7, BTX, SYP, and DAPI, was used to quantify SCs in the vicinity of NMJs. The Pax7 signal in each z-stack image was filtered using DAPI as a mask to identify Pax7+ cell nuclei. Afterwards the z-stacks were compacted into 2D images by maximal intensity projection (Figure 2A). The number of nuclear Pax7+ cells were counted within the circles of 75 µm diameter drawn around the NMJs (Figure 2B, see also Figure 2-Source Data 1). Similar to the NMJ denervation status, depletion of peri-NMJ SCs was most significant in hindlimb muscles of end-stage G93A mice and slightly milder in diaphragm, while no significant depletion was observed in EOMs (Figure 2C). G93A mice fed with NaBu had less depletion of SCs in hindlimb and diaphragm muscles.

Quantification of peri-NMJ SC abundance in muscles dissected from wild-type controls and end-stage G93A mice with or without NaBu feeding.

(A) Representative compacted z-stack scan images of whole-mount EDL, soleus diaphragm extraocular muscles stained with antibodies against Pax7 (labeling SCs), SYP and BTX-Alexa Fluor 488. Scale bars, 50 μm. (B) To measure the abundance of SCs around NMJs, circles of 75 μm diameter were drawn around the NMJs and the number of nuclear Pax7+ cells (co-localized with DAPI) were counted. Scale bars, 10 μm. (C) Mean number of SCs around NMJs in different types of muscles dissected from WT controls and end-stage G93A mice with or without NaBu feeding (see Figure 2-Source Data 1 for NMJs measured per muscle type per gender. Briefly, WT EDL group was from 2 male and 2 female mice; G93A EDL group was from 3 male and 3 female mice; G93A EDL with NaBu feeding group was from 2 male and 4 female mice; WT soleus group was from 2 male and 3 female mice; G93A soleus group was from 3 male and 2 female mice; G93A soleus with NaBu feeding group was from 1 male and 4 female mice; WT diaphragm group was from 1 male and 4 female mice; G93A diaphragm group was from 1 male and 4 female mice; G93A diaphragm with NaBu feeding group was from 4 female mice; WT EOM group was from 1 male and 3 female mice; G93A EOM group was from 5 female mice; G93A EOM with NaBu feeding group was from 1 male and 3 female mice). Each dot in the box-and-dot plots represents quantification result from a single mouse. * P < 0.05; ** P < 0.01; **** P < 0.0001; ns, not significant (t-test). ANOVA P values are also shown.

The proliferation and differentiation properties are conserved in SCs derived from G93A EOMs

We conducted FACS based isolation of SCs from hindlimb (combining tibialis anterior, EDL, quadriceps and gastrocnemius), diaphragm, and EOMs using a modified version of a published protocol [41]. The three groups were determined because they represent the most severely affected, moderately affected and least affected muscles by ALS progression, respectively. Soleus was not included in the hindlimb SCs pool because its less affected than other hindlimb muscles based on our study and others [6,42]. To separate different cell populations, Vcam1 was used as a marker for quiescent SCs, while CD31 (endothelial cell marker), CD45 (hematopoietic cell marker) and Sca-1 (mesenchymal stem cell marker) positive cells were discarded. Thus, in the FACS profile the target SC population (Vcam1+CD31-CD45-Sca-1-) was in gate P5-4 (Figure 3A). Single antibody staining controls are shown in Figure 3-figure supplement 1. We also applied methyl green (2 µg/ml) staining [43] to evaluate the viability of cells (Figure 3-figure supplement 2), only about 1% of Vcam1 positive population were methyl green stained (Figure 3-figure supplement 2D, Events P5-2 / (P5-2 + P5-4)*100%). Thus, most cells were alive when loaded into the flow cytometer.

Isolation of SCs from wildtype and G93A muscles using fluorescence activated cell sorting.

(A) Representative FACS profiles during the isolation of Vcam1+CD31-CD45-Sca1- cells (gate P5-4). The corresponding population hierarchies are shown right to the plots. For single antibody staining controls, unstained controls and viability test, see Figure 3-figure supplement 1 and Figure 3-figure supplement 2. (B) Comparing the percentage of events in the P5-4 gate between cells isolated from WT and G93A mice. Each dot in the box-and-dot plots represents result from one round of sorting. WT HL SCs were from 8 male and 6 female mice; G93A HL SCs were from 9 male and 5 female mice; WT diaphragm SCs were from 6 male and 3 female mice; G93A diaphragm SCs were from 12 male and 5 female mice. WT EOM SCs were from 6 batches of male and 1 batch of female mice (each batch contains 5-6 mice of the same gender). G93A EOM SCs were from 5 batches of male and 2 batches of female mice. * P < 0.05; *** P < 0.001; ns: not significant (t-test).

SCs isolated from WT hindlimb and diaphragm muscles exhibited relatively homogenous levels of Vcam1, which facilitated separation from other cells and debris (Figure 3A, gate P5-4), and also indicates they are mostly quiescent under resting conditions. In contrast, SCs derived from the G93A hindlimb and diaphragm displayed heterogeneous levels of Vcam1, potentially due to ongoing activation/differentiation triggered by disease progression (Figure 3A). Importantly, EOM SCs from both WT and G93A mice were heterogeneous in Vcam1 levels, implying they are spontaneously activated even in the absence of pathological stimulation (Figure 3A). While significant decreases of P5-4 population were observed in the G93A hindlimb and diaphragm derived SCs compared to WT, no significant difference was detected between G93A and WT EOM derived SCs (Figure 3B).

FACS-isolated SCs were next cultured in growth medium for 4 days, fixed, and stained for MyoD to confirm their myogenic lineage. Close to 100% of the cells were MyoD positive (Figure 4-figure supplement 1A). We examined the ratio of actively proliferating cells (Ki67+) among the SCs (labeled with Pax7) [44]. SCs isolated from both WT and G93A EOMs exhibited the highest Pax/Ki67 double positive ratio, and the lowest Pax7/Ki67 double negative ratio (Figure 4A). These observations are consistent with previous reports of superior proliferative potential of EOM SCs compared to limb and diaphragm muscles [36]. The direct comparison between WT and G93A muscle SCs is presented in Figure 4-figure supplement 2.

Proliferation and differentiation properties of FACS-isolated SCs in culture.

(A) Representative images of FACS-isolated SCs cultured for 4 days in growth medium stained with antibodies against Pax7, Ki67 (proliferating cell marker), MyoD (myogenic lineage marker). Measurement results for the ratios of Pax7+Ki67+, Pax7+Ki67- and Pax7-Ki67- cells are shown in the right two panels. Scale bars, 100 μm. Each dot in the box-and-dot plots represents one image analyzed. For WT HL SCs, 94 images from 3 rounds of sorting; For WT Dia SCs, 107 images from 3 rounds of sorting; For WT EOM SCs, 75 images from 3 rounds of sorting; For G93A HL SCs, 96 images from 3 rounds of sorting; For G93A Dia SCs, 62 images from 3 rounds of sorting; For G93A EOM SCs, 79 images from 3 rounds of sorting. For the 3 rounds of sorting, 1 was from male and 2 were from female mice. * P < 0.05; ** P < 0.01; **** P < 0.0001; ns, not significant (t-test). ANOVA P values are also shown. (B) Representative images of FACS-isolated SCs cultured for 4 days in growth medium and 2 days in differentiation medium stained with antibodies against Pax7, Ki67, myosin heavy chain (MHC, differentiated myotube marker). Measurement results for the ratios of Pax7+Ki67+ and Pax7+Ki67- cells, as well as the fusion indices are shown in the right two panels. Each dot in the box-and-dot plots represents one image analyzed. For WT HL SCs, 52 images from 3 rounds of sorting; For WT Dia SCs, 51 images from 3 rounds of sorting; For WT EOM SCs, 51 images from 3 rounds of sorting; For G93A HL SCs, 52 images from 3 rounds of sorting; For G93A Dia SCs, 47 images from 3 rounds of sorting; For G93A EOM SCs, 56 images from 3 rounds of sorting. For the 3 rounds of sorting, 1 was from male and 2 were from female mice. ANOVA P values are also shown.

We also compared myotube formation and cell proliferation of cultured SCs under differentiation conditions. After 4 days in growth medium, we switched to differentiation medium and cultured the cells for 2 more days. The cells were then stained for sarcomere myosin heavy chain (MHC), Pax7 and Ki67. The EOM-derived SCs exhibited a higher number of Pax7 positive cells (Pax7+/Ki67+ and Pax7+/Ki67-) compared to SCs derived from hindlimb and diaphragm muscles – from both WT and G93A mice (Figure 4B). In contrast, the fusion indices, which calculate the ratio of nucleus within the MHC positive myotubes, were the lowest for EOM SCs (Figure 4B). This does not necessarily mean the EOM SCs have deficiencies in differentiation after they are committed, but more likely a reflection of their superior preservation of actively proliferating population [45]. In support of this view, previous TA engraftment experiments showed that more myofibers were generated by EOM SCs than their limb counterparts [36].

Butyrate induces EOM-like preservation of renewability in FACS-isolated SCs under differentiation conditions in vitro

Since NaBu-supplemented feeding reduced SC depletion in end stage G93A mice (Figure 2) in vivo, we tested NaBu treatment on cultured SCs. G93A hindlimb SCs were cultured with 0, 0.1, 0.5, 1.0, 2.5 and 5 mM of NaBu added to the growth medium for 1 day. RNAs were collected for qRT-PCR to evaluate the short-term impact of NaBu on cell growth. The relative expression of the proliferation markers (Hmhg2 and Ccnd1) peaked at 0.5 mM. In contrast, the relative expression of proliferation inhibitor P21 bottomed at 0.5 mM (Figure 5A), indicating 0.5 mM as the optimal concentration of NaBu application in vitro.

Proliferation and differentiation properties of FACS-isolated SCs with different NaBu treatment conditions.

(A) qRT-PCR based relative quantification of Hmga2, Ccnd1 and P21 expression in FACS-isolated G93A HL SCs cultured with different doses of NaBu in growth medium for 1 day (n = 6, each dot in the box-and-dot plots represents one replicate of culture). HL SCs were from male mice. *** P < 0.001; **** P < 0.0001 (t-test). ANOVA P values are also shown. (B) Left two panels show the representative images of FACS-isolated G93A hindlimb and diaphragm SCs cultured in growth medium for 4 days and differentiation medium for 2 days and experienced different NaBu treatment conditions and stained with Pax7, Ki67 and MHC antibodies. Cells in top row were not treated with NaBu. Cells in the middle row were treated with NaBu for 1 day before the induction of differentiation (Day 4 of culture). Cells in the bottom row were treated with NaBu continuously for 3 days from the day before the induction of differentiation (Day 4-6 of culture). Scale bars, 100 μm. Measurement results for the ratios of Pax7+Ki67+ and Pax7+Ki67- cells, as well as the fusion indices are shown in the right two panels. Each dot in the box-and-dot plots represents one image analyzed. For G93A HL SCs, 52 images from 3 rounds of sorting; 1-day NaBu treatment, 45 images from 3 rounds of sorting; 3-day NaBu treatment, 51 images from 3 rounds of sorting; For G93A Dia SCs, 47 images from 3 rounds of sorting; 1-day NaBu treatment, 60 images from 3 rounds of sorting; 3-day NaBu treatment, 57 images from 3 rounds of sorting. For the 3 rounds of sorting, 2 were from male and 1 was from female mice. * P < 0.05; ns, not significant (t-test). ANOVA P values are also shown.

We designed two NaBu treatment plans for G93A diaphragm and hindlimb SCs as follows: one with 0.5 mM NaBu applied in the growth medium for 1 day before switching to differentiation medium (1 day in growth medium), the other with NaBu continuously applied for 3 days (1 day in growth medium + 2 days in differentiation medium). As shown in Figure 5B, the G93A hindlimb and diaphragm SCs in culture showed significant increase in the ratio of Pax7 positive cells under the 1-day NaBu treatment, associated with a decrease in fusion index. This trend was more pronounced under the 3-day NaBu treatment. These data indicate that NaBu treatment results in EOM-like preservation of renewability of the G93A SCs (Figure 4B). Aside from the drop of fusion indices after 3-day treatment with NaBu, there was notable decrease of MHC levels, indicating long-term exposure to NaBu did inhibit the differentiation process (Figure 5B), which is in line with the published result observed in myoblasts of different origins [46–48].

EOM-derived SCs show distinct transcriptome profile supporting self-renewal and axonal growth when compared to the SCs derived from other muscle origins

We extracted RNAs from SCs of hindlimb, diaphragm and EOMs under the following two culture conditions: first, in growth medium for 4 days (with suffix “G”); second, in growth medium for 4 days and then in differentiation medium for 2 days (with suffix “D”). The RNAs were send for bulk RNA-Seq. Additionally, RNA-Seq analyses were conducted for SCs of G93A hindlimb and diaphragm muscles cultured with NaBu for 3 days (with suffix “NaBu3”) (Figure 6).

Comparing the transcriptomic profile and homeostasis preferences of SCs from different muscle origins cultured in growth and differentiation medium.

(A) PCA analysis and hierarchical clustering results of different RNA-Seq samples. Samples with suffix “G” were SCs cultured in growth medium for 4 days. Suffix “D” denote SCs cultured in growth medium for 4 days and differentiation medium for 2 days. Suffix “NaBu3” denote SCs treated with 0.5 mM NaBu continuously for 3 days (Day 4-6 of culture). Dashed-line circles in PCA plot highlight samples close to each other in distance. All SC samples shown here were from female mice. (B) We identify genes differentially expressed in EOM SCs compared to their diaphragm and HL counterparts cultured in growth medium (leftmost panel) and differentiation medium (middle panel) by screening for genes shared by the four DEG lists (gold, magenta, pink and green in color) shown in the Venn diagrams. All DEG lists were generated based on the standard of |log2FC|>= 0.4 and FDR<= 0.001. Yellow box highlights the shared genes. To identify NaBu treatment signature genes, we screened for genes shared by the two DEG lists (green and gold in color) shown in the rightmost Venn diagram. (C) Dot plots in the upper row show the Log2FC of activation signature genes (magenta) and differentiation signature genes (cyan) identified in the 1119, 3318 and 3018 genes highlighted in yellow boxes in (B). Dot plots in the lower row show the Log2 FC of quiescence signature genes identified in the yellow-box highlighted genes in (B). Additionally, genes belonging to the following subgroups of quiescent signature genes: most commonly used SC markers, Notch signaling components and pluripotency signaling components are colored in dark blue, red and gold, respectively.

Similarities of the transcriptomic profiles of RNA-Seq samples were examined using the Principle Component Analysis (PCA) and hierarchical clustering (Figure 6A). We found that both WT and G93A EOM SCs cultured in differentiation medium were more similar to hindlimb and diaphragm SCs in growth medium than their counterparts in differentiation medium, highlighting the superior renewability of EOM SCs in differentiation medium and is consistent with the immunostaining results showed in Figure 4B.

Considering that cultured WT and G93A SCs from the same muscle origins exhibited relatively similar gross transcriptomic profiles in PCA and hierarchical clustering analysis, and that the proliferation and differentiation properties of cultured WT and G93A SCs from the same muscles were only marginally different in immunofluorescence assays (Figure 4-figure supplement 2), we speculate that the differential homeostasis seen in the FACS profiles of hindlimb and diaphragm SCs (Figure 3) may relate to the pathological milieu in vivo. When G93A SCs are isolated and cultured in vitro, their properties become similar to those of WT counterparts.

Additionally, we listed top 20 differentially expressed genes (DEGs) (ranked by Log2FC, both the upregulated and downregulated) by comparing: (1) EOM SCs to their hindlimb and diaphragm counterparts (Figure 6-Source Data 1); (2) G93A SCs to WT SCs of the same muscle origin (Figure 6-Source Data 2); (3) G93A hindlimb and diaphragm SCs with and without 3 day-NaBu treatment (Figure 6-Source Data 3).

The top 20 DEGs comparing EOM SCs to their hindlimb and diaphragm counterparts are mainly homeobox transcription factors (Figure 6-Source Data 1), reflecting different anatomic origin of these SCs [49]. For example, Alx4 and Pitx1 are expressed significantly higher in EOM SCs, while Lbx1, Tbx1, Hoxa and Hoxc genes are expressed significantly lower in EOM SCs than hindlimb and diaphragm counterparts acquired from either WT or G93A mice. These observations are consistent with previous reports [50,51]. Additionally, Pax3 is expressed significantly lower in EOM SCs than hindlimb and diaphragm counterparts cultured in growth medium, which reflects the unique Pax3-independent induction of myoD expression reported in EOM SCs [50]. It is also worth mentioning that BMP4 is expressed significantly higher in G93A EOM SCs compared to G93A hindlimb and diaphragm counterparts cultured in differentiation medium, potentially contributing to the continued expansion and slowed myogenic differentiation of EOM SCs [52]. This trend is also seen in SCs acquired from WT mice, but not pronounced enough to make it to the top 20 list. However, most of the top 20 differentially expressed genes between G93A and WT SCs (Figure 6-Source Data 2) are with poor functional annotation records and there exist great variations in the lists between muscles of different origins. We also examined the top 20 differentially expressed genes following 3 day-NaBu treatment in G93A hindlimb and diaphragm SCs. It did not turn out to be very informative (Figure 6-Source Data 3).

To better understand RNA-Seq data in the perspective of SC homeostasis, which refers to the transition between quiescence (dormant, slow self-renewal), activation (rapid amplification with symmetric and asymmetric cell division) and differentiation states during tissue regeneration [45], we compiled the lists of quiescence, activation and differentiation signature genes respectively. For quiescence signature genes, a list of 507 genes highly upregulated in quiescent SCs compared to activated SCs have been generated previously by Fukada et al [53]. We adopted this list together with other 18 genes denoted by other investigators, including Notch1, Notch2, Hes1, Hey1, Rbpj, Dtx4, Jag1, Sox8, Sdc3, Itga7, Itgb1, Met, Vcam1, Ncam1, Cdh2, Cdh15, Cxcr4, Cav1 [45,54–57]. The activation signature gene list was generated from our RNA-seq data using those expressed significantly higher in SCs (both from WT and G93A) in growth medium (1187 genes in total, |log2FC| ≥ 0.4 and FDR ≤ 0.001). Accordingly, the differentiation signature gene list was generated from those expressed significantly higher in SCs cultured in differentiation medium (1349 genes in total).

To verify that our signature gene lists recapitulate the SC properties at activation and differentiation state reported in previous studies, we did functional annotation analysis (KEGG pathway, Gene ontology_biological process, Gene ontology_cellular component). As shown in Figure 6-figure supplement 4, the top functional annotations enriched in our activation signature genes are all related to cell cycle and DNA replication, while the top functional annotations for our differentiation signature genes are associated with muscle contraction, muscle-specific subcellular structures and metabolic changes.

To explore whether the homeostatic states of EOM SCs are different from that of SCs from other muscle origins, we looked for differentially expressed genes (DEGs) shared-in-common among the following four groups: WT EOM SC-vs-WT diaphragm SC; WT EOM SC-vs-WT HL SC; G93A EOM SC-vs-G93A diaphragm SC; G93A EOM SC-vs-G93A HL SC (Figure 6B, left two panels). Within the 1119 shared DEGs identified in SCs cultured in growth medium, slightly more activation signature genes and quiescence signature genes were expressed higher in EOM SCs compared to diaphragm and hindlimb derived SCs. Meanwhile more differentiation signature genes were expressed lower in EOM SCs (Figure 6C, leftmost panel and Figure 6-Source Data 4). These trends became much more prominent in the 3318 shared DEGs identified in SCs cultured in differentiation medium, implying that EOM SCs are notably superior at preserving renewability in a differentiation environment than their diaphragm and hindlimb counterparts (Figure 6C, middle panel). Consistently, subgroups of quiescence signature genes belonging to commonly used stem cell markers (Pax7, Cd34, Cxcr4, Vcam1, Sdc3, Sdc4, Cdh15, Met, Cav1), Notch signaling pathway (Notch1, Notch2, Notch3, Jag1, Heyl, Dtx4, Msc) and pluripotency signaling pathway (Klf4, Igf1, Bmp2, Bmp4, Fzd4, Fgfr4, Pik3r1, Lifr) were all expressed higher in EOM SCs in differentiation medium than SCs derived from diaphragm and hindlimb muscles (Figure 6C, middle panel and Figure 6-Source Data 5).

In parallel, we examined whether the 3-day NaBu treatment could alter the homeostasis of G93A diaphragm and hindlimb SCs (Figure 6B, rightmost panel). There were 3018 DEGs shared between the following two comparing groups: G93A diaphragm SC_D_NaBu3-vs-G93A diaphragm SC_D and G93A HL SC_D_NaBu3-vs-G93A HL SC_D. Similar to the trends observed in EOM SCs in differentiation medium, the majority of activation and quiescence signature genes found in these DEGs were upregulated in the NaBu treated group, while the differentiation signature genes were downregulated (Figure 6C, rightmost panel). The elevated quiescence signature genes of the NaBu treated group include commonly used stem cell markers (Pax7, Cd34, Vcam1, Sdc4, Cdh2, Met, Cav1, Itga7), Notch signaling pathway members (Notch3, Rbpj) and pluripotency signaling pathway components (Fzd4, Bmp4). However, other three pluripotency related genes (Klf4, Wnt4 and Bmp2) were found downregulated (Figure 6-Source Data 6), indicating that NaBu treatment also has a mixed impact on the stemness of the SCs.

To compile EOM SCs signature gene list, we identified 621 genes expressed higher in EOM SCs than diaphragm and hindlimb SCs cultured in growth medium from the 1119 DEGs (Figure 6B, leftmost panel). Next, we pulled out 2424 genes that are expressed higher in EOM SCs than diaphragm and hindlimb SCs cultured in differentiation medium from the 3318 DEGs (Figure 6B, middle panel). These two lists have 478 genes shared-in-common. Particularly, many of these genes are related to axon guidance, cell migration and neuron projection (Figure 7A). Figure 7A-figure supplement 1, Figure 7-Source Data 1 and 2 further demonstrate this phenotype when the functional annotation analysis was done separately for the growth or differentiation culture conditions. The functional annotation analysis for the 3018 genes altered by the 3-day NaBu treatment (Figure 6B, rightmost panel) also spotted axon guidance, cell migration and neuron projection on the top of the annotation lists (Figure 7B and Figure7-Source Data 3). Importantly, the chemokine Cxcl12 shows up in all three lists. Cxcl12 is known to promote axonal extension and NMJ regeneration [58]. These data suggest that enhanced attraction of motor neuron axons by peri-NMJ SCs might be a shared mechanism underlying the NMJ preservation in EOMs and NaBu treatment observed in G93A mice.

Functional annotation analysis of EOM SC signature genes and NaBu treatment signature genes and qRT-PCR results.

(A) Top KEGG pathway, GO_P and GO_C annotations of the 478 EOM signature genes (expressed higher in EOM SCs compared to diaphragm and HL counterparts cultured in both growth medium and differentiation medium). Arrows highlight annotations related to axon guidance and cell migration. (B) Top KEGG pathway, GO_P and GO_C annotations of the 3018 NaBu treatment signature genes shown in Figure 6B, rightmost panel. (C) qRT-PCR based relative quantification (RQ, fold of change against WT HL SCs) of Hmga2 (activation signature gene), Actn3 (differentiation signature gene), Notch3 (quiescence signature gene) and Cxcl12 (axon guidance molecule) expression in HL, diaphragm and EOM SCs cultured in growth (upper row) and differentiation medium (lower row), respectively. RNA samples were collected from 3-6 rounds of sorting and sorted cells were seeded into 3 dishes as replicates. Each dot in the box-and-dot plots represents one replicate of culture. WT HL SCs were from 3 male and 1 female mice. WT diaphragm SCs were from 2 male and 2 female mice; WT EOM SCs were from 3 male mice; G93A HL SCs were from 4 male and 2 female mice. G93A diaphragm SCs were from 1 male and 3 female mice; G93A EOM SCs were from 3 male mice. * P < 0.05; ** P < 0.01; *** P < 0.001; **** P < 0.0001; ns, not significant (t-test). ANOVA P values are also shown. (D) qRT-PCR based relative quantification (fold of change against G93A HL SCs without NaBu treatment) of Hmga2, Actn3, Notch3 and Cxcl12 expression in G93A hindlimb and diaphragm SCs with or without the 3-day NaBu treatment. RNA samples were collected from 6 rounds of sorting and sorted cells were seeded into 3 dishes as replicates. G93A HL SCs were from 4 male and 2 female mice; G93A diaphragm SCs were from 2 male and 4 female mice. ANOVA P values are also shown.

We selected one representative gene from each of the following categories: activation (Hmga2), differentiation (Actn3) and quiescence (Notch3) signature genes, as well as an axon guidance gene (Cxcl12) for qRT-PCR. For each muscle type and culture condition, we collected samples from 3-6 rounds of sorting (the individual results are listed in Figure 7-Source Data 4-6). We observed significantly higher expression of Hmga2 in EOM SCs compared to diaphragm and hindlimb SCs cultured in either growth or differentiation medium (Figure 7C, leftmost panel), confirming the superior renewability of EOM SCs. The 3-day NaBu treatment of diaphragm and hindlimb SCs derived from G93A mice also upregulated Hmga2 (Figure 7D, leftmost panel). Actn3 was expressed lower in EOM SCs than diaphragm SCs cultured in either growth or differentiation medium, but not necessarily lower than hindlimb SCs (Figure 7C, 2nd panel from the left). Meanwhile NaBu treatment significantly decreased Actn3 expression (Figure 7D, 2nd panel from the left). Consistent with the RNA-Seq data, the quiescence marker Notch3 was expressed significantly higher in EOM SCs than diaphragm and hindlimb SCs cultured in differentiation medium but not in growth medium (Figure 7C, 2nd panel from the right). Similarly, NaBu treatment also elevated Notch3 expression (Figure 7D, 2nd panel from the right) in G93A diaphragm and hindlimb SCs cultured in differential medium. Importantly, the axon guidance molecule Cxcl12 was expressed significantly higher in EOM SCs than diaphragm and hindlimb SCs cultured in both growth and differentiation medium (Figure 7C, rightmost panel). NaBu treatment elevated Cxcl12 expression as well (Figure 7D, rightmost panel), confirming enhanced axon attraction as a common feature shared by EOM SCs and NaBu treatment.

AAV mediated overexpression of Cxcl12 in myotubes promotes motor neuron axon extension

Since recombinant Cxcl12 is known to promote axon extension [58], we tested whether overexpressing Cxcl12 in myotubes via adeno-associated virus (AAV) transduction could achieve the same effect. For these experiments, we co-cultured motor neurons and SC-derived myotubes on compartmentalized microfluidic chambers (XonaChip). Because AAV vectors do not transduce satellite cells efficiently [59,60], we first seeded the satellite cells into one compartment of the XonaChips, then induced them to differentiate for two days before AAV transduction. Four types of AAV-transduced myotubes were used for the coculture assays: a) WT hindlimb SC-derived myotubes transduced with AAV-CMV-eGFP; b) G93A hindlimb SC-derived myotubes transduced with AAV-CMV-eGFP; c) G93A hindlimb SC-derived myotubes transduced with AAV-CMV-Cxcl12-IRES-eGFP; and d) G93A EOM SC-derived myotubes transduced with AAV-CMV-eGFP. One day after AAV transduction, rat spinal motor neurons (RSMN) were seeded into the other compartment of the XonaChips and the culture medium switched to Neurobasal Plus Medium (NBPM) (Figure 8A). The cocultures were maintained for 9 more days. Immunostaining confirmed increased presence of Cxcl12 protein in the form of cytosolic vesicles in myotubes transduced with AAV8-CMV-Cxcl12-IRES-eGFP but not those transduced with AAV8-CMV-eGFP (Figure 8-figure supplement 1). These are likely vesicles for secretion [61].

Coculture of AAV transduced SC derived myotubes with rat spinal motor neurons in compartmentalized microfluidic chambers.

(A) Schematic representation of XonaChip spatial configuration and coculture experiment timeline. NBPM: Neurobasal Plus Medium; RSMNs: rat spinal motor neurons. (B) Representative composite images of RSMNs cocultured with: 1. WT HL SC-derived myotubes transduced with AAV-CMV-eGFP; 2. G93A HL SC-derived myotubes transduced with AAV-CMV-eGFP; 3. G93A HL SC-derived myotubes transduced with AAV-CMV-Cxcl12-IRES-eGFP; 4. G93A EOM SC-derived myotubes transduced with AAV-CMV-eGFP. Arrows highlight RSMN neurites crossed the microgrooves but stayed close to the edges and grew backward. Also see Figure 8-figure supplement 2A. Boxed regions are enlarged below. Scale bars: 50 μm. (C) Zoom-in views of RSMN neurites innervated or missed the AChR clusters (BTX positive patches). Boxed regions are further enlarged below. Scale bars: 50 μm, 10 μm. (D) Measured lengths of the longest neurites derived from RSMN cells (presumably axons) crossing the 450 μm microgrooves. n = 24, 31, 34,31 for the 4 groups, respectively (from 3 rounds of coculture experiments). Each dot in the box-and-dot plot represents one neurite measured. HL and EOM SCs used for co-culture experiments were all from male mice. ** P < 0.01; **** P < 0.0001 (t-test). ANOVA P value is also shown.

By measuring the lengths of the longest RSMN neurites (presumably axons) crossing the 450 μm microgrooves, we noticed that hindlimb-SC derived myotubes from G93A mice, but not from WT littermates, appeared repulsive to distal axons. The axons stayed close to the edge of the microgrooves and even grew backward towards the neuronal compartment, which we termed “edge effect” (Figure 8B and Figure 8-figure supplement 2A, arrows). This was rarely seen in the WT hindlimb SC-derived myotubes. In contrast, RSMN axon extension was dramatically enhanced in the co-culture system either with G93A hindlimb SC-derived myotubes overexpressing Cxcl12, or G93A EOM SC-derived myotubes overexpressing GFP (as control) (Figure 8B, C). Those neurites located close to the microgrooves were often seen to closely grow along nearby myotubes (Figure 8-figure supplement 2B). G93A hindlimb SC-derived myotubes overexpressing Cxcl12 occasionally were even seen to overcome the “edge effect” and redirect the axons towards their proximities (Figure 8-figure supplement 2B, arrow) - a persuasive indication of chemoattractive effects of those myotubes. Meanwhile, RSMN innervation of AChR clusters (BTX positive patches) were not seen in co-cultures with G93A hindlimb SC-derived myotubes overexpressing GFP, while they were consistently observed in the other three types of AAV-transduced myotubes (Figure 8C).

The overall innervation incidence was too low for statistical analysis due to the sparsity of axons in the myotube compartments, thus we conducted co-culture experiments where RSMNs were added to SC-derived myotubes in a uni-compartmental co-culture setting. As shown in Figure 9A-C and Figure 9-figure supplement 1A-C, when co-cultured in a single compartment, Cxcl12 overexpressing G93A hindlimb SC-derived myotubes exhibited increased innervation compared to GFP overexpressing controls. Furthermore, individual Cxcl12 overexpressing EOM SC derived myotubes were frequently innervated by multiple (3 or more) times (Figure 9A-C and Figure 9-figure supplement 1D), recapitulating the “multi-innervation” property of EOMs in vivo [26]. The extra “anchor points” seemed to encourage better spatial alignment of the RSMN axon network to the myotubes globally in the co-cultures. This spatial alignment of RSMN axons and myotubes was not observed in co-cultures of RSMNs and hindlimb SC-derived myotubes (without Cxcl12 overexpression), which also did not exhibit multi-innervation (Figure 9A-C and Figure 9-figure supplement 1A-C). Under higher magnification, we often observed that axons sending branches to adjacent BTX positive patches on EOM SC-derived myotubes (Figure 9B, arrows). Of note, there was higher expression of several Netrin and Semaphorin family members reported to promote axon-branching in EOM SCs compared to other muscle-derived SCs cultured in differentiation medium (Figure 7-Source Data 2) [62–67]. This milieu, comprised of abundant axon guidance molecules, in combination with the robust self-renewability of satellite cells, may contribute to the relative EOM NMJ resistance in ALS.

Coculture of AAV transduced SC derived myotubes with rat spinal motor neurons within the same compartment.

(A) Representative composite images of RSMNs cocultured with: 1. WT HL SC-derived myotubes transduced with AAV-CMV-eGFP; 2. G93A HL SC-derived myotubes transduced with AAV-CMV-eGFP; 3. G93A HL SC-derived myotubes transduced with AAV-CMV-Cxcl12-IRES-eGFP; 4. G93A EOM SC-derived myotubes transduced with AAV-CMV-eGFP. The timeline was the same as that in Figure 8 but RSMNs were seeded on top of myotubes. Boxed regions are enlarged in (B). Scale bars: 50 μm. (B) Zoom-in views of innervated and non-innervated AChR clusters on myotubes having contact with RSMN neurites. Arrows highlight branching phenomena observed in RSMN neurites innervating AChR clusters. Scale bars: 10 μm. (C) Quantification of innervation ratios of AChR clusters (area > 10 μm2) in myotubes in contact with RSMN neurites. n = 37, 48, 40, 52 z-stack images for the 4 groups, respectively (from 3 rounds of coculture experiments). Also see methods for detailed description. Each dot in the box-and-dot plot represents one image analyzed. HL and EOM SCs used for co-culture experiments were all from male mice. *** P < 0.001; **** P < 0.0001; ns, not significant (t-test). ANOVA P value is also shown. (D) Graphic summary of the potential mechanisms contributing to EOM NMJ integrity maintenance during ALS progression mimicked by NaBu treatment.

Discussion

It is known that the EOMs are complex muscles. Besides the developmental myosin isoforms, EOMs also express both adult fast and slow myosin contractile elements [68], suggesting that the sparing may not be solely linked to the fast or slow twitch nature of the muscle fibers, rather the changes in SCs may play a pivotal role in preserving the EOM function during the progression of ALS. We have demonstrated an association between the degeneration of NMJs and depletion of peri-NMJ SCs in hindlimb and diaphragm muscles in vivo in end-stage G93A mice. In contrast, EOM NMJs exhibited no signs of denervation or SCs depletion. The FACS profile of SCs isolated from EOM suggests spontaneous activation even without pathological stimulation, which is supported by previous in vivo BrdU labeling and staining results [69]. Furthermore, immunostaining, qRT-PCR and RNA-Seq analyses all imply that EOM SCs in vitro possess superior renewability better than hindlimb and diaphragm SCs, under differentiation-inducing culture conditions. Comparative functional annotation analyses of the transcriptomes highlight higher expression of axon guidance molecules in EOM SCs. Neuromuscular coculture experiments demonstrate that EOM SC-derived myotubes are more attractive to potential axons derived from rat spinal motor neurons than their hindlimb counterparts – which in fact appeared to repel the branching axons. Thus, the unique homeostasis regulation and axon attraction properties of EOM SCs may contribute to the resistance to denervation of EOM under ALS (Figure 9D).

Earlier report of SC quantification from muscle biopsy samples using electron microscopy detected no significant difference between ALS patients and non-ALS controls [70]. Other studies using primary cultured SCs from muscle biopsy samples implied compromised differentiation program in ALS SCs compared to controls [71,72]. As to proliferation, one study reported enhanced proliferation in ALS SCs [72], while the other cued for deficits in proliferation in ALS SCs [71]. The various findings may reflect variations in the disease stage of patients during sampling. For animal studies, Manzano et al measured the amount of Pax7+ SCs in dissociated EDL and soleus myofibers from G93A mice and WT controls and observed large variations across muscle-type and disease stage [73]. In another study carried out by the same group, primarily cultured EDL and soleus SCs from G93A mice generally showed lower proliferation potential than the WT counterparts [74].

In our whole-mount muscle imaging data and FACS profiles of isolated SCs, SC depletion, at least for the most quiescent population, is observed in hindlimb and diaphragm muscles from end-stage G93A mice. Additionally, the heterogeneity of Vcam1 levels observed in hindlimb and diaphragm SCs derived from end stage G93A mice compared to those from WT controls indicates occurrence of denervation-related pathological activation. Although this activation may transiently elevate the number of SCs, the continuous dominance of asymmetric over symmetric cell division, as well as the differentiation pressure to regenerate myofibers could gradually exhaust the quiescent stem cell pool, leading to the depletion phenotype seen in the end stage G93A mice. In contrast, the superior preservation of renewability of EOM SCs seen in differentiation medium is an effective measure against this catastrophe. However, the spontaneous activation and superior renewability of EOM SCs may not always be an advantage. As mentioned in the introduction section, EOMs are preferentially involved in some other neuromuscular disorders. In the case of mitochondrial myopathies, mitochondrial DNA defects accumulate over time, resulting in compromised oxidative phosphorylation represented by cytochrome c oxidase (COX) deficient myofibers in patients [7,8]. It is possible that the more frequent self-renewal and spontaneous activation of EOM SCs contribute to higher rate of mitochondrial DNA replication, leading to accelerated spreading of mitochondrial DNA defects, resulting in higher proportion of COX-deficient myofibers than other muscles, such as EDL, whose SCs remain dormant under physiological conditions.

In this study, overexpressing Cxcl12 in G93A hindlimb SC-derived myotubes promotes motor neuron axon extension under the co-cultured condition. Cxcl12 has been known as a critical guidance cue for motor and oculomotor axons during development [75,76], as well as the migration of neural crest cells to sympathetic ganglia in the formation of sympathetic nerve system [77]. In adult animals, peri-synaptic Schwann cells have recently been reported to transiently produce Cxcl12 after acute motor axon terminal degeneration [58]. Intraperitoneal injection of antibodies neutralizing secreted Cxcl12 protein slows down the regeneration process. In contrast, the exposure of primary motor neurons to recombinant Cxcl12 protein stimulates neurite outgrowth and directional axonal extension in compartmentalized culture setup, which could be reversed by the treatment of the Cxcr4 specific inhibitor AMD3100. Published studies have demonstrated that local administration of Cxcl12 protein to neurotoxin injected hindlimb muscles of mice improved evoked junctional potentials of soleus [58]. Thus, the axon guidance role of Cxcl12 has been consolidated both in vitro and in vivo. Additionally, myoblasts also express Cxcr4 and their fusion into myofibers/myotubes involves Cxcl12 guidance as well [78–80]. The abundant peri-NMJ SCs in EOMs expressing Cxcl12 may achieve synergistic effects with peri-synaptic Schwann cells, improving the long-term integrity of NMJs. However, it is possible that motor neurons carrying ALS mutations may respond differently to Cxcl12 mediated axon guidance than WT motor neurons. This is a limitation of the current study, which will be investigated in future co-culture studies.

Epigenetic regulation of SC transcriptome by histone deacetylase (HDAC) inhibitors, such as butyrate [81], could modulate the myogenesis process both in vitro and in vivo [46,48,82,83]. Butyrate is a gut-bacteria fermentation product that also acts as an energy source to improve mitochondrial respiration [81,84]. Previous studies by our lab and others showed NaBu-supplemented feeding prolongs the life span of G93A mice and reverses CypD-related mitoflash phenotypes in myofibers derived from G93A mice, which implies a potential decrease of oxidative stress. Additionally, NaBu treatment also improved mitochondrial respiratory function of NSC34 motor neuron cell line overexpressing hSOD1G93A [85,86]. NaBu treatment has been reported by other groups to directly promote neuronal survival and axonal transport by inhibiting HDAC6 [87–90]. Aside from neurons, feeding NaBu also changed intestinal microbiota and intestinal epithelial permeability of G93A mice [37]. The effect of NaBu treatment is certainly multi-faceted. Yet previously it was not known whether the beneficiary effects of NaBu are linked to any phenotypic changes of SCs, and whether NaBu treatment induced epigenetic modulations similar to that of EOM SCs. The study here revealed its beneficiary role on the preservation of NMJ integrity and peri-NMJ SCs affected in hindlimb and diaphragm muscles by ALS progression. NaBu treatment of G93A hindlimb and diaphragm SCs in culture induced transcriptomic changes mimicking some features seen in EOM SCs, including the upregulation of activation and quiescence signature genes and the downregulation of differentiation signature genes, indicating NaBu improved the renewability of G93A hindlimb and diaphragm SCs. Additionally, some axon guidance molecules, particularly Cxcl12, were also induced by NaBu application (Figure 9D). Thus, NaBu-induced EOM SC-like transcriptomic pattern could be a crucial contributor to its therapeutic effect against ALS progression. Combining NaBu treatment with application on top of Cxcl12 overexpression may have additive therapeutic benefits to slow ALS progression.

Overall, our studies have defined several mechanisms which (1) may explain why EOMs are relatively spared in ALS, and (2) provides a rationale for targeting these differences as a therapeutic strategy. Furthermore, analysis of these signatures in patient muscle biopsies could be used to select subsets of ALS patients who may be most likely to benefit from therapies targeting this signaling network, and to provide possible "response biomarkers" in pre-clinical and clinical studies. As a cocktail of sodium phenylbutyrate (a derivative of NaBu) and Taurursodiol (an antioxidant) under the Commercial name of Relyvrio has been recently approved in US and Canada as a new ALS therapy [91], our study could help gain more insights into the molecular mechanisms related to this type of ALS therapy.

Materials and methods

Key Resources Table

Animals

All animal experiments were carried out in accordance with the recommendations in the Guide for the Care and Use of Laboratory Animals of the National Institutes of Health. The protocol on the usage of mice was approved by the Institutional Animal Care and Use Committee of the University of Texas at Arlington (approved IACUC protocol #A19.001). WT mice used in this study were of B6SJL background. The ALS transgenic mouse model (hSOD1G93A) with genetic background of B6SJL was originally generated by Drs. Deng and Siddique’s group at Northwestern University and deposited to the Jackson Lab as B6SJL-Tg (SOD1*G93A) [92]. For the butyrate feeding experiment, the animals were fed with 2% NaBu (SAFC ARK2161) from the age of 3 months (at the ALS disease onset) and lasting for 4 weeks.

Primary culture of satellite cells isolated by fluorescence-activated cell sorting

The hindlimb muscles including tibialis anterior (TA), extensor digitorum longus (EDL), gastrocnemius, and quadriceps were dissected from one mouse for each round of sorting. For diaphragm, both the crural and costal domains were dissected from one mouse for each round of sorting. For EOMs, the four rectus muscles in the eye sockets were dissected from 5-6 mice of the same gender for each round of sorting [93]. The collected muscles were placed in 0 mM Ca2+ Ringer’s solution. Excessive connective tissue, tendons, fat, blood clots and vessels were removed as much as possible. After cleanup, the muscles were transferred into the digestion media (DMEM containing 5% FBS, 1% Pen/Strep and 10 mM HEPES) and minced into small blocks using scissors. Collagenase II (Worthington LS004176, ≥125 units/mg), dispase II (Sigma D46931G, ≥0.5 units/mg), hyaluronidase (Worthington LS002592, ≥300 USP/NF units/mg) and Dnase I (Worthington LS002139, ≥2,000 Kunitz units/mg) were added at the final concentration of 0.26%, 0.24%, 0.16% and 0.04% (weight/volume), respectively. The digestion system was placed in an orbital shaker running at 70 rpm at 37 °C for 45 min. The digested muscles were triturated 10 times with a 20-gauge needle attached to a 5 ml syringe. Afterwards the triturated mixture was pipetted onto a pre-wetted 100 μm strainer and centrifuge at 100 g for 2.5 min to precipitate the bulky myofiber debris. The supernatant was transferred onto a pre-wetted 40 um strainer and cells were collected by centrifuged at 1200 g for 6.5 min. The cells were resuspended in sorting media (PBS containing 0.3% BSA, 10 mM HEPES and 1 mM EGTA) and stained with antibodies or dyes at 4 °C for 45 min with shaking. The antibodies used include PE anti-mVcam1 (Biolegend 105713, 600 ng/106 cells), APC anti-mCD31 (Biolegend 102409, 400 ng/106 cells), APC anti-mCD45 (Biolegend 103111, 400 ng/106 cells) APC anti-mSca-1 (Biolegend 108111, 200 ng/106 cells). This antibody combination followed a previous published protocol [41]. Vcam1 is a SC marker, CD31 is a endothelial cell marker, CD45 is a hematopoietic cell marker and Sca-1 is a mesenchymal stem cell marker [41]. For sorting, our target population is VCAM+CD31-CD45-Sca-1-. When needed, the cells were stained with methyl green (2 µg/ml) to check mortality rate. As a positive control, half of the cells were transferred to a new Eppendorf tube and methanol (precooled at -20°C) was added dropwise with gentle mixing and left on ice for 5 min, the dead cells were wash twice before being mixed with the rest half of cells for methyl green staining. The sorting process was conducted using BD FACS Melody Cell Sorter using a 100 µm nozzle. The excitation laser and emission filter setup for methyl green is the same as APC. The flow rate was tuned between 1-7 to adjust the event rate to 2000-4000/sec. Gating conditions of FSC-A/SSC-A, SSC-H/SSC-W, FSC-H/FSC-W, Vcam1/CD31_CD45_Sca-1 and Vcam1/Methyl green plots were shown in Figure 3-figure supplement 1 and Figure 3-figure supplement 2. The sorted cells were collected in growth media (DMEM containing 20%FBS, 1% Pen/Strep, 1% chick embryonic extract and 10 mM HEPES). Afterwards the cells were seeded into laminin (Corning 354232) -coated plates or glass-bottom dishes containing growth media (without HEPES). Culture media was changed on the third day and the fourth day. To induce differentiation, cells were washed three times with PBS and cultured in differentiation media (DMEM containing 2% horse serum) for 2 more days.

Neuromuscular coculture assay

For neuromuscular coculture in compartmentalized chambers (XonaCHIPS XC450), XoanCHIPS were coated with poly-L-ornithine (0.5 mg/ml) for 1 hr at 37 °C followed by laminin (0.1 mg/ml) for 1 hr at 37 °C. Primary cultured satellite cells were seeded at 1×10^5 into the myotube compartment in growth medium for attachment overnight and then switched to differentiation medium and cultured for 2 more days. Afterwards AAV8-CMV-eGFP (Vector Biolabs, 1×10^11 GC/ml) or AAV8-CMV-Cxcl12-IRES-eGFP (Vector Biolabs, 1.5×10^11 GC/ml, the Cxcl12 gene is of mouse origin) transduction was conducted for one day. The usage of higher titer for the second virus is for matching GFP signal intensity, as IRES-dependent expression is significantly lower than those directly driven by CMV [94]. The chambers were washed with Neurobasal Plus Medium twice before seeding rat spinal motor neurons into the neuronal compartment (1×10^5). The coculture was kept in Neurobasal Plus Medium for 9 more days and half medium change was conducted every 2 days.

For neuromuscular coculture in the same glass-bottom chamber, 8-well chambered cover glasses (Cellvis C8-1.5H-N) were coated with poly-L-ornithine and laminin as described above. Primary cultured satellite cells were seeded at 1×10^5/chamber in growth medium for attachment overnight and then switched to differentiation medium and cultured for 2 more days. AAV transduction was conducted for one day as described above. The infected myotubes were washed with Neurobasal Plus Medium (Thermo Fisher A3582901, with B-27 and GlutaMax supplement) before the addition of rat spinal motor neurons (4×10^3/chamber, iXCells 10RA-033) to the same chamber. The coculture was kept in Neurobasal Plus Medium for 9 more days and half medium change was conducted every 2 days.

Immunofluorescence (IF) and imaging of whole mount muscle samples

EDL, soleus and EOM (the four rectus muscles) were dissected out and fixed in 4% paraformaldehyde (PFA) for 1 hr. Afterwards the samples were washed with PBS containing 1% glycine once and two more times with PBS. Further fixation and permeabilization was done with methanol at -20°C for 10 min. For diaphragm, the costal muscles and their surrounding rib cages were dissected and immersed in methanol directly for fixation (PFA fixation was skipped because it made the diaphragm harder to flatten and introduced more non-specific noises in staining). The samples were rehydrated and washed with PBS. The epimysium and tendons were removed as much as possible. EDL and soleus were further split into smaller bundles along the longitudinal axis. Then the samples were immersed in blocking buffer (PBS containing 2% BSA, 2% horse serum, 0.1% Tween-20, 0.1% Triton X-100 and 0.05% sodium azide) overnight at 4 °C. Next day the samples were incubated with primary antibodies for 1-2 days. After washing with PBS for three times, the samples were incubated with Alexa Fluor labelled secondary antibodies (Thermo Fisher 1:1000) and Alexa Fluor 488 conjugated α-Bungarotoxin (Thermo Fisher B13422, 1:1000) for 1-2 days. The samples were then washed with PBS, counterstained with DAPI and mounted in antifade mounting media (20 mM Tris, 0.5% N-propyl gallate, 80% glycerol) for imaging. Primary antibodies used: neurofilament (DSHB 2H3 concentrate, 1:250), synaptophysin (Thermo Fisher PA11043, 1:400), Pax7 (SCBT sc-81648, 1:100), Cxcl12 (R&D MAB350, 1:200).

Z-stack scans of glycerol-cleared whole muscles were obtained using 40x high working distance lens of Leica TCS SP8 confocal microscope. The z-stacks were compacted into 2D images by maximal intensity projection. To quantify SC number surrounding NMJs, circles (75 µm in diameter) were drawn around each NMJ in Image J. The DAPI channel was used as a mask to filter out nonspecific staining of Pax7 in the extracellular matrix.

Immunofluorescence and imaging of cultured myoblasts/myotubes and motor neurons

The cells were fixed in 4% PFA for 10 min and then washed with 1% glycine and PBS. Further fixation and permeabilization was done either with methanol at -20°C for 10 min or 0.25% TritonX-100 for 15 min. The samples were then washed with PBS and immersed in blocking buffer for 40 min. Primary antibody incubation was done at 4 °C overnight. Next day the cells were washed with PBS and incubated with Alexa Fluor labelled secondary antibodies (1:1000) for 2 hours at room temperature, counterstained with DAPI and mounted in antifade mounting media for imaging. Primary antibodies used: Alexa Fluor 488 conjugated α-Bungarotoxin (Thermo Fisher B13422, 1:1000), Pax7 (SCBT sc-81648, 1:100), MyoD (SCBT sc-377460, 1:100), Ki-67 (CST 12202S, 1:300), MHC (DSHB MF20 concentrate, 1: 200), neurofilament (DSHB 2H3 concentrate, 1:250), eGFP (Novus NBP-22111AF488, 1:250).

For quantifying innervation ratio, AChR clusters (BTX patches) larger than 10 μm^2 were counted on myotubes that have contact with RSMN neurites. Those myotubes having 0 pixel overlap with RSMN neurites were not included in measurement. The ratio of AChR clusters having overlaps with the RSMN neurites after visual inspection of the whole z-stacks of each view were calculated. The innervation ratio of each view is one dot in the box-and-dot plot in Figure 9C. The measurements were done for 3 rounds of coculture assays.

RNA extraction and qRT-PCR

For cultured SCs, cell culture medium was removed before RNA extraction and Trizol reagent was added to the well before scratching cells off using pipette tips. The homogenate was transferred into tubes containing 5 PRIME Phase Lock Gel (Quantabio 2302830), and 1/10 volume of 1-bromo-3-chloroporpane (Alfa Aesar A11395) was added. The samples were shaken vigorously for 12 s and centrifuged for 15 min at 16,000× g. Afterwards, the upper phase was transferred to another Eppendorf tube and isopropanol (1/2 volume of the initial homogenate) was added for precipitation overnight at −20 °C. Next day, after 15 min centrifugation at 16000 g the precipitate was washed briefly with 75% ethanol. After another 5 min centrifugation, the ethanol was removed. The precipitate was briefly dried and resuspended in RNase free water. For whole muscle samples, the freshly dissected muscles were snap-frozen in liquid nitrogen, kept at -80 °C freezer overnight and then soaked in RNALater ICE (Thermo Fisher AM7030) for 2-3 days at -20 °C. Before RNA extraction, leftover tendons and neurites were removed and muscles were transferred to 1 ml Trizol. Homogenization was conducted in FastPrep-24 Classic bead beating grinder (MP Biomedicals 116004500). The tissue homogenate was transfer to centrifuge tubes containing phase lock gel (QuantaBio 2302830) and 1/10 volume of 1-bromo-3-chloropropane was added. The tubes were hand shaken for 12 seconds, left at bench top for 2 minutes and then centrifuged at 16000 g at 4 °C for 15 minutes. The upper phase was transferred to a new centrifuge tube and mixed with equal volume of ethanol. The following steps of RNA purification was conducted with Direct-zol RNA Miniprep Plus kit (Zymo Research R2070). RNA concentration was measured with Quantus Fluorometer (Promega E6150). RNA quality was assessed by RNA ScreenTape analysis at Genomics Core Facility at the University of North Texas Health Science Center. Only RNAs with RIN ≥ 9 and 28s/18s ≥ 1 were used for qRT-PCR or RNA-Seq. Reverse transcription was done using Promega GoScript Reverse Transcription system. The qPCR reactions used Maxima SYBR Green/ROX qPCR Master Mix (Thermo Fisher) and were carried out by the StepOnePlus Real-Time PCR system (Thermo Fisher). Relative quantification (RQ) of gene expression was generated by ΔΔCt method. The sequences of primers used for qPCR were listed in Key Resource Table.

Bulk RNA-Seq and analysis

Non-stranded mRNA sequencing was carried out by BGI Americas Corporation. The average number of clean reads per sample was 45.16 million (ranging from 44.11 to 48.51 million). The average percentage of clean reads was 93.32% (ranging from 92.42% to 93.97%). The average percentage of aligned reads (to Genome Reference Consortium Mouse Build 38 patch release 6) was 96.01% (ranging from 95.40% to 96.67%) and the average percentage of uniquely aligned reads was 86.46% (ranging from 85.12% to 87.63%). To identify differentially expressed genes (DEGs) between different samples (using Poisson distribution method to compare uniquely mapped reads of the gene), only genes with |log2FC| ≥ 0.4 and FDR ≤ 0.001 were considered. To identify DEGs between different groups (containing more than one sample, using DESeq2 method), only genes with |log2FC| ≥ 0.4 and Qvalue ≤ 0.05 were considered. PCA analysis, hierarchical clustering, functional annotation analysis (KEGG pathway, GO_C, GO_P) were performed in Dr. TOM platform of BGI. Dot plots showing the distribution of activation, differentiation and quiescence signature genes were generated in ggplot2 in R. Fastq files and the excel file containing TPM values has been uploaded to GEO database with record number GSE249484.

Data analysis and statistics

The experimenters were not blinded to the samples in data collection and analysis. Box-and-dot plots were generated using the open-source ggplot2 data visualization package for the statistical programming language R. Two independent sample t-tests were carried out by adding the stat_compare_means function. For multi-group data, one-way ANOVA P-values were generated using the R function aov(). For each two-group comparison, we used two independent sample t-tests and used brackets to highlight the groups compared.

Supporting information